### 2021-02-28 manuscript/manuscript.tex
initial commit of manuscript files

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmc_template/bmcart}

%%% Load packages
%\usepackage{amsthm,amsmath}
\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \def\includegraphic{}
% \def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\newcommand{\bm}[1]{{\bf #1}}
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Research}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Continuous estimates of heat emission at Mt. Ruapehu using
the unscented Kalman Smoother}

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   noteref={n1},                        % id's of article notes, if any
   email={y.behr@gns.cri.nz}   % email address
]{\inits{YB}\fnm{Yannik} \snm{Behr}}
\author[
   addressref={aff1},
   email={s.sherburn@gns.cri.nz}
]{\inits{SS}\fnm{Steve} \snm{Sherburn}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%
    \orgname{GNS Science},
  \street{114 Karetoto Rd},
  \postcode{3384}
  \city{Taupo},
  \cny{New Zealand}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract} % abstract
\parttitle{First part title} %if any
Text for this section.

\parttitle{Second part title} %if any
Text for this section.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{sample}
\kwd{article}
\kwd{author}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Introduction}

The amount and rate of heat emitted by a volcano provides important clues about
volcanic processes at depth. For example, a sudden rise in the heat output rate
combined with increased microgravity signals may result from the emplacement of
fresh magma at shallow depth \cite{Brown1991}. A slow decrease in the heat
emission rate together with a high carbon-sulfur ratio of the emitted gas may
point to the formation of a hydrothermal seal.

Volcanic lakes help with estimating heat emissions. Often, the majority of a
volcano's heat and gas output passes through the crater lake. Crater lakes act thus
like a filter, integrating heat and gas input from the vents entering the lake.

Monitoring changes in temperature, water mass, and ion concentration in these
lakes can thus serve as proxies for changes in gas- and steam-release by the
hydrothermal system and, ultimately, the melt zone beneath the volcano.

Crater Lake (Te Wai A-Moe) on Mt. Ruapehu, an active andesitic stratovolcano in
the center of New Zealand's North Island (Figure \ref{overview}), is one such
example with records of observations dating back to the 19th century
\citep[e.g.][]{Friedlander1898}. Mt. Ruapehu erupted at least 602 times since
1830 with many phreatic and phreato-magmatic eruptions and two major magmatic
episodes \citep{Scott2013}. Although the latest magmatic event between June
1995 to November 1997 and a subsequent dam collapse changed the shape and
volume of the lake, it has been largely unchanged since then and the lake's
bathymetry is well known from times when it was empty [Tony Hurst, personal
communication].

Increases in heat and steam input may be caused by the ascent of magma or
changes in the permeability of the hydrothermal system and are, therefore,
important parameters to be monitored. First attempts to quantify the amount of
heat and steam entering the lake date back to \citep{Dibble1966}. Over the past
decade water temperature and lake level have been monitored continuously
and water samples for ion concentration measurements have been taken roughly
every two to three months. This enables us to continuously estimate heat and
steam input into the lake and, in principle, automatically detect changes in
these variables as soon as they occur.

Inference of the heat and steam input is typically done through a mass and
energy balance calculation: the sum of mass and energy flow into and out of the
lake has to account for the observed changes in lake temperature, water level
and ion dilution \citep[e.g.,][]{Hurst1981, Hurst1991, Stevenson1992,
Fournier2009, Scott1994}. At Crater Lake, and probably at most other volcanic
lakes, this is an underdetermined problem, that is the number of unknowns
exceeds the number of constraints. Hence, it is necessary to constrain some of
the parameters in the mass and energy balance model using prior knowledge. Further,
some observations and input parameters come with large uncertainties and are sampled
irregularly.

\section*{Section title}
Text for this section \ldots
\subsection*{Sub-heading for section}
Text for this sub-heading \ldots
\subsubsection*{Sub-sub heading for section}
Text for this sub-sub-heading \ldots
\paragraph*{Sub-sub-sub heading for section}
Text for this sub-sub-sub-heading \ldots
In this section we examine the growth rate of the mean of $Z_0$, $Z_1$ and $Z_2$. In
addition, we examine a common modeling assumption and note the
importance of considering the tails of the extinction time $T_x$ in
studies of escape dynamics.
We will first consider the expected resistant population at $vT_x$ for
some $v>0$, (and temporarily assume $\alpha=0$)
%
\[
 E \bigl[Z_1(vT_x) \bigr]= E
\biggl[\mu T_x\int_0^{v\wedge
1}Z_0(uT_x)
\exp \bigl(\lambda_1T_x(v-u) \bigr)\,du \biggr].
\]
%
If we assume that sensitive cells follow a deterministic decay
$Z_0(t)=xe^{\lambda_0 t}$ and approximate their extinction time as
$T_x\approx-\frac{1}{\lambda_0}\log x$, then we can heuristically
estimate the expected value as
%
\begin{eqnarray}\label{eqexpmuts}
E\bigl[Z_1(vT_x)\bigr] &=& \frac{\mu}{r}\log x
\int_0^{v\wedge1}x^{1-u}x^{({\lambda_1}/{r})(v-u)}\,du
\nonumber\\
&=& \frac{\mu}{r}x^{1-{\lambda_1}/{\lambda_0}v}\log x\int_0^{v\wedge
1}x^{-u(1+{\lambda_1}/{r})}\,du
\nonumber\\
&=& \frac{\mu}{\lambda_1-\lambda_0}x^{1+{\lambda_1}/{r}v} \biggl(1-\exp \biggl[-(v\wedge1) \biggl(1+
\frac{\lambda_1}{r}\biggr)\log x \biggr] \biggr).
\end{eqnarray}
%
Thus we observe that this expected value is finite for all $v>0$ (also see \cite{koon,khar,zvai,xjon,marg}).
%\nocite{oreg,schn,pond,smith,marg,hunn,advi,koha,mouse}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Competing interests}
  The authors declare that they have no competing interests.

\section*{Author's contributions}
    Text for this section \ldots

\section*{Acknowledgements}
  Text for this section \ldots
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc_template/bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{zotero.bib}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

%% \section*{Figures}
%%   \begin{figure}[h!]
%%   \caption{\csentence{Sample figure title.}
%%       A short description of the figure content
%%       should go here.}
%%       \end{figure}
%% 
%% \begin{figure}[h!]
%%   \caption{\csentence{Sample figure title.}
%%       Figure legend text.}
%%       \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
%% \section*{Tables}
%% \begin{table}[h!]
%% \caption{Sample table title. This is where the description of the table should go.}
%%       \begin{tabular}{cccc}
%%         \hline
%%            & B1  &B2   & B3\\ \hline
%%         A1 & 0.1 & 0.2 & 0.3\\
%%         A2 & ... & ..  & .\\
%%         A3 & ..  & .   & .\\ \hline
%%       \end{tabular}
%% \end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% \section*{Additional Files}
%%   \subsection*{Additional file 1 --- Sample additional file title}
%%     Additional file descriptions text (including details of how to
%%     view the file, if it is in a non-standard format or the file extension).  This might
%%     refer to a multi-page table or a figure.
%% 
%%   \subsection*{Additional file 2 --- Sample additional file title}
%%     Additional file descriptions text.

\end{backmatter}
\end{document}

### 2021-07-01 manuscript/manuscript.tex
add more sections; change synthetic test section and bibliography

@@
 %\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+\usepackage{graphicx}
 
@@
 \title{Continuous estimates of heat emission at Mt. Ruapehu using
-the unscented Kalman Smoother}
+the Unscented Kalman Smoother}
 
@@
 ]{\inits{SS}\fnm{Steve} \snm{Sherburn}}
+\author[
+   addressref={aff1},
+   email={t.hurst@gns.cri.nz}
+]{\inits{TH}\fnm{Tony} \snm{Hurst}}
 
@@
 
+We propose a probabilistic (Bayesian) inference approach which naturally
+handles these difficulties and transparently propagates uncertainties of
+observations and prior assumptions into posterior uncertainties of heat and
+steam input. The accurate description of uncertainties is of particular
+importance when publishing automatically generated estimates of heat and steam
+input, as they could potentially be used in decision making or automatic
+alerting of monitoring staff.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/overview.png}  
+\caption{Image of Mt. Ruapehu's crater lake from 2008 (\textcopyright Karen
+	Britten, GNS Science). The red O marks the location of the lake outlet. The
+	location of Mt. Ruapehu is shown as a red triangle on the insert map.}
+  \label{overview}
+\end{figure}
+
+\section*{Probabilistic inference method}\label{Pim}
+
+\subsection*{Physical model}\label{phm}
+
+Following \citet{Hurst1991} and \citet{Stevenson1992} the mass and energy balance
+can be written in terms of three coupled ordinary differential equations. The
+rate of lake water mass change can be described as:
+\begin{equation}\label{ode_M}
+	\frac{dM}{dt} = M_s - M_o - M_e + M_i
+\end{equation}
+
+Where $M_s$ is the rate of water inflow through melt and precipitation, $M_o$
+is the outflow rate through seepage and overflow, $M_e$ is the rate of mass
+loss through evaporation and $M_i$ is the rate of steam input through the vent
+system beneath the lake.
+
+The thermal energy content of a mass flow can be described with the notion of
+enthalpy, $H$, which is the ratio of energy and mass. Enthalpy can also be defined
+on a relative scale as:
+\begin{equation}
+	H(T) = H(T_0) + \int_{T_0}^{T}c_p\tau d\tau
+\end{equation}
+
+where $T$ is temperature and $c_p$ is the specific heat. For all following
+equations we take $H(T_0=0)$ as our reference enthalpy and $c_p=c_w=constant$,
+with $c_w$ the specific heat of water, such that the enthalpy of lake water at
+temperature $T$ can be written as $H(T) = c_w T$. Because Ruapehu Crater Lake
+is at an altitude of ~2500 m and big parts of its catchment are covered by snow
+and ice all year round we assume the temperature of incoming water to be at 0
+$^{\circ}C$.
+
+The rate of energy change of the lake water can thus be written as:
+\begin{equation}\label{ode_E}
+    \frac{dE}{dt}=\frac{d(c_wTM)}{dt}=c_wT\frac{dM}{dt} + c_wM\frac{dT}{dt} =
+	-Q_e - c_wTM_o + Q_{eff} + Q_r
+\end{equation}
+
+where $T$ is the lake water temperature, $M$ is the lake water mass, $Q_r$ is
+the rate of energy gain due to solar irradiation (short-wavelength radiation),
+and $c_wTM_o$ is the rate of energy loss through outflow/seepage. $Q_e$ is
+the rate of surface energy losses due to evaporation (forced and free
+convection), sensible heat, and long-wavelength radiation (net loss). Several
+empirical and theoretical equations exist to describe the surface losses and
+both \citet{Stevenson1992} and \citet{Hurst2015} discuss them in detail. We
+follow here mostly the choice of \citet{Hurst2015} as it includes the recent
+development by \citet{Sartori2000} on forced (i.e. wind-induced) convection
+(see Appendix \ref{A} for more details). $Q_{eff}$ is the effective rate of
+energy gain from volcanic sources after heating incoming water:
+\begin{equation}
+	Q_{eff}=Q_i - M_s c_w T
+\end{equation}
+with $Q_i$ the rate of energy input from volcanic sources.
+
+Following from Equation \ref{ode_E} the temperature change $\frac{dT}{dt}$ can then be written as:
+\begin{equation}\label{ode_T}
+	\frac{dT}{dt}=\frac{1}{c_wM}\left(-Q_e - c_wTM_o + Q_{eff} + Q_r\right )
+	-\frac{T}{M}\frac{dM}{dt}
+\end{equation}
+
+Finally, the change in ion concentration of, for example, $Mg^{2+}$ can be expressed as:
+\begin{equation}\label{ode_X}
+	\frac{dX}{dt}=-M_o\frac{X}{M}
+\end{equation}
+
+Where $X$ is the total ion amount in the lake. This equation is only true under
+the assumption that $X$ only changes due to dilution and there is no re-supply
+between eruptive episodes. Equation \ref{ode_X} gives an estimate of the average
+outflow over periods on the order of months but does not capture shorter term
+variations in the outflow.
+
+\subsection*{Inference}\label{inf}
+
+The objective of the inference problem is to estimate the amount and energy
+content of steam and gas entering the lake from the hydrothermal system below.
+To constrain these we use the currently available observations of lake
+temperature ($T$), lake level ($L$), ion concentration ($X$) as well as
+numerical weather model data on windspeed. Putting it more concisely, we are
+looking for
+
+\begin{equation}\label{bsm}
+    p(\bm{x}_k|\bm{y}_{1:T})
+\end{equation}
+
+where $\bm{y}_{1:T}$ is the time series of $m$ observations and $x_k$ comprises
+the $n$ model parameters described in Section \ref{phm} at time step $k$. Note
+that $x_k$ also includes the lake's temperature, lake level, and ion
+concentration. We distinguish here between the state of the lake and
+observations of this state ($T$, $L$, $X$) which are to some degree uncertain.
+
+The following equations are based on \citet{Sarkka2010} which we refer the
+reader to for more proofs and more detailed explanations.
+
+Equation \ref{bsm} is the marginal probability of $\bm{x}_k$ given all
+observations until time $T$. We assume that the states of the lake form a
+Markov chain, meaning that the state of the lake at time step $k$, $\bm{x}_k$,
+only depends on the state at adjacent time steps:
+
+\begin{equation}\label{markov1}
+    p(\bm{x}_k|\bm{x}_{1:k-1}, \bm{y}_{1:k-1}) = p(\bm{x}_k|\bm{x}_{k-1})
+\end{equation}
+
+and
+
+\begin{equation}\label{markov2}
+    p(\bm{x}_{k-1}|\bm{x}_{k:T}, \bm{y}_{k:T}) = p(\bm{x}_{k-1}|\bm{x}_{k})
+\end{equation}
+
+With this assumption and using \textit{Bayes' rule}, Equation \ref{bsm} can be
+rewritten as:
+
+\begin{equation}\label{bsm1}
+p(\bm{x}_k|\bm{y}_{1:T}) = p(\bm{x}_k|\bm{y}_{1:k})
+    \int\left[\frac{p(\bm{x}_{k+1}|\bm{x}_k)p(\bm{x}_{k+1}|\bm{y}_{1:T})}{p(\bm{x}_{k+1}|\bm{y}_{1:k})}\right]d\bm{x}_{k+1}
+\end{equation}
+
+with
+
+\begin{eqnarray}\label{bfilt1}
+p(\bm{x}_{k+1}|\bm{y}_{1:k}) & = & \int p(\bm{x}_{k+1}|\bm{x}_k)p(\bm{x}_k|\bm{y}_{1:k})d\bm{x}_k \\
+        \label{bfilt}
+    p(\bm{x}_k|\bm{y}_{1:k}) & = & \frac{1}{z_k} p(\bm{y}_k|\bm{x}_k)p(\bm{x}_k|\bm{y}_{1:k-1}) \\
+        \label{zetk}
+    z_k & = & \int p(\bm{y}_k|\bm{x}_k) p(\bm{x}_k| \bm{y}_{1:k-1}) d\bm{x}_k  
+\end{eqnarray}
+
+using the \textit{Chapman-Kolmogorov} equation we can write:
+
+\begin{equation}
+p(\bm{x}_k|\bm{y}_{1:k-1}) = \int p(\bm{x}_k|\bm{x}_{k-1})p(\bm{x}_{k-1}|\bm{y}_{1:k-1})d\bm{x}_{k-1} 
+\end{equation}
+
+Equations \ref{bsm1} to \ref{zetk} describe a recursive way to compute Equation
+\ref{bsm} in which we only need to know $\bm{x}_1$,
+$p(\bm{x}_k|\bm{x}_{k-1})$, and $p(\bm{y}_k|\bm{x}_k)$. Equation \ref{bsm} and
+\ref{bfilt} are also known as the Bayesian Smoothing and Bayesian Filtering
+equation, respectively. If $p(\bm{x}_k|\bm{x}_{k-1})$ and
+$p(\bm{y}_k|\bm{x}_k)$ are linear transformations with normally distributed
+errors, the \textit{Kalman filter} \citep{Kalman1960} and Kalman smoother
+\citep{Rauch1965} represent closed form solutions. In our case, these
+transformations are:
+
+\begin{eqnarray}
+p(\bm{x}_k|\bm{x}_{k-1}) & = & f(\bm{x}_{k-1}) + \bm{q}_{k-1} \\
+    p(\bm{y}_k|\bm{x}_k) & = & \bm{x}_k[1:m] + \bm{r}_k   
+\end{eqnarray}
+
+where $f(\bm{x}_{k-1})$ is the physical model described in Section \ref{phm}
+which is highly non-linear. If we assume the errors, $\bm{r}_k$ and
+$\bm{q}_{k-1}$, to be normally distributed ($\bm{q}_{k-1} \sim N(\bm{0},
+\bm{Q}_{k-1})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$), we can compute an
+approximate solution to \ref{bsm} using the \textit{Unscented Kalman Filter}
+\citep{Merwe2004}.
+
+We estimate $\bm{R}_k$ from the daily variations in the observations but we do
+not know $\bm{x}_1$ and $\bm{Q}_{k-1}$. If $\bm{\theta}$ stands for all unknown
+parameters than we are looking for $p(\bm{\theta}|\bm{y}_{1:T})$. Using again
+\textit{Bayes' Theorem} and ignoring the normalising constant we can write:
+
+\begin{equation}\label{pe1}
+    p(\bm{\theta}|\bm{y}_{1:T}) \propto p(\bm{y}_{1:T}|\bm{\theta})p(\bm{\theta})
+\end{equation}
+
+Assuming that $\bm{Y}_{1:T}$ form a Markov chain and that $p(\bm{\theta})$ follows
+a uniform distribution we can rewrite Equation \ref{pe1} as:
+
+\begin{equation}\label{pe2}
+    p(\bm{\theta}|\bm{y}_{1:T}) \propto \prod_{k=1}^Tp(\bm{y}_{k}|\bm{y}_{k-1}, \bm{\theta})
+\end{equation}
+
+Each factor on the right-hand side of Equation \ref{pe2} can be written as:
+\begin{equation}
+    p(\bm{y}_k|\bm{y}_{1:k-1}, \bm{\theta}) = \int p(\bm{y}_k|\bm{x}_k, \bm{\theta})p(\bm{x}_k|\bm{y}_{1:k-1})d\bm{x}_k
+\end{equation}
+
+which is readily computed as a by-product of Equation \ref{bfilt}. We then run
+a non-linear optimisation to find the \textit{maximum a posteriori (MAP)}
+estimate of Equation \ref{pe2}.
+
+* sensitivity analysis: high vs. low lake temperature; windspeed dependence;
+  enthalpy dependence
+
+\section*{Results}
+\subsection*{Synthetic test}\label{syn_test}
+To test the inversion scheme we design a synthetic test, consisting of a very
+large cylinder, the same volume as Ruapehu Crater Lake. We keep the windspeed
+and surface water inflow constant at 4.5 m/s and 10 kt/day, respectively. To
+model the outflow we assume the outlet to be a pipe 0.2 $m^2$ in cross section
+and then use Bernoulli's equation to model the water outflow based on the water
+level in the cylinder above the pipe. We assume that the incoming heat follows
+a combination of a third-order polynomial and a sinusoid function and that the
+enthalpy of the incoming steam is 2.8 which corresponds to the enthalpy of
+saturated steam at hydrostatic pressures similar to those at the bottom of the
+lake \ref{Mayhew1978}. To the resulting synthetic observations we add
+uncertainties similar to those estimated from real observations (see Section
+\ref{realdata}).
+
+Figure \ref{syn_example} shows the result of the probabilistic inference of
+heat input compared to the true input of the synthetic model. The observations
+are well fit within their error bounds but we see some descrepancies between
+the original and the recovered heat input rates, especially for very steep
+changes in the input rate. This is the result of preferring a smooth solution
+as already mentioned in \ref{inf}.
+
+It is noteworthy that the mass influx and outflux rates are recovered quite accurately despite having no data in the former and very little data in the latter case.
+
+\begin{figure}
+	\includegraphics[width=.9\textwidth]{figures/synthetic_inversion_uks.png}  
+    \caption{Inversion result for the synthetic test described in Section 
+        \ref{syn_test}. Turquoise lines and markers represent hypothetical
+        observations; red dashed lines are the inversion results and black
+        dashed lines are the true input to the synthetic test.}
+  \label{syn_example}
+\end{figure}
+
 \section*{Section title}
@@
 % if your bibliography is in bibtex format, use those commands:
-\bibliographystyle{bmc_template/bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{zotero.bib}      % Bibliography file (usually '*.bib' )
+\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
+\bibliography{zotero}      % Bibliography file (usually '*.bib' )
 % for author-year bibliography (bmc-mathphys or spbasic)

### 2021-07-01 src/pumahu/syn_model.py
remove old test scenarios; add function to make synthetic obs sparse

@@ -78,57 +78,12 @@ def setup_realistic(q_in=1000., tmax=90., sinterval=10):
     # Mi
     data[1, :] = np.ones(t.size)*10.
     # H
-    data[2, :] = np.ones(t.size)*6
+    data[2, :] = np.ones(t.size)*2.8
     # W
     data[3, :] = np.ones(t.size)*4.5
     return xr.DataArray(data, coords=[('params', params),
                                       ('times', dates)])
-#        if mode == 'sinus':
-#            qi = np.sin(2.*np.pi*self.f*t + np.deg2rad(15))*q_in/2.+q_in/2.
-#            qi *= tukey(nsteps, 0.9)
-#        if mode == 'tukey':
-#            tail = int(10/dt)
-#            qi = np.ones(nsteps-tail)*q_in
-#            qi *= tukey(nsteps-tail, 0.9)
-#            qi_tmp = np.zeros(nsteps)
-#            qi_tmp[0:nsteps-tail] = qi
-#            qi = qi_tmp
-#        if mode == 'step':
-#            tail = int(10/dt)
-#            qi = np.ones(nsteps-tail)*q_in
-#            qi_tmp = np.zeros(nsteps)
-#            qi_tmp[0:nsteps-tail] = qi
-#            qi = qi_tmp
-#        if mode == 'constant':
-#            qi = np.ones(nsteps)*q_in
-#        if mode == 'gamma':
-#            y = gamma.pdf(t, 3., loc=4)
-#            y /= y.max()
-#            y *= q_in
-#            qi = y
-#
-#        if mode == 'observed':
-#            # a polynomial model of a real inversion
-#            z = np.array([-6.93795125e-30,  1.17673746e-26, -9.19940930e-24,
-#                          4.39229128e-21, -1.43037875e-18,  3.36004438e-16,
-#                          -5.87098984e-14,  7.75486512e-12, -7.79213224e-10,
-#                          5.94657977e-08, -3.41657063e-06,  1.45392771e-04,
-#                          -4.46901292e-03,  9.56540064e-02, -1.35189993e+00,
-#                          1.16488040e+01, -5.37668791e+01,  1.05017651e+02,
-#                          2.91772946e+01])
-#            p = np.poly1d(z)
-#            dates = pd.date_range(start='2018-4-1', end='2018-9-30',
-#                                  periods=183)
-#            t = np.r_[0, np.cumsum(np.diff(dates)/np.timedelta64(1, 'D'))]
[…]

### 2021-07-01 src/pumahu/visualise.py
change how nans are handled in trellis plot; add dashed lines to trellis plot

@@ -27,26 +27,53 @@ class TrellisPlot:
         self.line_colours = ['rgb(26,166,183)',
                              'rgb(255,65,77)',
                              'rgb(0,45,64)']
+        self.line_dash = ['solid', 'dash', 'dash'] 
         self.error_colours = ['rgba(26,166,183,.3)',
                               'rgba(255,65,77, .3)',
                               'rgba(0,45,64,.3)']
         self.legend_traces = []
         self._ntraces = defaultdict(lambda: -1)
 
-    def get_traces(self, data, key, err_min=1e-3):
+    def get_traces(self, data, key, err_min=1e-3,
+                   nanthresh=0.4):
+        """
+        Extract traces
+        
+        Parameters:
+        -----------
+        data : :class:`xarray.DataArray`
+               Data structure that contains the values
+               and error estimates.
+        key : str
+              Dimension name of the data to extract.
+        err_min : float
+                  Minimum error to plot in case max
+                  and min traces are too close.
+        nanthresh : float
+                    Interpolate if the number of NaNs is
+                    less than nanthresh.
+        """
         if 'val_std' in data.dims:
             # ignore warnings due to NaNs
-            with np.errstate(invalid='ignore'):
-                ymean = data.loc[:, key, 'val'].values
-                ymean = np.where(ymean < 0., 0., ymean)
-                yerr = np.sqrt(data.loc[:, key, 'std'].values)
-                ymin = ymean - 3*yerr
-                ymin = np.where(ymin < 0., 0., ymin)
-                ymax = ymean + 3*yerr
-                if np.nanmax(ymax-ymin) < err_min:
-                    ymax = ymean + err_min
-                    ymin = ymean + err_min
-                return [ymean, ymin, ymax]
+            ymean = data.loc[:, key, 'val'].values
+            ymean = np.where(ymean < 0., 0., ymean)
+            yvar = data.loc[:, key, 'std'].values
+            yerr = np.sqrt(yvar, where=yvar>=0)
+            # Interpolate over negative values
+            idx = np.where(yerr<0)[0]
+            yerr[idx] = np.nan
+            # calculate the percentage of NaN values
+            nanperc = float(np.sum(data.loc[:, key, 'val'].isnull())/data.shape[0])
+            if nanperc < nanthresh:
+                yerr = pd.Series(yerr).interpolate().values
+                #yerr = yerr.interpolate_na(method='linear').values
[…]

### 2021-07-02 src/pumahu/uks.py
add more plots; tweak Q

@@ -20,7 +20,7 @@ from .syn_model import SynModel
 from .forward_model import Forwardmodel
 from .sigma_points import MerweScaledSigmaPoints
 from .data import LakeData
-from .visualise import trellis_plot
+from .visualise import trellis_plot, plot_qin_uks
  
 import ipdb
 
@@ -105,10 +105,9 @@ class UnscentedKalmanSmoother:
         if P0 is not None:
             self.P0 = P0
             
-        _Q = OrderedDict(T=1e-3, M=1e-3, X=1e-3, q_in=1e1,
-                         m_in=1e1, m_out=1e1, h=1e-3, W=1e1,
-                         dqi=1e3, dMi=1e4, dMo=1e4, dH=1e-3, 
-                         dW=1e1)
+        _Q = OrderedDict(T=1e-1, M=1e1, X=1e-3, q_in=1e2,
+                         m_in=1e1, m_out=1e1, h=1e-3, W=1e-3,
+                         dqi=0, dMi=0, dMo=0, dH=0, dW=0)
         _Q = np.eye(len(_Q))*list(_Q.values())
         self.Q = _Q
         if Q is not None:
@@ -366,11 +365,7 @@ def mainCore(args):
         args.endtime = datetime.utcnow()
     data = ld.get_data(args.starttime, args.endtime, smoothing='dv')
     # Setup path for results file
-    tstart = pd.to_datetime(data['dates'].values[0])
-    tend = pd.to_datetime(data['dates'].values[-1])
-    res_fn = 'uks_{:s}_{:s}.nc'
-    res_fn = res_fn.format(tstart.strftime('%Y-%m-%d'),
-                           tend.strftime('%Y-%m-%d'))
+    res_fn = 'uks.nc'
     if args.pretxt is not None:
         res_fn = args.pretxt + '_' + res_fn
     res_fn = os.path.join(args.rdir, res_fn)
@@ -381,7 +376,14 @@ def mainCore(args):
     if args.plot:
         xdf = xr.open_dataset(res_fn)
         fout_trellis = os.path.join(args.rdir, 'uks_trellis.png')
-        trellis_plot(xdf, filename=fout_trellis)
+        if args.benchmark is not None:
+            data2 = xr.open_dataset(args.benchmark)
+            trellis_plot(xdf, data2=data2, filename=fout_trellis)
+            plot_qin_uks(xdf, data_mcmc=data2,
[…]

### 2021-07-02 src/pumahu/mcmc.py
change default output file

@@ -279,11 +279,7 @@ def mainCore(args):
         args.endtime = datetime.utcnow()
     data = ld.get_data(args.starttime, args.endtime)
     # Setup path for results file
-    tstart = pd.to_datetime(data['dates'].values[0])
-    tend = pd.to_datetime(data['dates'].values[-1])
-    res_fn = 'mcmc_sampling_{:s}_{:s}.nc'
-    res_fn = res_fn.format(tstart.strftime('%Y-%m-%d'),
-                           tend.strftime('%Y-%m-%d'))
+    res_fn = 'mcmc_sampling.nc'
     if args.pretxt is not None:
         res_fn = args.pretxt + '_' + res_fn
     res_fn = os.path.join(args.rdir, res_fn)

### 2021-07-02 src/pumahu/visualise.py
add MAP plot

@@ -63,10 +63,10 @@ class TrellisPlot:
             idx = np.where(yerr<0)[0]
             yerr[idx] = np.nan
             # calculate the percentage of NaN values
-            nanperc = float(np.sum(data.loc[:, key, 'val'].isnull())/data.shape[0])
+            nanperc = float(np.sum(np.isnan(ymean))/data.shape[0])
             if nanperc < nanthresh:
                 yerr = pd.Series(yerr).interpolate().values
-                #yerr = yerr.interpolate_na(method='linear').values
+                ymean = pd.Series(ymean).interpolate().values
             ymin = ymean - 3*yerr
             ymin = np.where(ymin < 0., 0., ymin)
             ymax = ymean + 3*yerr
@@ -275,6 +275,41 @@ def mcmc_heat_input(data, filename=None):
     axs[1].set_ylim(0, 1100)
     if filename is not None:
         plt.savefig(filename, dpi=300, bbox_inches='tight')
+    return fig
+
+
+def plot_qin_uks(data_uks, data_mcmc=None, data2y=None, filename=None):
+    """
+    Plot the MAP of the heat input rate for the 
+    Unscented Kalman Smoother solution.
+    Parameters:
+    -----------
+    data_uks: :class:`xarray.DataArray`
+              The MAP values from the UKS solution.
+    data_mcmc: :class:`xarray.DataArray`
+               The MAP values from the MCMC solution
+    data2y: (list-like, list-like, str)
+            Data to plot on the secondary y-axes in order
+            to compare it to the heat input rate. The last
+            entry in the tuple defines the legend name.
+    filename : str
+               Filename to save output to.
+    """
+    tp = TrellisPlot()
+    with plt.style.context('bmh'):
+        fig = make_subplots(rows=1, cols=1, specs=[[{"secondary_y": True}]])
+        tp.plot_trace(fig, data_uks, 'q_in', 'Qi [MW]', 1, filled_error=True,
+                      plotvars=[('exp', 'UKS')])
+        if data_mcmc is not None:
+            tp.plot_trace(fig, data_mcmc, 'q_in', 'Qi [MW]', 1, filled_error=True,
+                          plotvars=[('exp', 'MCMC')])
+        if data2y is not None:
+            x, y, name = data2y
+            fig.add_trace(go.Scatter(x=x, y=y, mode='lines', name=name),
+                          secondary_y=True)
+            fig.update_yaxes(title_text=name, secondary_y=True)
[…]

### 2021-07-02 job_scheduler.py
bugfix

@@ -4,7 +4,9 @@ import time
 
 def job():
     subprocess.run(['heat_mcmc', '--rdir', '/opt/data', '-f', '-p', '-d'])
-    subprocess.run(['heat_mcmc', '--rdir', '/opt/data', '-f', '-p', '-d'])
+    subprocess.run(['heat_uks', '--rdir', '/opt/data',
+                    '-f', '-p', '-d',
+                    '--benchmark', 'mcmc_sampling.nc'])
 
 schedule.every().day.at("11:00").do(job)
 

### 2021-07-16 manuscript/manuscript.tex
document compiles with all sections added

--- 2021-02-28 manuscript/manuscript.tex
@@
 
-Increases in heat and steam input may be caused by the ascent of magma or
+Increases in heat input may be caused by the ascent of magma or
 changes in the permeability of the hydrothermal system and are, therefore,
 important parameters to be monitored. First attempts to quantify the amount of
-heat and steam entering the lake date back to \citep{Dibble1966}. Over the past
-decade water temperature and lake level have been monitored continuously
+heat entering the lake date back to \citep{Dibble1966}. Since 2016
+water temperature and lake level have been monitored continuously
 and water samples for ion concentration measurements have been taken roughly
-every two to three months. This enables us to continuously estimate heat and
-steam input into the lake and, in principle, automatically detect changes in
-these variables as soon as they occur.
+every two to three months. This enables us to continuously estimate heat input
+into the lake and, in principle, automatically detect changes as soon as they occur.
 
-Inference of the heat and steam input is typically done through a mass and
+Inference of the heat input is typically done through a mass and
 energy balance calculation: the sum of mass and energy flow into and out of the
@@
 
-\section*{Section title}
-Text for this section \ldots
-\subsection*{Sub-heading for section}
-Text for this sub-heading \ldots
-\subsubsection*{Sub-sub heading for section}
-Text for this sub-sub-heading \ldots
-\paragraph*{Sub-sub-sub heading for section}
-Text for this sub-sub-sub-heading \ldots
-In this section we examine the growth rate of the mean of $Z_0$, $Z_1$ and $Z_2$. In
-addition, we examine a common modeling assumption and note the
-importance of considering the tails of the extinction time $T_x$ in
-studies of escape dynamics.
-We will first consider the expected resistant population at $vT_x$ for
-some $v>0$, (and temporarily assume $\alpha=0$)
-%
-\[
- E \bigl[Z_1(vT_x) \bigr]= E
-\biggl[\mu T_x\int_0^{v\wedge
-1}Z_0(uT_x)
-\exp \bigl(\lambda_1T_x(v-u) \bigr)\,du \biggr].
-\]
-%
-If we assume that sensitive cells follow a deterministic decay
-$Z_0(t)=xe^{\lambda_0 t}$ and approximate their extinction time as
-$T_x\approx-\frac{1}{\lambda_0}\log x$, then we can heuristically
-estimate the expected value as
-%
-\begin{eqnarray}\label{eqexpmuts}
-E\bigl[Z_1(vT_x)\bigr] &=& \frac{\mu}{r}\log x
-\int_0^{v\wedge1}x^{1-u}x^{({\lambda_1}/{r})(v-u)}\,du
-\nonumber\\
-&=& \frac{\mu}{r}x^{1-{\lambda_1}/{\lambda_0}v}\log x\int_0^{v\wedge
-1}x^{-u(1+{\lambda_1}/{r})}\,du
-\nonumber\\
-&=& \frac{\mu}{\lambda_1-\lambda_0}x^{1+{\lambda_1}/{r}v} \biggl(1-\exp \biggl[-(v\wedge1) \biggl(1+
-\frac{\lambda_1}{r}\biggr)\log x \biggr] \biggr).
+In the following, we propose a probabilistic (Bayesian) inference approach which
+naturally handles these difficulties and transparently propagates uncertainties of
+observations and prior assumptions into posterior uncertainties of heat input.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{figures/overview.png}  
+\caption{Image of Mt. Ruapehu's crater lake from 2008 (\textcopyright Karen
+	Britten, GNS Science). The red O marks the location of the lake outlet. The
+	location of Mt. Ruapehu is shown as a red triangle on the insert map.}
+  \label{overview}
+\end{figure}
+
+\section*{Probabilistic inference method}\label{Pim}
+
+\subsection*{Physical model}\label{phm}
+
+Following \cite{Hurst1991} and \cite{Stevenson1992} the mass and energy balance
+can be written in terms of three coupled ordinary differential equations. The
+rate of lake water mass ($M$) change can be described as:
+\begin{equation}\label{ode_M}
+	\frac{dM}{dt} = \dot{M_s} - \dot{M_o} - \dot{M_e} + \dot{M_i}
+\end{equation}
+
+Where $\dot{M_s}$ is the rate of water inflow through melt and precipitation,
+$\dot{M_o}$
+is the outflow rate through seepage and overflow, $\dot{M_e}$ is the rate of mass
+loss through evaporation and $\dot{M_i}$ is the rate of steam input through the vent
+system beneath the lake.
+
+The energy content of the lake can be described with the notion of
+enthalpy, $H$. At constant pressure, enthalpy is defined as:
+\begin{equation}
+	H(T) = H(T_0) + \int_{T_0}^{T}c_pM\tau d\tau
+\end{equation}
+
+where $T$ is temperature and $c_p$ is the specific heat. For all following
+equations we take $H(T_0=0)$ as our reference enthalpy and $c_p=c_w=constant$,
+with $c_w$ the specific heat of water, such that the enthalpy of lake water at
+temperature $T$ can be written as $H(T) = c_wMT$. Because Ruapehu Crater Lake is
+ at an altitude of $\sim$~2500~m and big parts of its catchment are covered by snow
+and ice all year round we assume the temperature of incoming water to be at 0
+$^{\circ}C$.
+
+The rate of enthalpy change of the lake water can thus be written as:
+\begin{equation}\label{ode_E}
+    \frac{dH}{dt}=\frac{d(c_wTM)}{dt}=c_wT\dot{M} + c_wM\dot{T} =
+	-\dot{Q_e} - c_wT\dot{M_o} + \dot{Q_{eff}} + \dot{Q_r}
+\end{equation}
+
+where $T$ is the lake water temperature, $\dot{Q_r}$ is
+the rate of energy gain due to solar irradiation (short-wavelength radiation),
+and $c_wT\dot{M_o}$ is the rate of energy loss through outflow/seepage. $\dot{Q_e}$ is
+the rate of surface energy losses due to evaporation (forced and free
+convection), sensible heat, and long-wavelength radiation (net loss). Several
+empirical and theoretical equations exist to describe the surface losses and
+both \cite{Stevenson1992} and \cite{hurstCraterLakeEnergy2015} discuss them in detail. We
+follow here mostly the choice of \cite{hurstCraterLakeEnergy2015} as it includes the recent
+development by \cite{sartoriCriticalReviewEquations2000} on forced (i.e. wind-induced) convection
+(see Appendix \ref{A} for more details). $\dot{Q_{eff}}$ is the effective rate of
+energy gain from volcanic sources after heating incoming water:
+\begin{equation}
+	\dot{Q_{eff}}=\dot{Q_i} - c_wT\dot{M_s}
+\end{equation}
+with $\dot{Q_i}$ the rate of energy input from volcanic sources.
+
+Following from Equation \ref{ode_E} the temperature change $\frac{dT}{dt}$ can then be written as:
+\begin{equation}\label{ode_T}
+	\frac{dT}{dt}=\frac{1}{c_wM}\left(-\dot{Q_e} - c_wT\dot{M_o} + \dot{Q_{eff}} + \dot{Q_r}\right )
+	-\frac{T}{M}\dot{M}
+\end{equation}
+
+Finally, the change in ion concentration of, for example, $Mg^{2+}$ can be expressed as:
+\begin{equation}\label{ode_X}
+	\frac{dX}{dt}=-\dot{M_o}\frac{X}{M}
+\end{equation}
+
+Where $X$ is the total ion amount in the lake. This equation is only true under
+the assumption that $X$ only changes due to dilution and there is no re-supply
+between eruptive episodes. Equation \ref{ode_X} gives an estimate of the average
+outflow over periods on the order of months but does not capture shorter term
+variations in the outflow.
+
+In a sensitivity analysis for $\dot{T}$ (Figure \ref{sens}) we can see a linear
+relationship with respect to $\dot{Q_i}$ and $\dot{M_o}$ and a mildly
+non-linear relationship with respect to $\dot{Q_i}$ and current season
+(note that this assumes southern hemisphere seasons). The relationship with respect
+to $T$ and windspeed, however, is strongly non-linear. This is because $\dot{Q_e}$
+and $\dot{M_e}$ are non-linear functions of windspeed and $\dot{Q_i}$. This further 
+complicates the inference of $\dot{Q_i}$.
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{figures/sensitivity_analysis.png}  
+    \caption{Sensitivity analysis of $\dot{T}$ with respect to $\dot{Q_i}$ 
+             and $\dot{M_o}$ (top panel), $\dot{Q_i}$ and season 
+             (center panel), and $T$ and windspeed (bottom panel). The model
+              parameters that have been held fixed are described in each 
+              panel's title. Warmer colors indicate a temperature increase
+              ($\dot{T} > 0$) and cooler colors a temperature decrease
+              ($\dot{T} < 0$)}
+      \label{sens}
+\end{figure}
+
+\subsection*{Inference}\label{inf}
+
+The objective of the inference problem is to estimate the rate at which heat
+is entering the lake from the hydrothermal system below.
+To constrain this we use the currently available observations of $T$, $M$,
+$X$, and $\dot{M_o}$ as well as numerical weather model data on windspeed.
+Putting it more concisely, we are looking for
+
+\begin{equation}\label{bsm}
+    p(\bm{x}_k|\bm{y}_{1:t})
+\end{equation}
+
+where $\bm{y}_{1:t}$ is the time series of $m$ observations and $x_k$ comprises
+the $n$ model parameters described in Section \ref{phm} at time step $k$. Note
+that $x_k$ also includes $T$, $M$, $\dot{M_o}$, and $X$. We distinguish here
+between the state of the lake and observations of this state
+($T$, $M$, $X$, $\dot{M_o}$) which are to some degree uncertain.
+
+The following equations are based on \cite{sarkkaBayesianFilteringSmoothing2013}
+which we refer the reader to for more proofs and detailed explanations.
+
+Equation \ref{bsm} is the marginal probability of $\bm{x}_k$ given all
+observations until time $t$. We assume that the states of the lake form a
+Markov chain, meaning that the state of the lake at time step $k$, $\bm{x}_k$,
+only depends on the state at adjacent time steps:
+
+\begin{equation}\label{markov1}
+    p(\bm{x}_k|\bm{x}_{1:k-1}, \bm{y}_{1:k-1}) = p(\bm{x}_k|\bm{x}_{k-1})
+\end{equation}
+
+and
+
+\begin{equation}\label{markov2}
+    p(\bm{x}_{k-1}|\bm{x}_{k:t}, \bm{y}_{k:t}) = p(\bm{x}_{k-1}|\bm{x}_{k})
+\end{equation}
+
+With this assumption and using \textit{Bayes' rule}, Equation \ref{bsm} can be
+rewritten as:
+
+\begin{equation}\label{bsm1}
+p(\bm{x}_k|\bm{y}_{1:T}) = p(\bm{x}_k|\bm{y}_{1:k})
+    \int\left[\frac{p(\bm{x}_{k+1}|\bm{x}_k)p(\bm{x}_{k+1}|\bm{y}_{1:T})}{p(\bm{x}_{k+1}|\bm{y}_{1:k})}\right]d\bm{x}_{k+1}
+\end{equation}
+
+with
+
+\begin{eqnarray}\label{bfilt1}
+p(\bm{x}_{k+1}|\bm{y}_{1:k}) & = & \int p(\bm{x}_{k+1}|\bm{x}_k)p(\bm{x}_k|\bm{y}_{1:k})d\bm{x}_k \\
+        \label{bfilt}
+    p(\bm{x}_k|\bm{y}_{1:k}) & = & \frac{1}{z_k} p(\bm{y}_k|\bm{x}_k)p(\bm{x}_k|\bm{y}_{1:k-1}) \\
+        \label{zetk}
+    z_k & = & \int p(\bm{y}_k|\bm{x}_k) p(\bm{x}_k| \bm{y}_{1:k-1}) d\bm{x}_k  
 \end{eqnarray}
-%
-Thus we observe that this expected value is finite for all $v>0$ (also see \cite{koon,khar,zvai,xjon,marg}).
+
+using the \textit{Chapman-Kolmogorov} equation we can write:
+
+\begin{equation}
+p(\bm{x}_k|\bm{y}_{1:k-1}) = \int p(\bm{x}_k|\bm{x}_{k-1})p(\bm{x}_{k-1}|\bm{y}_{1:k-1})d\bm{x}_{k-1} 
+\end{equation}
+
+Equations \ref{bsm1} to \ref{zetk} describe a recursive way to compute Equation
+\ref{bsm} in which we only need to know $\bm{x}_1$,
+$p(\bm{x}_k|\bm{x}_{k-1})$, and $p(\bm{y}_k|\bm{x}_k)$. Equation \ref{bsm1} and
+\ref{bfilt} are also known as the Bayesian Smoothing and Bayesian Filtering
+equation, respectively. If $p(\bm{x}_k|\bm{x}_{k-1})$ and
+$p(\bm{y}_k|\bm{x}_k)$ are linear transformations with normally distributed
+errors, the \textit{Kalman filter} \cite{Kalman1960} and Kalman smoother
+\cite{Rauch1965} represent closed form solutions. In our case, these
+transformations are:
+
+\begin{eqnarray}
+p(\bm{x}_k|\bm{x}_{k-1}) & = & f(\bm{x}_{k-1}) + \bm{q}_{k-1} \\
+    p(\bm{y}_k|\bm{x}_k) & = & \bm{x}_k[1:m] + \bm{r}_k   
+\end{eqnarray}
+
+where $f(\bm{x}_{k-1})$ is the physical model described in Section \ref{phm}
+which is strongly non-linear as shown in Figure \ref{sens}. If we assume the errors, $\bm{r}_k$ and
+$\bm{q}_{k-1}$, to be normally distributed ($\bm{q}_{k-1} \sim N(\bm{0},
+\bm{Q}_{k-1})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$), we can compute an
+approximate solution to \ref{bsm1} using the \textit{Unscented Kalman Filter}
+\citep{Merwe2004}.
+
+We estimate $\bm{R}_k$ from the daily variations in the observations but we do
+not know $\bm{x}_1$ and $\bm{Q}_{k-1}$. Setting $\bm{x}_1$ to an arbitrary value
+within physically reasonable bounds but large uncertainty works well in practice.
+While $\bm{Q}_{k}$ can be formally optimised we treat it here as a design
+parameter that we modify such that the observations are matched within their
+error bounds and the inferred heat input rate varies smoothly. The latter is
+motivated by the fact that the physical model described in \ref{phm} is not
+able to model short-term transient changes in lake volume and temperature.
+These can be caused by rain or increased meltwater influx and the assumption
+of a perfectly mixed lake typically does not apply in these situations.
+
+\section*{Results}
+\subsection*{Synthetic test}\label{syn_test}
+To evaluate the inference scheme we design a synthetic test, consisting of a very
+large cylinder, the same volume as Ruapehu Crater Lake. We keep the windspeed
+and surface water inflow constant at 4.5 m/s and 10 kt/day, respectively. To
+model the outflow we assume the outlet to be a pipe 0.2 $m^2$ in cross section
+and then use Bernoulli's equation to model the water outflow based on the water
+level in the cylinder above the pipe. We assume that the incoming heat follows
+a combination of a third-order polynomial and a sinusoid function and that the
+enthalpy of the incoming steam is 2.8 which corresponds to the enthalpy of
+saturated steam at hydrostatic pressures similar to those at the bottom of the
+lake \cite{Mayhew1978}. To the resulting synthetic observations we add
+uncertainties similar to those estimated from real observations (see Section
+\ref{realdata}).
+
+Figure \ref{syn_example} shows the result of the probabilistic inference of
+heat input compared to the true input of the synthetic model. The observations
+are well fit within their error bounds but we see some descrepancies between
+the original and the recovered heat input rates, especially for very steep
+changes in the input rate. This is the result of preferring a smooth solution
+as already mentioned in Section \ref{inf}.
+
+It is noteworthy that the mass influx and outflux rates are recovered
+quite accurately despite having no data in the former and very little data
+in the latter case.
+
+\begin{figure}
+	\includegraphics[width=\textwidth]{figures/synthetic_inversion_uks.png}  
+    \caption{Inversion result for the synthetic test described in Section 
+        \ref{syn_test}. Turquoise lines and markers represent hypothetical
+        observations; red dashed lines are the inversion results and black
+        dashed lines are the true input to the synthetic test.}
+  \label{syn_example}
+\end{figure}
+
+\subsection*{Real data}\label{realdata}
+Temperature at RCL is measured by a integrated-circuit temperature device
+(Texas Instruments LM35) located roughly 1.9 m below the lake's current
+overflow level. The water level sensor is a hydrostatic pressure sensor (First
+Sensor KTE/KTU/KTW6000\ldots CS Series) and it is co-located with the
+temperature sensor. Both sensors have a sampling interval of 15 minutes and
+transmit data several times a day via satellite to GNS Science, which monitors
+New Zealand's geohazards through its GeoNet project (www.geonet.org.nz
+ last accessed 16/07/2021).
+Water samples are taken 8-10 times a year from which subsequently, among many
+other analytes, concentrations of $Mg^{++}$ and $Cl^-$ are determined at GNS
+Science's Geothermal Analytical Laboratory. We will focus here on $Mg^{++}$
+concentration as it shows less deviation from the assumption of constant
+dilution than $Cl^-$.
+
+To estimate uncertainties of temperature and water level observations, we 
+compute their daily averages and standard deviations. Uncertainties of $Mg^{++}$
+concentrations are estimated directly when two samples were taken on the same day.
+Otherwise, their uncertainty is taken to be the median of directly estimated
+uncertainties. $\dot{M_o}$ is measured once or twice per year and we further know
+from field observations that the lake is not overflowing beneath a certain water
+level.
+
+As there is no weather station close to RCL we assume the wind speed from the
+New Zealand Metservice's numerical weather prediction model with a standard
+deviation of ??.
+
+\begin{figure}
+% 	\includegraphics[width=.9\textwidth]{clemb_inference_result}  
+% 	\caption{Inference results for the energy input rate (A),
+% 		the steam input rate (B), the water inflow (C) and 
+% 		outflow rate (D) for temperature, water level and 
+% 		$Mg^{++}$ observations at RCL in 2018.
+% 		The color shows an interpolation of 1D kernel density 
+% 		estimates (kdes) of the posterior samples for each day's 
+% 		inference result. The solid black line marks the expected
+% 		energy input rate. (E) shows the Bhattacharyya distance 
+% 	        between the probability distributions of the observations
+% 	        and the probability distributions of the inference results.}
+\label{inference_result}
+\end{figure}
+
+\section*{Discussion and conclusion}
+Although we use a tighter constraint on the
+outflow than just the dilution of $Mg^{++}$, inflow and outflow are still
+poorly constraint, especially when the lake level is above overflow level. As we already
+showed in the synthetic test in Section \ref{syn_test}, the $MG^{++}$ alone has
+too large uncertainties to properly constrain outflow rate. The probabilistic
+inference results thus help us determine future priorities for field
+observations and work is underway to establish a quantitative relationship
+between lake level and outflow rate.
+
+\cite{hurstCraterLakeEnergy2015} and \cite{Stevenson1991} both found that they required very
+high values of enthalpy for the incoming steam to avoid violating physical
+boundary conditions such as negative values for the inflow rate of meteoric
+water. Their explanation was the existence of a heat pipe beneath the lake where
+gas and fluid phases coexist. As the rising gas enters the lake relatively cool
+lake water enters the vent leading to an overall lower addition of mass to the
+lake and resulting in an enthalpy value that is double than expected for
+saturated steam at hydrostatic pressures likely to be found at the bottom of the
+lake. By including uncertainties in our observations we no longer require the
+high enthalpy values to satisfy the physical constraints. In fact, since we \textit{a
+priori} only allow physically reasonable parameter ranges we find the
+combinations of parameter values that reproduce the observations within their
+respective uncertainties.
+
+Inferring the heat input rate into a crater lake is a corner piece in estimating the
+total energy budget of a volcanic system as, for example, described by \cite{Adams1989a}
+for Po\'as volcano in Costa Rica. Similar to a volcano's gas budget, it can either
+indicate migration of magma or changes in permeability within the vent system. To
+answer whether it is one or the other requires more detailed numerical modelling
+of the magmatic and hydrothermal system. Many studies have demonstrated the 
+value of non-linear Kalman Filters for data assimilation with very complex numerical 
+models \citep{}. The next logical step, therefore, is to include models of magma
+degassing and hydrothermal fluid flow to model the heat input rate, volume, and
+enthalpy of the steam entering the lake. Because Kalman Filters return a probabilistic
+estimate of model parameters, even crude, first-order models could be used
+as long as appropriate model uncertainties are assigned.
+
+* probabilistic approach allows for Bayesian comparison between evaporation models
+
+Although an increased heat input rate at Ruapehu
+often goes along with an increase in seismic tremor levels (see Figure ??), this is
+not always the case and so it is an independent indicator of potential changes in
+the volcano.
+
+Figure ?? shows the comparison of daily averages of volcanic tremor amplitudes
+(RSAM; reference), heat input rate and water temperature. We can see that heat
+input rate correlates well with tremor values with a correlation coefficient of
+?. If we only consider heat pulses above ? MW (see dashed line in Figure ?), the
+correlation coefficient increases to ?. Although correlation does not imply
+causality it would seem plausible that high heat, and therefore gas, fluxes
+could generate volcanic tremor. A mechanism of heat flow generating tremor
+has been described by \citet{Iwamura2005}.
+
+In conclusion, the combination of the mass and energy model of \citet{Hurst2015}
+and \citet{Stevenson1992} combined with the probabilistic inference method we
+developed here provides a powerful method to continuously infer the heat input
+rate into Ruapehu Crater Lake. It allows us to combine disparate data streams,
+quantify our beliefs in the results and thereby augment our monitoring
+capabilities.
+
+\appendix
+\section{Surface losses}\label{A}
+\subsection{long-wavelength radiation}
+The net loss through long-wavelength radiation can be written as:
+\begin{equation}
+    Q_{rad} = A\sigma(\epsilon_w T_s^4 - \epsilon_a T_a^4)
+\end{equation}
+
+where
+
+\begin{tabular}{cp{8cm}}
+    A & is the surface of the lake (in $m^2$) \\
+    $\sigma$ & is the Stefan-Boltzmann coefficient ($5.67e^{-8} \frac{W}{m^2 K^4}$) \\
+    $\epsilon_w$ & is the lake emissivity \\
+    $T_s$ & is the water surface temperature (here assumed to be 1 $^{\circ}C$ less than the
+    measured temperature) \\
+    $\epsilon_a$ & is the effective atmospheric emissivity \\
+    $T_a$ & is the effective air temperature (here assumed to be 0.9 $^{\circ}C$) \\ 
+\end{tabular}
+
+\subsection{Evaporation}
+As proposed by \citet{Hurst2015} we use the equation of \citet{Adams1990} but
+replace the term for forced convection by that of \citet{Sartori2000}:
+
+\begin{equation}
+    Q_{evap} = [(2.2(T_{sv} - T_{av})^{1/3}(e_s - e_a))^2 + 
+    (L(0.00407u^{0.8}X_0^{-0.2})(e_s - e_a)/P_a)^2]^{1/2}
+\end{equation}
+
+where
+
+\begin{tabular}{cp{8cm}}
+    $T_{sv}$, $T_{av}$ & are the virtual surface and air temperatures respectively.
+    This correction accounts for the extra buoyancy of water vapour compared to air
+    (see Equation \ref{vtemp} for details) \\
+    $e_s$, $e_a$ & are the saturation vapour pressure and the ambient air vapour
+    pressure respectively \\
+    $L$ & is the latent heat of evaporation \\
+    $u$ & is the wind speed \\
+    $X_0$ & is the fetch, or characteristic length, of the lake\\
+    $P_a$ & is the atmospheric pressure \\
+\end{tabular}
+
+\begin{equation}\label{vtemp}
+T_{xv} = \frac{T_x}{(1-0.378e_x/P_a)} \qquad \mathrm{for}\quad x = s\ \mathrm{or}\ a
+\end{equation}
+
+\subsection{Sensible heat}
+This describes the heat loss due to conduction and convection above a hot lake and according to
+\citet{Stevenson1992} the ratio between sensible heat, $Q_{sh}$, and heat loss due
+to evaporation, $Q_{evap}$ can be written as:
+
+\begin{equation}
+    \frac{Q_{sh}}{Q_{evap}}=R_{sh}=\frac{\rho c_a}{L} \frac{(T_s - T_a)}{(e_s - e_a)}
+    \underbrace{\left [ \frac{(T_s-T_a)}{(T_{sv} - T_{av})} \right ]^{1/3}}_A 
+\end{equation}
+
+where
+
+\begin{tabular}{cp{8cm}}
+    $\rho$ & is the air density \\
+    $c_a$ & is the specific heat of air\\
+\end{tabular}
+
+We decided to ignore term A as it tends to be very close to 1.
+
+\subsection{Putting it all together}
+The final equation to compute surface heat losses is then:
+\begin{equation}
+    Q_e = Q_{rad} + Q_{evap}(1+R_{sh})
+\end{equation}
+
+and the mass-loss due to evaporation is:
+\begin{equation}
+    M_e = Q_{evap}/L
+\end{equation}
+
 %\nocite{oreg,schn,pond,smith,marg,hunn,advi,koha,mouse}

### 2021-08-02 src/pumahu/data.py
change default windspeed; fix caching

@@ -17,6 +17,14 @@ import xarray as xr
 
 from . import Forwardmodel, get_data
 
+def fits_hash(args, kwargs):
+    key = []
+    for _a in args[1:]:
+        key.append(_a)
+    key = tuple(key)    
+    key += tuple(sorted(kwargs.items()))
+    hashValue = hash(key)
+    return hashValue
 
 class LakeData:
     """
@@ -24,7 +32,7 @@ class LakeData:
     """
 
     def __init__(self, url="https://fits.geonet.org.nz/observation",
-                 csvfile=None, windspeed=4.5, m_out=10.,
+                 csvfile=None, windspeed=5., m_out=10.,
                  enthalpy=3.):
         """
         Parameters:
@@ -59,7 +67,7 @@ class LakeData:
 
 
     @cachier(stale_after=timedelta(weeks=2),
-             cache_dir='~/.cache')
+             cache_dir='~/.cache', hash_params=fits_hash)
     def get_data_fits(self, start, end, smoothing='kf'):
         """
         Request data from the FITS database unless it has been already cached.

### 2021-08-02 src/pumahu/visualise.py
add width to static export

@@ -197,7 +197,7 @@ def trellis_plot(data, data2=None, data2_params=None,
     fig.update_xaxes(showticklabels=False, ticks='inside')
     fig.update_xaxes(showticklabels=True, row=nparams)
     if filename is not None:
-        fig.write_image(file=filename)
+        fig.write_image(file=filename, width=1500)
     return fig
 
 

### 2021-08-02 manuscript/manuscript.tex
version sent to Steve and Tony

@@
 \startlocaldefs
-\newcommand{\bm}[1]{{\bf #1}}
+\newcommand{\bm}[1]{{\textbf{#1}}}
 \endlocaldefs
 
+%\newcommand{\bm}[1]{{\textbf{#1}}}
 %%% Begin ...
@@
 %%
-\section*{Introduction}
+\section{Introduction}
 
@@
 
-Crater Lake (Te Wai A-Moe) on Mt. Ruapehu, an active andesitic stratovolcano in
-the center of New Zealand's North Island (Figure \ref{overview}), is one such
-example with records of observations dating back to the 19th century
-\citep[e.g.][]{Friedlander1898}. Mt. Ruapehu erupted at least 602 times since
-1830 with many phreatic and phreato-magmatic eruptions and two major magmatic
-episodes \citep{Scott2013}. Although the latest magmatic event between June
-1995 to November 1997 and a subsequent dam collapse changed the shape and
-volume of the lake, it has been largely unchanged since then and the lake's
-bathymetry is well known from times when it was empty [Tony Hurst, personal
-communication].
+Ruapehu Crater Lake (Te Wai A-Moe in Te Reo Maori and referred to as RCL from
+hereon) on Mt. Ruapehu, an active andesitic stratovolcano in the center of New
+Zealand's North Island (Figure \ref{overview}), is one such example with records
+of observations dating back to the 19th century \cite{Friedlander1898}. Mt.
+Ruapehu erupted at least 602 times since 1830 with many phreatic and
+phreato-magmatic eruptions and two major magmatic episodes \citep{Scott2013}.
+Although the latest magmatic event between June 1995 to November 1997 and a
+subsequent dam collapse changed the shape and volume of the lake, it has been
+largely unchanged since then and the lake's bathymetry is well known from times
+when it was empty [Tony Hurst, personal communication].
 
-Increases in heat input may be caused by the ascent of magma or
-changes in the permeability of the hydrothermal system and are, therefore,
-important parameters to be monitored. First attempts to quantify the amount of
-heat entering the lake date back to \citep{Dibble1966}. Since 2016
-water temperature and lake level have been monitored continuously
-and water samples for ion concentration measurements have been taken roughly
-every two to three months. This enables us to continuously estimate heat input
-into the lake and, in principle, automatically detect changes as soon as they occur.
+First attempts to quantify the amount of heat entering the lake date back to
+1966 \citep{Dibble1966}. Since 2016 water temperature and lake level have been
+monitored continuously and water samples for ion concentration measurements have
+been taken roughly every two to three months. This enables us to continuously
+estimate heat input into the lake and, in principle, automatically detect
+changes as soon as they occur.
 
-Inference of the heat input is typically done through a mass and
-energy balance calculation: the sum of mass and energy flow into and out of the
-lake has to account for the observed changes in lake temperature, water level
-and ion dilution \citep[e.g.,][]{Hurst1981, Hurst1991, Stevenson1992,
-Fournier2009, Scott1994}. At Crater Lake, and probably at most other volcanic
-lakes, this is an underdetermined problem, that is the number of unknowns
-exceeds the number of constraints. Hence, it is necessary to constrain some of
-the parameters in the mass and energy balance model using prior knowledge. Further,
-some observations and input parameters come with large uncertainties and are sampled
-irregularly.
+Inference of the heat input is typically done through a mass and energy balance
+calculation: the sum of mass and energy flow into and out of the lake has to
+account for the observed changes in lake temperature, water level and ion
+dilution \cite{Hurst1981, Hurst1991, Stevenson1992, Fournier2009, Scott1994}. At
+RCL, and probably at most other volcanic lakes, this is an underdetermined
+problem, that is the number of unknowns exceeds the number of constraints.
+Hence, it is necessary to constrain some of the parameters in the mass and
+energy balance model using prior knowledge. Further, some observations and input
+parameters come with large uncertainties and are sampled irregularly.
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/overview.png}  
+\includegraphics[width=12 cm]{figures/overview.png}  
 \caption{Image of Mt. Ruapehu's crater lake from 2008 (\textcopyright Karen
@@
 
-\section*{Probabilistic inference method}\label{Pim}
+\section{Probabilistic inference method}\label{Pim}
 
-\subsection*{Physical model}\label{phm}
+\subsection{Physical model}\label{phm}
 
@@
 where $T$ is temperature and $c_p$ is the specific heat. For all following
-equations we take $H(T_0=0)$ as our reference enthalpy and $c_p=c_w=constant$,
-with $c_w$ the specific heat of water, such that the enthalpy of lake water at
-temperature $T$ can be written as $H(T) = c_wMT$. Because Ruapehu Crater Lake is
- at an altitude of $\sim$~2500~m and big parts of its catchment are covered by snow
-and ice all year round we assume the temperature of incoming water to be at 0
-$^{\circ}C$.
+equations we take $H(T_0=0~^{\circ}C)$ as our reference enthalpy and
+$c_p=c_w=constant$, with $c_w$ the specific heat of water, such that the
+enthalpy of lake water at temperature $T$ can be written as $H(T) = c_wMT$.
+Because RCL is at an altitude of $\sim$~2500~m and big parts of its catchment
+are covered by snow and ice all year round we assume the temperature of incoming
+water to be at 0 $^{\circ}C$.
 
@@
 
-where $T$ is the lake water temperature, $\dot{Q_r}$ is
-the rate of energy gain due to solar irradiation (short-wavelength radiation),
-and $c_wT\dot{M_o}$ is the rate of energy loss through outflow/seepage. $\dot{Q_e}$ is
-the rate of surface energy losses due to evaporation (forced and free
-convection), sensible heat, and long-wavelength radiation (net loss). Several
-empirical and theoretical equations exist to describe the surface losses and
-both \cite{Stevenson1992} and \cite{hurstCraterLakeEnergy2015} discuss them in detail. We
-follow here mostly the choice of \cite{hurstCraterLakeEnergy2015} as it includes the recent
-development by \cite{sartoriCriticalReviewEquations2000} on forced (i.e. wind-induced) convection
-(see Appendix \ref{A} for more details). $\dot{Q_{eff}}$ is the effective rate of
-energy gain from volcanic sources after heating incoming water:
+where $T$ is the lake water temperature, $\dot{Q_r}$ is the rate of energy gain
+due to solar irradiation (short-wavelength radiation), and $c_wT\dot{M_o}$ is
+the rate of energy loss through outflow/seepage. $\dot{Q_e}$ is the rate of
+surface energy losses due to evaporation (forced and free convection), sensible
+heat, and long-wavelength radiation (net loss). Several empirical and
+theoretical equations exist to describe the surface losses and we refer the
+reader to previous publications for more detailed discussions
+\cite{Stevenson1992, hurstCraterLakeEnergy2015}. There is no consensus on the
+best quantitative model of evaporation from warm lakes and, in fact, the most
+suited equation may depend on the location of and available observations from
+the lake. Appendix \ref{A} describes the equations used in this study in detail
+which include the recent development on forced (i.e. wind-induced) convection
+\cite{sartoriCriticalReviewEquations2000}.
+
+$\dot{Q_{eff}}$ is the effective rate of energy gain from volcanic sources after
+heating incoming water:
 \begin{equation}
@@
 \begin{figure}
-    \includegraphics[width=\textwidth]{figures/sensitivity_analysis.png}  
+    \includegraphics[width=12 cm]{figures/sensitivity_analysis.png}  
     \caption{Sensitivity analysis of $\dot{T}$ with respect to $\dot{Q_i}$ 
@@
 
-\subsection*{Inference}\label{inf}
+\subsection{Inference}\label{inf}
 
 The objective of the inference problem is to estimate the rate at which heat
-is entering the lake from the hydrothermal system below.
+is entering the lake from the hydrothermal system below ($\dot{Q_i}$).
 To constrain this we use the currently available observations of $T$, $M$,
-$X$, and $\dot{M_o}$ as well as numerical weather model data on windspeed.
-Putting it more concisely, we are looking for
+$X$, and $\dot{M_o}$. Putting it more concisely, we are looking for
 
@@
 
-The following equations are based on \cite{sarkkaBayesianFilteringSmoothing2013}
-which we refer the reader to for more proofs and detailed explanations.
+Solving Equation \ref{bsm} using standard strategies like Markov Chain Monte
+Carlo is computationally very expensive as the number of variables grows
+exponentially with the length of the timeseries. An alternative are Bayesian
+Filtering and Smoothing algorithms of which the Kalman Filter and Smoother are
+the closed form solution for linear models \cite{Kalman1960, Rauch1965,
+sarkkaBayesianFilteringSmoothing2013}.
 
-Equation \ref{bsm} is the marginal probability of $\bm{x}_k$ given all
+Equation \ref{bsm} is the conditional probability of $\bm{x}_k$ given all
 observations until time $t$. We assume that the states of the lake form a
@@
 With this assumption and using \textit{Bayes' rule}, Equation \ref{bsm} can be
-rewritten as:
+rewritten \cite{sarkkaBayesianFilteringSmoothing2013} as:
 
 \begin{equation}\label{bsm1}
-p(\bm{x}_k|\bm{y}_{1:T}) = p(\bm{x}_k|\bm{y}_{1:k})
-    \int\left[\frac{p(\bm{x}_{k+1}|\bm{x}_k)p(\bm{x}_{k+1}|\bm{y}_{1:T})}{p(\bm{x}_{k+1}|\bm{y}_{1:k})}\right]d\bm{x}_{k+1}
+p(\bm{x}_k|\bm{y}_{1:t}) = p(\bm{x}_k|\bm{y}_{1:k})
+    \int\left[\frac{p(\bm{x}_{k+1}|\bm{x}_k)p(\bm{x}_{k+1}|\bm{y}_{1:t})}{p(\bm{x}_{k+1}|\bm{y}_{1:k})}\right]d\bm{x}_{k+1}
 \end{equation}
@@
 
-using the \textit{Chapman-Kolmogorov} equation we can write:
+using the \textit{Chapman-Kolmogorov} equation
+\cite{sarkkaBayesianFilteringSmoothing2013} we can write:
 
@@
 where $f(\bm{x}_{k-1})$ is the physical model described in Section \ref{phm}
-which is strongly non-linear as shown in Figure \ref{sens}. If we assume the errors, $\bm{r}_k$ and
-$\bm{q}_{k-1}$, to be normally distributed ($\bm{q}_{k-1} \sim N(\bm{0},
-\bm{Q}_{k-1})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$), we can compute an
+propagated forward in time using a fourth-order Runge-Kutta scheme. If we 
+assume the errors, $\bm{r}_k$ and $\bm{q}_{k}$, to be normally distributed
+($\bm{q}_{k} \sim N(\bm{0},
+\bm{Q}_{k})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$), we can compute an
 approximate solution to \ref{bsm1} using the \textit{Unscented Kalman Filter}
-\citep{Merwe2004}.
+\citep{Merwe2004, sarkkaBayesianFilteringSmoothing2013}.
 
 We estimate $\bm{R}_k$ from the daily variations in the observations but we do
-not know $\bm{x}_1$ and $\bm{Q}_{k-1}$. Setting $\bm{x}_1$ to an arbitrary value
-within physically reasonable bounds but large uncertainty works well in practice.
-While $\bm{Q}_{k}$ can be formally optimised we treat it here as a design
-parameter that we modify such that the observations are matched within their
-error bounds and the inferred heat input rate varies smoothly. The latter is
-motivated by the fact that the physical model described in \ref{phm} is not
-able to model short-term transient changes in lake volume and temperature.
-These can be caused by rain or increased meltwater influx and the assumption
-of a perfectly mixed lake typically does not apply in these situations.
+not know $\bm{x}_1$ and $\bm{Q}_{k}$. Setting $\bm{x}_1$ to an arbitrary value
+within physically reasonable bounds but large uncertainty works well in
+practice. While $\bm{Q}_{k}$ can be formally optimised we treat it here as a
+design parameter that we modify such that the observations are matched within
+their error bounds and the inferred heat input rate varies smoothly. The latter
+is motivated by the fact that the physical model described in \ref{phm} is not
+able to model short-term transient changes in lake volume and temperature. These
+can be caused by rain or increased meltwater influx and the underlying
+assumption of a perfectly mixed lake typically does not apply in these
+situations.
 
-\section*{Results}
-\subsection*{Synthetic test}\label{syn_test}
-To evaluate the inference scheme we design a synthetic test, consisting of a very
-large cylinder, the same volume as Ruapehu Crater Lake. We keep the windspeed
-and surface water inflow constant at 4.5 m/s and 10 kt/day, respectively. To
-model the outflow we assume the outlet to be a pipe 0.2 $m^2$ in cross section
-and then use Bernoulli's equation to model the water outflow based on the water
-level in the cylinder above the pipe. We assume that the incoming heat follows
-a combination of a third-order polynomial and a sinusoid function and that the
-enthalpy of the incoming steam is 2.8 which corresponds to the enthalpy of
+\section{Results}
+\subsection{Synthetic test}\label{syn_test}
+
+To evaluate the inference scheme we design a synthetic test, consisting of a
+very large cylinder, the same volume as RCL. We keep the windspeed and surface
+water inflow constant at 4.5 m/s and 10 kt/day, respectively. To model the
+outflow we assume the outlet to be a pipe 0.2 $m^2$ in cross section and then
+use Bernoulli's equation to model the water outflow based on the water level in
+the cylinder above the pipe. We assume that the incoming heat follows a
+combination of a third-order polynomial and a sinusoid function and that the
+enthalpy of the incoming steam is 2.8 MJ/kg which corresponds to the enthalpy of
 saturated steam at hydrostatic pressures similar to those at the bottom of the
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{figures/synthetic_inversion_uks.png}  
+	\includegraphics[width=12 cm]{figures/synthetic_inversion_uks.png}  
     \caption{Inversion result for the synthetic test described in Section 
@@
 
-\subsection*{Real data}\label{realdata}
-Temperature at RCL is measured by a integrated-circuit temperature device
+\subsection{Real data}\label{realdata}
+Temperature at RCL is measured by an integrated-circuit temperature device
 (Texas Instruments LM35) located roughly 1.9 m below the lake's current
@@
 Otherwise, their uncertainty is taken to be the median of directly estimated
-uncertainties. $\dot{M_o}$ is measured once or twice per year and we further know
-from field observations that the lake is not overflowing beneath a certain water
-level.
+uncertainties.
 
-As there is no weather station close to RCL we assume the wind speed from the
-New Zealand Metservice's numerical weather prediction model with a standard
-deviation of ??.
+$\dot{M_o}$ is measured once or twice per year and we further know from field
+observations that the lake is not overflowing beneath a certain water level. We
+fit a sigmoid function to these observations assuming a maximum outflow rate of
+250 l/s. We assume the uncertainty in $\dot{M_o}$ as the 95 percent confidence
+interval of the curve-fitting.
+
+As there is no permanent weather station close to RCL we assume the wind speed
+to be 5 m/s. This was the average windspeed derived from a temporary deployment
+of a weather buoy on RCL and adjacent permanent weather stations
+\cite{hurstUseWeatherBuoy2012}.
+
+Figure \ref{inference_result} shows $\dot{Q_i}$ for RCL for lake measurements
+between 4 March 2016 and 7 July 2021. During this time the heat input rate
+reached a maximum of 753 MW and was on average 165 MW. The average standard
+deviation was 34 MW and generally increased for lower values of $\dot{Q_i}$. To
+put this in context, currently installed geothermal power plants in New Zealand
+have a combined output of $\sim$~1000MW.
 
 \begin{figure}
-% 	\includegraphics[width=.9\textwidth]{clemb_inference_result}  
-% 	\caption{Inference results for the energy input rate (A),
-% 		the steam input rate (B), the water inflow (C) and 
-% 		outflow rate (D) for temperature, water level and 
-% 		$Mg^{++}$ observations at RCL in 2018.
-% 		The color shows an interpolation of 1D kernel density 
-% 		estimates (kdes) of the posterior samples for each day's 
-% 		inference result. The solid black line marks the expected
-% 		energy input rate. (E) shows the Bhattacharyya distance 
-% 	        between the probability distributions of the observations
-% 	        and the probability distributions of the inference results.}
+ 	\includegraphics[width=13 cm]{figures/uks_real_data.png}  
+ 	\caption{Inference results for the heat input rate $\dot{Q_i}$ at RCL
+             between 4 March 2016 and 7 July 2021. The solid line shows the
+             marginal probability and the uncertainty is displayed as shaded
+             region.}
 \label{inference_result}
@@
 
-\section*{Discussion and conclusion}
-Although we use a tighter constraint on the
-outflow than just the dilution of $Mg^{++}$, inflow and outflow are still
-poorly constraint, especially when the lake level is above overflow level. As we already
-showed in the synthetic test in Section \ref{syn_test}, the $MG^{++}$ alone has
-too large uncertainties to properly constrain outflow rate. The probabilistic
-inference results thus help us determine future priorities for field
-observations and work is underway to establish a quantitative relationship
-between lake level and outflow rate.
+\section{Discussion and conclusion}
+Inferring the heat input rate into a crater lake is an important part in
+estimating the total energy budget of a volcanic system \cite{Brown1989a}.
+Similar to a volcano's gas budget, it can either indicate migration of magma or
+changes in permeability within the vent system.
 
-\cite{hurstCraterLakeEnergy2015} and \cite{Stevenson1991} both found that they required very
+Figure \ref{psd} shows the power spectral density (PSD) for the marginal heat
+input rate shown in Figure \ref{inference_result}. It was computed using the
+multi taper method \cite{prietoComputersGeosciencesFortran2009} which provides
+not only an estimate of the PSD but also its uncertainty. The PSD shows that the
+dominant periods in the heat input rate lie between $\sim$100 and $\sim$250
+days. Whether these periods are due to cyclic processes in the magmatic system,
+the hydrothermal system, or a combination of both requires further modelling of
+these processes. Many studies have demonstrated the value of non-linear Kalman
+Filters for inference and data-assimilation with non-linear, high-dimension
+numerical models \cite{White2018a}. While beyond the scope of this study, the
+next logical step, therefore, is to include numerical models of magma degassing
+and hydrothermal fluid flow to model the heat input rate, volume, and enthalpy
+of the steam entering the lake.
+
+\begin{figure}
+    \includegraphics[width=12 cm]{figures/psd_analysis.png}  
+ 	\caption{Power spectral density of the heat input rate 
+             (Figure \ref{inference_result}) shown as a solid line
+             with uncertainty displayed as shaded region.}
+    \label{psd}
+\end{figure}
+
+Irrespective of the process that is responsible for the periodicity, it appears
+that it has not changed significantly over the past five years. This may
+suggest that the volcano has been in a dynamic equilibrium during this time.
+This is further supported by the cumulative heat input (Figure \ref{cumheat}).
+It shows a slight increase in heat input in mid-2017 but a fairly constant rate
+since then. Of course this does not in any way forecast the volcano's future
+behaviour but it may serve as a useful baseline to detect changes in the
+volcano's state more easily.
+
+\begin{figure}
+    \includegraphics[width=12 cm]{figures/cumulative_heat.png}
+    \caption{Cumulative heat input into RCL (solid line) and its 
+    uncertainty (shaded region).}
+    \label{cumheat}
+\end{figure}
+
+Previous studies of RCL's mass and energy balance found that they required very
 high values of enthalpy for the incoming steam to avoid violating physical
 boundary conditions such as negative values for the inflow rate of meteoric
-water. Their explanation was the existence of a heat pipe beneath the lake where
-gas and fluid phases coexist. As the rising gas enters the lake relatively cool
-lake water enters the vent leading to an overall lower addition of mass to the
-lake and resulting in an enthalpy value that is double than expected for
-saturated steam at hydrostatic pressures likely to be found at the bottom of the
-lake. By including uncertainties in our observations we no longer require the
-high enthalpy values to satisfy the physical constraints. In fact, since we \textit{a
-priori} only allow physically reasonable parameter ranges we find the
-combinations of parameter values that reproduce the observations within their
-respective uncertainties.
+water \cite{hurstCraterLakeEnergy2015, Stevenson1992}. Their explanation was the
+existence of a heat pipe beneath the lake where gas and fluid phases coexist. As
+the rising gas enters the lake relatively cool lake water enters the vent
+leading to an overall lower addition of mass to the lake and resulting in an
+enthalpy value that is double than expected for saturated steam at hydrostatic
+pressures likely to be found at the bottom of the lake. By including
+uncertainties in our observations we no longer require these high enthalpy
+values to satisfy the physical constraints. Numerical modelling of hydrothermal
+processes may also shed further light on the heating processes and whether or
+not a heat pipe is a plausible mechanism.
 
-Inferring the heat input rate into a crater lake is a corner piece in estimating the
-total energy budget of a volcanic system as, for example, described by \cite{Adams1989a}
-for Po\'as volcano in Costa Rica. Similar to a volcano's gas budget, it can either
-indicate migration of magma or changes in permeability within the vent system. To
-answer whether it is one or the other requires more detailed numerical modelling
-of the magmatic and hydrothermal system. Many studies have demonstrated the 
-value of non-linear Kalman Filters for data assimilation with very complex numerical 
-models \citep{}. The next logical step, therefore, is to include models of magma
-degassing and hydrothermal fluid flow to model the heat input rate, volume, and
-enthalpy of the steam entering the lake. Because Kalman Filters return a probabilistic
-estimate of model parameters, even crude, first-order models could be used
-as long as appropriate model uncertainties are assigned.
-
-* probabilistic approach allows for Bayesian comparison between evaporation models
-
-Although an increased heat input rate at Ruapehu
-often goes along with an increase in seismic tremor levels (see Figure ??), this is
-not always the case and so it is an independent indicator of potential changes in
-the volcano.
-
-Figure ?? shows the comparison of daily averages of volcanic tremor amplitudes
-(RSAM; reference), heat input rate and water temperature. We can see that heat
-input rate correlates well with tremor values with a correlation coefficient of
-?. If we only consider heat pulses above ? MW (see dashed line in Figure ?), the
-correlation coefficient increases to ?. Although correlation does not imply
-causality it would seem plausible that high heat, and therefore gas, fluxes
-could generate volcanic tremor. A mechanism of heat flow generating tremor
-has been described by \citet{Iwamura2005}.
-
-In conclusion, the combination of the mass and energy model of \citet{Hurst2015}
-and \citet{Stevenson1992} combined with the probabilistic inference method we
-developed here provides a powerful method to continuously infer the heat input
-rate into Ruapehu Crater Lake. It allows us to combine disparate data streams,
-quantify our beliefs in the results and thereby augment our monitoring
-capabilities.
+In conclusion, the combination of the mass and energy model, developed in
+previous studies \cite{hurstCraterLakeEnergy2015, Stevenson1992}, with the
+probabilistic inference method we developed here provides a powerful method to
+continuously infer the heat input rate into RCL. It allows us to combine
+disparate data streams, quantify our beliefs in the results and thereby augment
+our monitoring capabilities by keeping a finger on Mt. Ruapehu's energy budget.
 
@@
 where
-
+\begin{table}[h!]
 \begin{tabular}{cp{8cm}}
@@
 \end{tabular}
+\end{table}
 
 \subsection{Evaporation}
-As proposed by \citet{Hurst2015} we use the equation of \citet{Adams1990} but
-replace the term for forced convection by that of \citet{Sartori2000}:
+As proposed by \cite{hurstCraterLakeEnergy2015} we use the equation of
+\cite{adamsEvaporationHeatedWater1990} but replace the term for forced
+convection by that of \cite{sartoriCriticalReviewEquations2000}:
 
@@
 \end{tabular}
+\end{table}
 
@@
 \subsection{Sensible heat}
-This describes the heat loss due to conduction and convection above a hot lake and according to
-\citet{Stevenson1992} the ratio between sensible heat, $Q_{sh}$, and heat loss due
-to evaporation, $Q_{evap}$ can be written as:
+This describes the heat loss due to conduction and convection above a hot lake
+and according to \cite{Stevenson1992} the ratio between sensible heat,
+$Q_{sh}$, and heat loss due to evaporation, $Q_{evap}$ can be written as:
 
@@
 
+\begin{table}[h!]
 \begin{tabular}{cp{8cm}}

### 2021-08-02 manuscript/figures/cumulative_heat.py
scripts to produce plots for manuscript

@@ -1,0 +1,40 @@
+import plotly.graph_objects as go
+import numpy as np
+import xarray as xr
+import pandas as pd
+
+fn = './data/uks.nc'
+xdf = xr.open_dataset(fn)
+data = xdf.exp.loc[:, 'q_in', 'val'].values
+sigma = xdf.exp.loc[:, 'q_in', 'std'].values
+# Convert MW to TJ by multiplying with 86400 s
+data *= 0.864
+sigma *= 0.864
+# Convert to kWh by deviding with 3.6e6 J
+data /= 3.6
+sigma /= 3.6
+cum_mean = np.cumsum(data)
+cum_std = np.sqrt(np.cumsum(sigma))
+dates = pd.to_datetime(xdf.dates.values)
+fig = go.Figure()
+fig.add_trace(go.Scatter(x=dates, y=cum_mean, mode='lines',
+                         showlegend=False))
+fig.add_trace(go.Scatter(name='Upper Bound',
+                         x=dates,
+                         y=cum_mean + 3*cum_std,
+                         mode='lines',
+                         marker=dict(color="#444"),
+                         line=dict(width=0),
+                         showlegend=False))
+fig.add_trace(go.Scatter(name='Lower Bound',
+                         x=dates,
+                         y=cum_mean - 3*cum_std,
+                         mode='lines',
+                         marker=dict(color="#444"),
+                         line=dict(width=0),
+                         fillcolor='rgba(68, 68, 68, 0.3)',
+                         fill='tonexty',
+                         showlegend=False))
+fig.update_yaxes(title='$10^6 kWh$')
+fig.update_layout(font_size=22)
+fig.write_image('cumulative_heat.png', width=1500, scale=2)

### 2021-08-02 manuscript/figures/psd_analysis.py

@@ -1,0 +1,42 @@
+import plotly.graph_objects as go
+import numpy as np
+from mtspec import mtspec
+from scipy import signal
+import xarray as xr
+
+fn = 'data/uks.nc'
+xdf = xr.open_dataset(fn)
+data = xdf.exp.loc[:, 'q_in', 'val'].values
+data -= data.mean()
+spec, freq, jackknife, _, _ = mtspec(data=data, delta=1.0, time_bandwidth=3.5,
+    number_of_tapers=10, nfft=2048, statistics=True)
+idx = freq<0.2
+freq = np.r_[1./freq[1:], 0]
+
+fig = go.Figure()
+fig.add_trace(go.Scatter(x=freq, y=spec, name='Multi taper', mode='lines',
+                         showlegend=False))
+if False:
+    freqs, psd = signal.welch(data, nfft=2048, nperseg=256)
+    freqs = np.r_[1./freqs[1:], 0]
+    fig.add_trace(go.Scatter(x=freqs, y=psd, mode='lines', line_dash='dash',
+                             name='Welch'))
+fig.add_trace(go.Scatter(name='Upper Bound',
+                         x=freq,
+                         y=jackknife[:, 1],
+                         mode='lines',
+                         marker=dict(color="#444"),
+                         line=dict(width=0),
+                         showlegend=False))
+fig.add_trace(go.Scatter(name='Lower Bound',
+                         x=freq,
+                         y=jackknife[:, 0],
+                         mode='lines',
+                         marker=dict(color="#444"),
+                         line=dict(width=0),
+                         fillcolor='rgba(68, 68, 68, 0.3)',
+                         fill='tonexty',
+                         showlegend=False))
+fig.update_xaxes(type="log", title='Period [days]')
+fig.update_layout(font_size=22)
+fig.write_image('psd_analysis.png', width=1500, scale=2)

### 2021-08-02 manuscript/figures/synthetic_inversion_uks.py

@@ -1,0 +1,31 @@
+from collections import OrderedDict
+
+import numpy as np
+
+from pumahu.syn_model import (SynModel,
+                              setup_realistic,
+                              resample,
+                              make_sparse)
+from pumahu.visualise import trellis_plot
+from pumahu.uks import UnscentedKalmanSmoother
+
+
+xds = SynModel().run(setup_realistic(sinterval=120), addnoise=True)
+
+na = resample(xds)
+na = make_sparse(na, ['m_out', 'X'])
+Q = OrderedDict(T=5e-1, M=1e-1, X=1e-3, q_in=1e1,
+                m_in=1e1, m_out=1e1, h=1e-3, W=1e-3,
+                dqi=1e2, dMi=0, dMo=0, dH=0, dW=0)
+Q = np.eye(len(Q))*list(Q.values())
+
+uks = UnscentedKalmanSmoother(data=na, Q=Q,
+                              initvals={'qi':200., 'm_in': 0,
+                                        'm_out':0., 'X':3.})
+xds_uks = uks(test=True, smooth=True, alpha=0.35)
+fig = trellis_plot(xds_uks, data2=xds,
+                   data2_params=['q_in', 'm_in', 'm_out'],
+                   filled_error=True)
+fig.update_layout(font_size=20, height=1200, showlegend=False)
+filename="synthetic_inversion_uks.png"
+fig.write_image(filename, scale=2)

### 2021-08-02 manuscript/figures/uks_real_data.py

@@ -1,0 +1,15 @@
+import os
+import xarray as xr
+from pumahu.visualise import plot_qin_uks
+from pumahu.uks import main as main_uks
+
+fn = 'data/uks.nc'
+if not os.path.isfile(fn):
+    main_uks(['--rdir', './data', '-s', '2016-03-04',
+              '-e', '2021-07-07', '-f'])
+
+xdf = xr.open_dataset('data/uks.nc')
+fig = plot_qin_uks(xdf)
+fig.update_layout(showlegend=False,
+                  font_size=22)
+fig.write_image('uks_real_data.png', width=1500)

### 2021-08-03 buildnrun.sh
bump version to 0.1

@@ -6,7 +6,7 @@
 #########################################
 
 IMAGE=pumahu
-TAG=0.0.2
+TAG=0.1
 BUILD=false
 PUSH=false
 INTERACTIVE=false

### 2021-08-03 docker-deploy.sh

@@ -1,7 +1,7 @@
 #!/bin/bash
 
 ######## Set up default project variables #######
-IMAGE='huta17-d.gns.cri.nz:5000/yannik/pumahu:0.0.2'
+IMAGE='huta17-d.gns.cri.nz:5000/yannik/pumahu:0.1'
 APP_NAME=Pumahu
 APP_SERVER="Vulkan"
 TEAM="Volcano"

### 2021-08-03 setup.py

@@ -9,7 +9,7 @@ from setuptools import setup
 # do the build/install
 setup(
     name="pumahu",
-    version="0.0.4",
+    version="0.1",
     description="Python package to compute crater lake energy and mass balance.",
     long_description="Python package to compute crater lake energy and mass balance.",
     author="Yannik Behr",

### 2022-02-03 src/pumahu/uks.py
move more functionality into data; add m_in to modelling

@@ -19,7 +19,7 @@ from . import get_data
 from .syn_model import SynModel
 from .forward_model import Forwardmodel
 from .sigma_points import MerweScaledSigmaPoints
-from .data import LakeData
+from .data import LakeData, get_outflow
 from .visualise import trellis_plot, plot_qin_uks
  
 import ipdb
@@ -65,25 +65,19 @@ def h_x(x):
     """
     Measurement function
     """
-    return [x[0], x[1], x[2], x[5], x[6], x[7]]
+    return [x[0], x[1], x[2], x[4], x[5], x[6], x[7]]
 
 
 class UnscentedKalmanSmoother:
 
-    def __init__(self, data, X0=None, P0=None, Q=None,
-                 initvals={'qi':200., 'm_in': 10, 'm_out':10., 'X':2.}):
+    def __init__(self, data, X0=None, P0=None, Q=None):
         self.dates = pd.to_datetime(data['dates'].values)
         self.ndates = len(self.dates)
         self.dt = (self.dates[1] - self.dates[0])/pd.Timedelta('1D')
         self.params = ['T', 'M', 'X', 'q_in', 'm_in', 'm_out', 'h', 'W']
         self.nparams = len(self.params)
-        try:
-            self.data = data.loc[:, self.params, :]
-            (T, M, X, Mi, Mo, qi, H, W) = self.data.loc[:,  self.params, 'val'].values[0]
-        except KeyError:
-            self.data = data.loc[:, ['T', 'M', 'X', 'm_out', 'h', 'W'], :]
-            (T, M, X, Mo, H, W) = self.data.loc[:, ['T', 'M', 'X', 'm_out', 'h', 'W'], 'val'].values[0]
-            qi, Mi, Mo, X = list(initvals.values())
+        self.data = data.loc[:, self.params, :]
+        (T, M, X, Mi, Mo, qi, H, W) = self.data.loc[:,  self.params, 'val'].values[0]
             
         dqi = 1e-1
         dMi = 1e-1
@@ -148,7 +142,7 @@ class UnscentedKalmanSmoother:
         points = MerweScaledSigmaPoints(n=self.nvar, alpha=alpha, beta=beta,
                                         kappa=kappa)
         F = Fx(self.nparams, self.ngrads, test=test)
-        kf = UnscentedKalmanFilter(dim_x=self.nvar, dim_z=5, dt=self.dt, fx=F.run,
[…]

### 2022-02-03 src/pumahu/data.py
add m_in and q_in to data; make get_outflow and get_wind standalone functions

@@ -38,7 +38,8 @@ class LakeData:
     """
 
     def __init__(self, url="https://fits.geonet.org.nz/observation",
-                 csvfile=None, windspeed=5., m_out=10., m_out_err=.25,
+                 csvfile=None, windspeed=5., m_in=np.nan, m_in_err=np.nan,
+                 q_in=np.nan, q_in_err=np.nan, m_out=10., m_out_err=.25,
                  enthalpy=3.):
         """
         Parameters:
@@ -58,13 +59,17 @@ class LakeData:
         self.xdf = None
         self.ws = windspeed
         self.ws_err = 0.5
+        self.m_in = m_in
+        self.m_in_err = m_in_err 
         self.m_out = m_out
         self.m_out_err = m_out_err 
+        self.q_in = q_in
+        self.q_in_err = q_in_err 
         self.h = enthalpy
         self.h_err = 0.01
         self.prms = ['T', 'z', 'Mg', 'X', 'V',
                      'A', 'p', 'M', 'dV', 'h', 
-                     'W', 'm_out']
+                     'W', 'm_in', 'm_out', 'q_in']
         if csvfile is None:
             self.get_raw_data = self.FITS_request
         else:
@@ -75,7 +80,8 @@ class LakeData:
     @cachier(stale_after=timedelta(weeks=2),
              cache_dir='~/.cache', hash_params=fits_hash)
     def get_data(self, start, end, smoothing='kf',
-                 nerrsamples=1000):
+                 nerrsamples=1000,
+                 initvals={'q_in':200.,'m_in':10,'m_out':10.,'X':2.}):
         """
         Request data from the FITS database or a csv file.
         
@@ -138,12 +144,21 @@ class LakeData:
         result[:, self.prms.index('W'), 1] = np.ones(dsize)*self.ws_err
         result[:, self.prms.index('h'), 0] = np.ones(dsize)*self.h
         result[:, self.prms.index('h'), 1] = np.ones(dsize)*self.h_err
+        result[:, self.prms.index('m_in'), 0] = np.ones(dsize)*self.m_in
+        result[:, self.prms.index('m_in'), 1] = np.ones(dsize)*self.m_in_err
         result[:, self.prms.index('m_out'), 0] = np.ones(dsize)*self.m_out
         result[:, self.prms.index('m_out'), 1] = np.ones(dsize)*self.m_out_err
+        result[:, self.prms.index('q_in'), 0] = np.ones(dsize)*self.q_in
[…]

### 2022-02-03 src/pumahu/syn_model.py
add m_in and q_in to output; add remove_inputs

@@ -193,7 +193,7 @@ class SynModel:
         exp = np.zeros((nsteps, nprms, 2))
         error = {'T': 0.4, 'z': 0.01, 'M': 2.,
                  'X': 0.4, 'A': 30., 'V': 2.0,
-                 'Mg': 50, 'm_out': 5, 'W':.5,
+                 'Mg': 50, 'm_in': 5., 'm_out': 5, 'W':.5,
                  'h': .01}
  
         V = 8800
@@ -255,7 +255,7 @@ class SynModel:
         return xds
 
 
-def resample(xds, parameters=['T', 'M', 'X', 'm_out', 'z', 'W', 'h'], sinterval='1D'):
+def resample(xds, parameters=['T', 'M', 'X', 'q_in', 'm_in', 'm_out', 'z', 'W', 'h', 'Mg'], sinterval='1D'):
     """
     Resample a synthetic model to a given sampling interval.
     
@@ -276,12 +276,13 @@ def resample(xds, parameters=['T', 'M', 'X', 'm_out', 'z', 'W', 'h'], sinterval=
     mn = nxds.mean().to_array().values
     st = nxds.std().to_array().values
     new_data = np.concatenate((mn[:,:,:,np.newaxis], st[:,:,:,np.newaxis]), axis=3)
-    rxds = xr.DataArray(new_data[0], dims=('dates', 'parameters', 'val_std'),
-                        coords=(nxds.mean().dates, parameters,['val', 'std']) )
     # fill fields with zero variance with the average
     # variance for the respective parameter
-    idx = np.where(rxds.loc[dict(val_std='std')] == 0.)
-    rxds[idx[0], idx[1], 1] = rxds[:, :, 1].mean(axis=0)
+    idx = np.where(new_data[0, :, :, 1] == 0.)
+    std_mn = new_data[0, :, :, 1].mean(axis=0)
+    new_data[0, idx[0], idx[1], 1] = std_mn[idx[1]]
+    rxds = xr.DataArray(new_data[0], dims=('dates', 'parameters', 'val_std'),
+                        coords=(nxds.mean().dates, parameters,['val', 'std']) )
     return rxds
 
 
@@ -305,10 +306,29 @@ def make_sparse(xds, parameters, perc=0.8, seed=42):
     """
     orig = xds.loc[:, parameters, ['val', 'std']].values
     rs = np.random.default_rng(42)
[…]

### 2022-02-03 src/pumahu/visualise.py
minor cosmetic changes

@@ -84,7 +84,7 @@ class TrellisPlot:
 
     def plot_trace(self, fig, data, key, ylabel, row, filled_error=True,
                    plotvars=[('input', 'Input'),('exp', 'Result')],
-                   dropzeros=True):
+                   dropzeros=True, showerror=True):
         """
         :param plotvars: Variables to plot.
         """
@@ -114,7 +114,7 @@ class TrellisPlot:
                                              name=name,
                                              showlegend=showlegend),
                                   row=row, col=1)
-                if len(traces) > 1:
+                if len(traces) > 1 and showerror:
                     if np.any(np.isnan(traces[0])):
                         error_y = (traces[2] - traces[1])/2.
                         fig.add_trace(go.Scatter(x=dates, y=traces[0],
@@ -158,7 +158,8 @@ class TrellisPlot:
 
 
 def trellis_plot(data, data2=None, data2_params=None,
-                 filename=None, **kwds):
+                 filename=None, data2_showerror=True,
+                 **kwds):
     """
     A trellis plot for inversion results.
 
@@ -174,7 +175,7 @@ def trellis_plot(data, data2=None, data2_params=None,
     params = list(data.parameters.values)
     nparams = len(params)
     labels = dict(q_in='Qi [MW]', m_in='Mi [kt/day]',
-                  m_out='Mo [kt/day]', T='T [C]',
+                  m_out='Mo [kt/day]', T='T [C]',W='W [m/s]',
                   M='M [kt]', X='X [kt]', h='H [MJ/kg]')
     rowdict = {}
     with plt.style.context('bmh'):
@@ -182,7 +183,12 @@ def trellis_plot(data, data2=None, data2_params=None,
                             vertical_spacing=.01)
         tp = TrellisPlot()
         for i, p in enumerate(params):
-            tp.plot_trace(fig, data, p, labels.get(p, p), row=i+1, **kwds)
+            # Don't show q_in as a data input
+            plotvars=[('input', 'Input'),('exp', 'Result')]
+            if p == 'q_in':
+                plotvars=[('NaN', 'NaN'),('exp', 'Result')]
+            tp.plot_trace(fig, data, p, labels.get(p, p),
+                          plotvars=plotvars, row=i+1, **kwds)
             rowdict[p] = i+1
 
         if data2 is not None:
@@ -192,7 +198,8 @@ def trellis_plot(data, data2=None, data2_params=None,
                 data2_params = params
             for i, p in enumerate(data2_params):
[…]

### 2022-02-03 manuscript/manuscript.tex
update style file locations
Version 2 of the manuscript

@@
 %\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass{bmc_template/bmcart}
+\documentclass{bmcart}
 
@@
    corref={aff1},                       % id of corresponding address, if any
-   noteref={n1},                        % id's of article notes, if any
    email={y.behr@gns.cri.nz}   % email address
@@
 \begin{abstract} % abstract
-\parttitle{First part title} %if any
-Text for this section.
-
-\parttitle{Second part title} %if any
-Text for this section.
+    Volcanic lakes often capture a significant amount of volcanic heat emission
+    and thus provide a unique opportunity to monitor changes inside the volcano.
+    We present a Bayesian inversion method to automatically infer volcanic heat
+    emission at the base of a volcanic lake from lake monitoring data using a
+    non-linear Kalman Smoother. Our method accounts for the, in parts large,
+    uncertainties in observations and the underlying physics-based model to
+    generate probabilistic estimates of heat emission. We verify our results
+    using a synthetic test case and then estimate the daily heat input rate into
+    Mt. Ruapehu's Crater Lake, New Zealand, between 2016 and 2022. Time-frequency
+    analysis of the heat input rate shows dominant periods of heating cycles
+    ranged between 100 - 250 days. Some years are dominated by shorter cycles
+    and others by longer cycles which potentially points to changes in the
+    magmatic heat supply and the hydrothermal system.
 \end{abstract}
@@
 \begin{keyword}
-\kwd{sample}
-\kwd{article}
-\kwd{author}
+\kwd{Crater lakes}
+\kwd{Non-linear Kalman Smoother}
+\kwd{Volcanic heat emission}
 \end{keyword}
@@
 Monitoring changes in temperature, water mass, and ion concentration in these
-lakes can thus serve as proxies for changes in gas- and steam-release by the
-hydrothermal system and, ultimately, the melt zone beneath the volcano.
+lakes can, therefore, serve as proxies for changes in gas- and steam-release by
+the hydrothermal system and, ultimately, the melt zone beneath the volcano.
 
@@
 
-Following \cite{Hurst1991} and \cite{Stevenson1992} the mass and energy balance
-can be written in terms of three coupled ordinary differential equations. The
-rate of lake water mass ($M$) change can be described as:
+The mass and energy balance model can be written in terms of three coupled ordinary
+differential equations \cite{Hurst1991, Stevenson1992}. The rate of lake water
+mass ($M$) change is described as:
 \begin{equation}\label{ode_M}
@@
 which include the recent development on forced (i.e. wind-induced) convection
-\cite{sartoriCriticalReviewEquations2000}.
+\cite{sartoriCriticalReviewEquations2000}. It is important to note here that
+$Q_e$ is a function of $T$ and the windspeed above the lake.
 
@@
 
-In a sensitivity analysis for $\dot{T}$ (Figure \ref{sens}) we can see a linear
-relationship with respect to $\dot{Q_i}$ and $\dot{M_o}$ and a mildly
-non-linear relationship with respect to $\dot{Q_i}$ and current season
-(note that this assumes southern hemisphere seasons). The relationship with respect
-to $T$ and windspeed, however, is strongly non-linear. This is because $\dot{Q_e}$
-and $\dot{M_e}$ are non-linear functions of windspeed and $\dot{Q_i}$. This further 
-complicates the inference of $\dot{Q_i}$.
+The main challenge in solving the mass and energy balance equations for
+$\dot{Q_i}$ arises from the non-linear relationship between $Q_e$ and ${T}$
+as shown in Figure \ref{sens}.
 
@@
     \includegraphics[width=12 cm]{figures/sensitivity_analysis.png}  
-    \caption{Sensitivity analysis of $\dot{T}$ with respect to $\dot{Q_i}$ 
-             and $\dot{M_o}$ (top panel), $\dot{Q_i}$ and season 
-             (center panel), and $T$ and windspeed (bottom panel). The model
-              parameters that have been held fixed are described in each 
-              panel's title. Warmer colors indicate a temperature increase
-              ($\dot{T} > 0$) and cooler colors a temperature decrease
-              ($\dot{T} < 0$)}
+    \caption{Surface energy loss ($Q_e$) with respect to lake water temperature
+             ($T$). The model parameters that have been held fixed are shown in  
+             the title.}
       \label{sens}
@@
 the $n$ model parameters described in Section \ref{phm} at time step $k$. Note
-that $x_k$ also includes $T$, $M$, $\dot{M_o}$, and $X$. We distinguish here
-between the state of the lake and observations of this state
-($T$, $M$, $X$, $\dot{M_o}$) which are to some degree uncertain.
+that $x_k$ also includes $T$, $M$, and $X$. We distinguish here between the
+state of the lake and observations of this state ($T$, $M$, $X$) which are to
+some degree uncertain. Equation \ref{bsm} can be seen as the probabilistic
+formulation of the inverse problem inferring $\dot{Q_i}$ from available
+observations.
 
@@
 Carlo is computationally very expensive as the number of variables grows
-exponentially with the length of the timeseries. An alternative are Bayesian
-Filtering and Smoothing algorithms of which the Kalman Filter and Smoother are
-the closed form solution for linear models \cite{Kalman1960, Rauch1965,
-sarkkaBayesianFilteringSmoothing2013}.
+exponentially with the length of the timeseries and the number of model
+parameters. An alternative are Bayesian Filtering and Smoothing algorithms of
+which the Kalman Filter and Smoother are the closed form solution for linear
+models \cite{Kalman1960, Rauch1965, sarkkaBayesianFilteringSmoothing2013}.
 
@@
 
-where $f(\bm{x}_{k-1})$ is the physical model described in Section \ref{phm}
-propagated forward in time using a fourth-order Runge-Kutta scheme. If we 
-assume the errors, $\bm{r}_k$ and $\bm{q}_{k}$, to be normally distributed
-($\bm{q}_{k} \sim N(\bm{0},
-\bm{Q}_{k})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$), we can compute an
-approximate solution to \ref{bsm1} using the \textit{Unscented Kalman Filter}
-\citep{Merwe2004, sarkkaBayesianFilteringSmoothing2013}.
+where $f(\bm{x}_{k-1})$ is the non-linear physical model described in Section
+\ref{phm} propagated forward in time using a fourth-order Runge-Kutta scheme. If
+we assume the errors, $\bm{r}_k$ and $\bm{q}_{k}$, to be normally distributed
+($\bm{q}_{k} \sim N(\bm{0}, \bm{Q}_{k})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$),
+we can compute an approximate solution to \ref{bsm1} using the \textit{Unscented
+Kalman Smoother} \citep{Merwe2004, sarkkaBayesianFilteringSmoothing2013}, a
+non-linear Kalman Smoother.
 
@@
 uncertainties similar to those estimated from real observations (see Section
-\ref{realdata}).
+\ref{realdata}). As with real data, average and standard deviation of all values
+within a day are calculated. The standard deviation is taken as a proxy for the
+uncertainty in observations ($\bm{R}_k$). To simulate the disparity between sampling
+intervals of different data streams we randomly remove most of the observations
+of $X$, $\dot{M_o}$, and $\dot{M_s}$. This dataset is then inverted using the
+Unscented Kalman Smoother to recover the original heat input rate.
 
@@
 
-It is noteworthy that the mass influx and outflux rates are recovered
-quite accurately despite having no data in the former and very little data
-in the latter case.
+It is noteworthy that the mass outflux rates are recovered quite accurately
+despite providing very little synthetic observations.
 
@@
     \caption{Inversion result for the synthetic test described in Section 
-        \ref{syn_test}. Turquoise lines and markers represent hypothetical
+        \ref{syn_test}. Turquoise lines and markers represent synthetic 
         observations; red dashed lines are the inversion results and black
-        dashed lines are the true input to the synthetic test.}
+        dashed lines are the true input to the synthetic test. Error bars 
+        represent $\pm$ 3 standard deviations.}
   \label{syn_example}
@@
 transmit data several times a day via satellite to GNS Science, which monitors
-New Zealand's geohazards through its GeoNet project (www.geonet.org.nz
- last accessed 16/07/2021).
+New Zealand's geohazards through its GeoNet project \cite{GeoNetHome}.
 Water samples are taken 8-10 times a year from which subsequently, among many
@@
 
-Figure \ref{inference_result} shows $\dot{Q_i}$ for RCL for lake measurements
-between 4 March 2016 and 7 July 2021. During this time the heat input rate
+Figure \ref{inference_result}a shows $\dot{Q_i}$ for RCL for lake measurements
+between 4 March 2016 and 1 February 2022. During this time the heat input rate
 reached a maximum of 753 MW and was on average 165 MW. The average standard
@@
 
+To also investigate changes in periodicity we computed the continuous wavelet
+transform of of $\dot{Q_i}$ using a complex Morlet wavelet (Figure
+\ref{inference_result}b). We chose the wavelet transform over the more common
+windowed Fourier transform for its ability to detect non-stationary signals
+\cite{torrencePracticalGuideWavelet1998}. Figure \ref{inference_result}c shows
+the global wavelet spectrum which is an average of the wavelet analysis in
+Figure \ref{inference_result}b over the whole analysis period. The global
+wavelet spectrum can be interpreted similarly to the Power Spectral Density.
+Figure \ref{inference_result} shows that periodicity is dominated by periods
+between 100 and 250 days. Which period is dominant at any one time changes
+significantly and we will expand on this further in the following section.
+
 \begin{figure}
- 	\includegraphics[width=13 cm]{figures/uks_real_data.png}  
- 	\caption{Inference results for the heat input rate $\dot{Q_i}$ at RCL
-             between 4 March 2016 and 7 July 2021. The solid line shows the
+ 	\includegraphics[width=12 cm]{figures/time_frequency_analysis.png}  
+    \caption{ a) Inference results for the heat input rate $\dot{Q_i}$ at RCL
+             between 4 March 2016 and 1 February 2022. The solid line shows the
              marginal probability and the uncertainty is displayed as shaded
-             region.}
+             region; b) Continuous wavelet transform of the time series shown in
+             Panel a; c) Power spectral density of the heat input rate shown as a
+             solid line with uncertainty displayed as shaded region.}
 \label{inference_result}
@@
 estimating the total energy budget of a volcanic system \cite{Brown1989a}.
-Similar to a volcano's gas budget, it can either indicate migration of magma or
-changes in permeability within the vent system.
+Similar to a volcano's gas budget, changes in the heat input rate can either
+indicate changes in the magmatic heat source or in the hydrothermal system
+above.
 
-Figure \ref{psd} shows the power spectral density (PSD) for the marginal heat
-input rate shown in Figure \ref{inference_result}. It was computed using the
-multi taper method \cite{prietoComputersGeosciencesFortran2009} which provides
-not only an estimate of the PSD but also its uncertainty. The PSD shows that the
-dominant periods in the heat input rate lie between $\sim$100 and $\sim$250
-days. Whether these periods are due to cyclic processes in the magmatic system,
-the hydrothermal system, or a combination of both requires further modelling of
-these processes. Many studies have demonstrated the value of non-linear Kalman
-Filters for inference and data-assimilation with non-linear, high-dimension
-numerical models \cite{White2018a}. While beyond the scope of this study, the
-next logical step, therefore, is to include numerical models of magma degassing
-and hydrothermal fluid flow to model the heat input rate, volume, and enthalpy
-of the steam entering the lake.
-
-\begin{figure}
-    \includegraphics[width=12 cm]{figures/psd_analysis.png}  
- 	\caption{Power spectral density of the heat input rate 
-             (Figure \ref{inference_result}) shown as a solid line
-             with uncertainty displayed as shaded region.}
-    \label{psd}
-\end{figure}
-
-Irrespective of the process that is responsible for the periodicity, it appears
-that it has not changed significantly over the past five years. This may
-suggest that the volcano has been in a dynamic equilibrium during this time.
-This is further supported by the cumulative heat input (Figure \ref{cumheat}).
-It shows a slight increase in heat input in mid-2017 but a fairly constant rate
-since then. Of course this does not in any way forecast the volcano's future
-behaviour but it may serve as a useful baseline to detect changes in the
-volcano's state more easily.
+We would expect a significant change in the heat source to cause a change in
+slope of the cumulative heat input rate (Figure \ref{cumulative}). The fact that
+we do not observe such changes may indicate a more or less constant supply of
+heat during the studied period. Experiments on analog models of hydrothermal
+systems suggest that changes in the heat source and the hydrothermal system
+should also change the length of heating cycles
+\cite{vandemeulebrouckAnalogueModelingInstabilities2005,fitzgerald1997TwentySecondWorkshop}.
+Figure \ref{inference_result}b shows the heating cycles before mid-2017 and
+after mid-2019 were dominated by longer periods. From mid-2017 to mid-2019
+cycles were mostly of shorter periods which may indicate a change in the
+volcanic system during this time.
 
@@
  	\includegraphics[width=12 cm]{figures/cumulative_heat.png}  
-    \caption{Cumulative heat input into RCL (solid line) and its 
-    uncertainty (shaded region).}
-    \label{cumheat}
+    \caption{Cumulative heat input into RCL (solid line) and its uncertainty 
+             (shaded region) calculated for the time-series in
+             Figure \ref{inference_result} a.} 
+\label{cumulative}
 \end{figure}
+
+To further explore hypotheses like this, the next logical step is to include
+numerical models of magma degassing and hydrothermal heat and fluid flow. While
+beyond the scope of this study, our inversion method is well suited for such
+more complex models as its runtime scales linearly with the number of
+parameters. Several studies have demonstrated the value of non-linear Kalman
+Filters for inverse problems with non-linear, high-dimensional numerical models
+\cite[e.g.][]{White2018a,huangIteratedKalmanMethodology2022}.
 
@@
 boundary conditions such as negative values for the inflow rate of meteoric
-water \cite{hurstCraterLakeEnergy2015, Stevenson1992}. Their explanation was the
-existence of a heat pipe beneath the lake where gas and fluid phases coexist. As
-the rising gas enters the lake relatively cool lake water enters the vent
-leading to an overall lower addition of mass to the lake and resulting in an
-enthalpy value that is double than expected for saturated steam at hydrostatic
-pressures likely to be found at the bottom of the lake. By including
-uncertainties in our observations we no longer require these high enthalpy
-values to satisfy the physical constraints. Numerical modelling of hydrothermal
-processes may also shed further light on the heating processes and whether or
-not a heat pipe is a plausible mechanism.
+water ($\dot{M_s}$) \cite{hurstCraterLakeEnergy2015, Stevenson1992}. Their
+explanation was that as rising hot fluids enter the lake, relatively cool lake
+water flows back into the vent. This would lead to an overall lower addition of
+mass to the lake and resulting in an enthalpy value that is double than expected
+for saturated steam at hydrostatic pressures likely to be found at the bottom of
+the lake. By including uncertainties in our observations, this explanation is
+still a possibility but no longer a requirement to satisfy the physical
+constraints. Numerical modelling of hydrothermal processes may also shed further
+light on the heating processes and whether or not the lake and the vent form a
+single or two separate convective systems.
 

### 2022-02-03 manuscript/figures/sensitivity_analysis.py

@@ -20,29 +20,16 @@ results = sensitivity_analysis_grid()
 
 # data selection
 results['Date'] = pd.to_datetime([date(1,1,1) + timedelta(days=dt-1) for dt in results['Date']])
-df1 = select_data(results, Ws=2.0, T0=25.0, H=6.0, Date='2020-01-01')
-qi = df1.Qi.values.reshape((10, 10))
-mo = df1.Mo.values.reshape((10, 10))
-dT = df1.dT.values.reshape((10, 10))
-
-df2 = select_data(results, Mo=20., T0=25.0, H=6.0, Ws=2.0)
-dt = np.datetime_as_string(df2.Date.values.reshape((10, 10)))
-qi2 = df2.Qi.values.reshape((10, 10))
-dT2 = df2.dT.values.reshape((10, 10))
-
-df3 = select_data(results, Mo=20., Date='2020-01-01', H=6.0, Qi=200)
-ws = df3.Ws.values.reshape((10, 10))
-T = df3.T0.values.reshape((10, 10))
-qe = df3.Qe.values.reshape((10, 10))
-dT3 = df3.dT.values.reshape((10, 10))
-dM3 = df3.dM.values.reshape((10, 10))
+df3 = select_data(results, Mo=20., Ws=5, Date='2020-01-01', H=6.0, Qi=200)
+T = df3.T0.values
+qe = df3.Qe.values/0.0864
+dT3 = df3.dT.values
+dM3 = df3.dM.values
+me = df3.Me.values
 
 # Plotting
-fig = make_subplots(rows=3, cols=1,
-                    subplot_titles=["T=25 \u00b0C, H=6 MJ/kg, Ws=2 m/s, Date=2020-01-01",
-                                    "T=25 \u00b0C, H=6 MJ/kg, Ws=2 m/s, Mo=20 kt/day",
-                                    "Qi=200 MW, H=6 MJ/kg, Mo=20 kt/day, Date=2020-01-01"],
-                    vertical_spacing=0.09)
+fig = make_subplots(rows=1, cols=1,
+                    subplot_titles=["Qi=200 MW, Ws=5 m/s, H=6 MJ/kg, Mo=20 kt/day, Date=2020-01-01"])
 
 def add_contour_plot(x, y, z):
     return go.Contour(x=x, y=y, z=z,
@@ -51,16 +38,12 @@ def add_contour_plot(x, y, z):
                                                    color='black')),
[…]

### 2022-02-03 manuscript/figures/synthetic_inversion_uks.py

@@ -2,7 +2,7 @@ from collections import OrderedDict
 
 import numpy as np
 
-from pumahu.syn_model import (SynModel,
+from pumahu.syn_model import (SynModel, remove_inputs,
                               setup_realistic,
                               resample,
                               make_sparse)
@@ -10,22 +10,25 @@ from pumahu.visualise import trellis_plot
 from pumahu.uks import UnscentedKalmanSmoother
 
 
-xds = SynModel().run(setup_realistic(sinterval=120), addnoise=True)
+xds = SynModel().run(setup_realistic(sinterval=120), addnoise=True,
+                     ignore_cache=True)
+xds_ideal = SynModel().run(setup_realistic(sinterval=120), addnoise=False,
+                           ignore_cache=True)
+
 
 na = resample(xds)
-na = make_sparse(na, ['m_out', 'X'])
+na = make_sparse(na, ['m_out', 'm_in', 'X'])
+na = remove_inputs(na, ['m_in', 'q_in'])
 Q = OrderedDict(T=5e-1, M=1e-1, X=1e-3, q_in=1e1,
                 m_in=1e1, m_out=1e1, h=1e-3, W=1e-3,
                 dqi=1e2, dMi=0, dMo=0, dH=0, dW=0)
 Q = np.eye(len(Q))*list(Q.values())
 
-uks = UnscentedKalmanSmoother(data=na, Q=Q,
-                              initvals={'qi':200., 'm_in': 0,
-                                        'm_out':0., 'X':3.})
+uks = UnscentedKalmanSmoother(data=na, Q=Q)
 xds_uks = uks(test=True, smooth=True, alpha=0.35)
-fig = trellis_plot(xds_uks, data2=xds,
+fig = trellis_plot(xds_uks, data2=xds_ideal,
                    data2_params=['q_in', 'm_in', 'm_out'],
-                   filled_error=True)
+                   filled_error=True, data2_showerror=False)
 fig.update_layout(font_size=20, height=1200, showlegend=False)
 filename="synthetic_inversion_uks.png"
 fig.write_image(filename, scale=2)

### 2022-02-03 manuscript/figures/time_frequency_analysis.py

@@ -1,0 +1,197 @@
+import pandas as pd
+import numpy as np
+import os
+import xarray as xr
+
+import plotly.graph_objects as go
+import pywt
+
+from pumahu.visualise import TrellisPlot
+from pumahu.uks import main as main_uks
+
+if False:
+    df_modis = pd.read_csv(get_data('data/41561_2021_705_MOESM2_ESM.csv'), usecols=[5, 6, 7, 8],
+                            index_col=0, dayfirst=True, parse_dates=True)
+    df_modis = df_modis.loc[str(dates[0]):str(dates[-1])]
+    
+def wavelet_analysis(dates, data, nscales=128):
+    sst = data - data.mean()
+    dt = (dates[1]-dates[0])/pd.Timedelta('1D')
+
+    # Taken from http://nicolasfauchereau.github.io/climatecode/posts/wavelet-analysis-in-python/
+    wavelet = 'cmor1.5-1.0'
+    #wavelet = 'cmor2.0-6.0'
+    #wavelet = 'mexh'
+    scales = np.arange(1, nscales)
+
+    [cfs, frequencies] = pywt.cwt(sst, scales, wavelet, dt)
+    power = (abs(cfs)) ** 2
+
+    period = 1. / frequencies
+    return period, power
+
+   
+fn = 'data/uks.nc'
+if not os.path.isfile(fn):
+    main_uks(['--rdir', './data', '-s', '2016-03-04',
+              '-e', '2022-02-01', '-f'])
+
+xdf = xr.open_dataset(fn)
+data = xdf.exp.loc[:, 'q_in', 'val'].values[:]
+dates = xdf['dates'].values
+
+tp = TrellisPlot()
+ymean, ymin, ymax = tp.get_traces(xdf.exp, 'q_in')
+nscales = 256 
+period, power = wavelet_analysis(dates, data, nscales)
+mean_power = power.mean(axis=1)
+std_power = power.std(axis=1)
+min_power = mean_power - std_power
+max_power = mean_power + std_power
+min_power = np.where(min_power < 0., 0., min_power)
+
+fig = go.Figure()
+fig.add_trace(go.Contour(
+        x = pd.to_datetime(dates),
+        y = period,
+        z = power,
+        contours=dict(
+            start=0,
+            end=700e3,
+            size=100e3,
+        ),
+        contours_coloring='heatmap',
+        colorscale = 'Blues',
+        reversescale = True,
[…]

### 2022-04-19 manuscript/figures/cumulative_heat.py
Change to cumulative difference to the mean

@@ -1,9 +1,16 @@
+import os
+
 import plotly.graph_objects as go
 import numpy as np
 import xarray as xr
 import pandas as pd
 
+from pumahu.uks import main as main_uks
+
 fn = './data/uks.nc'
+if not os.path.isfile(fn):
+    main_uks(['--rdir', './data', '-f', '-s', '2016-03-04', '-e', '2022-02-01'])
+
 xdf = xr.open_dataset(fn)
 data = xdf.exp.loc[:, 'q_in', 'val'].values
 sigma = xdf.exp.loc[:, 'q_in', 'std'].values
@@ -13,7 +20,7 @@ sigma *= 0.864
 # Convert to kWh by deviding with 3.6e6 J
 data /= 3.6
 sigma /= 3.6
-cum_mean = np.cumsum(data)
+cum_mean = np.cumsum(data - data.mean())
 cum_std = np.sqrt(np.cumsum(sigma))
 dates = pd.to_datetime(xdf.dates.values)
 fig = go.Figure()

### 2022-04-19 manuscript/figures/overview.py
Minor formatting updates

@@ -18,13 +18,17 @@ ax1.text(1850,750, 'O', color='red', weight='bold', size='18')
 # Map
 tiler = StamenTerrain()
 mercator = tiler.crs
-ax2 = fig.add_axes([0.65,0.1,0.3,0.35], projection=mercator)
+ax2 = fig.add_axes([0.64,0.1,0.3,0.35], projection=mercator)
 ax2.set_extent([165, 180, -49, -32], crs=ccrs.PlateCarree())
 ax2.add_image(tiler, 6)
 ax2.coastlines(resolution='50m')
 
+label_style = {'color': 'black', 'weight': 'bold', 'size':10,
+               'bbox': {'facecolor':'white', 'edgecolor': 'white', 'visible': True}}
 gl = ax2.gridlines(crs=ccrs.PlateCarree(), draw_labels=True,
-                                     linewidth=1, color='gray', alpha=0.5, linestyle='--')
+                                     linewidth=1, color='gray', alpha=0.5,
+                                     linestyle='--', xlabel_style=label_style,
+                                     ylabel_style=label_style)
 gl.xlabels_top = False
 gl.ylabels_left = False
 gl.xlines = True
@@ -32,8 +36,6 @@ gl.ylines = True
 gl.xlocator = mticker.FixedLocator([160, 166, 172, 179, 180])
 gl.xformatter = LONGITUDE_FORMATTER
 gl.yformatter = LATITUDE_FORMATTER
-gl.xlabel_style = {'color': 'red', 'weight': 'bold', 'size':10}
-gl.ylabel_style = {'color': 'red', 'weight': 'bold', 'size':10}
 ax2.plot(175.564490, -39.281149, marker='^', color='red',
                   markersize=6, transform=ccrs.PlateCarree())
 fig.savefig('overview.png', dpi=300, bbox_inches='tight')

### 2022-04-19 manuscript/figures/synthetic_inversion_uks.py

@@ -30,5 +30,8 @@ fig = trellis_plot(xds_uks, data2=xds_ideal,
                    data2_params=['q_in', 'm_in', 'm_out'],
                    filled_error=True, data2_showerror=False)
 fig.update_layout(font_size=20, height=1200, showlegend=False)
+fig.update_yaxes(range=(0,40), row=5, col=1)
+fig.update_yaxes(range=(0,40), row=6, col=1)
+fig.update_yaxes(range=(2.5, 3.), row=7, col=1)
 filename="synthetic_inversion_uks.png"
 fig.write_image(filename, scale=2)

### 2022-04-19 manuscript/figures/time_frequency_analysis.py
Change to pywt to get rid of noise in contour lines

@@ -1,3 +1,4 @@
+from tkinter import W
 import pandas as pd
 import numpy as np
 import os
@@ -8,29 +9,53 @@ import pywt
 
 from pumahu.visualise import TrellisPlot
 from pumahu.uks import main as main_uks
+import pycwt
+from pycwt.helpers import find
+
 
 if False:
     df_modis = pd.read_csv(get_data('data/41561_2021_705_MOESM2_ESM.csv'), usecols=[5, 6, 7, 8],
                             index_col=0, dayfirst=True, parse_dates=True)
     df_modis = df_modis.loc[str(dates[0]):str(dates[-1])]
     
-def wavelet_analysis(dates, data, nscales=128):
-    sst = data - data.mean()
-    dt = (dates[1]-dates[0])/pd.Timedelta('1D')
+def wavelet_analysis(dates, data, wavelet_pkg='pycwt'):
+    if wavelet_pkg == 'pywt':
+        nscales = 256
+        sst = data - data.mean()
+        dt = (dates[1]-dates[0])/pd.Timedelta('1D')
 
-    # Taken from http://nicolasfauchereau.github.io/climatecode/posts/wavelet-analysis-in-python/
-    wavelet = 'cmor1.5-1.0'
-    #wavelet = 'cmor2.0-6.0'
-    #wavelet = 'mexh'
-    scales = np.arange(1, nscales)
+        # Taken from http://nicolasfauchereau.github.io/climatecode/posts/wavelet-analysis-in-python/
+        wavelet = 'cmor1.5-1.0'
+        #wavelet = 'cmor2.0-6.0'
+        #wavelet = 'mexh'
+        scales = np.arange(1, nscales)
 
-    [cfs, frequencies] = pywt.cwt(sst, scales, wavelet, dt)
-    power = (abs(cfs)) ** 2
+        [cfs, frequencies] = pywt.cwt(sst, scales, wavelet, dt)
+        power = (abs(cfs)) ** 2
 
-    period = 1. / frequencies
+        period = 1. / frequencies
+    elif wavelet_pkg == 'pycwt':
+        dt = (np.diff(pd.to_datetime(dates))/pd.Timedelta('1D'))[0]
+        N = data.size
+        t = np.arange(0, N) * dt
+        p = np.polyfit(t, data, 1)
+        data_notrend = data - np.polyval(p, t)
+        std = data_notrend.std()  # Standard deviation
+        var = std ** 2  # Variance
[…]

### 2022-04-19 manuscript/manuscript.tex
Submitted version

@@
 %\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass{bmcart}
+\documentclass[doublespacing, linenumbers]{bmcart}
 
@@
 %\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-\usepackage{graphicx}
+%\usepackage{graphicx}
 
@@
 
-% \def\includegraphic{}
-% \def\includegraphics{}
+\def\includegraphic{}
+\def\includegraphics{}
 
@@
     emission at the base of a volcanic lake from lake monitoring data using a
-    non-linear Kalman Smoother. Our method accounts for the, in parts large,
+    non-linear Kalman Smoother. Our method accounts for the, sometimes large,
     uncertainties in observations and the underlying physics-based model to
@@
     using a synthetic test case and then estimate the daily heat input rate into
-    Mt. Ruapehu's Crater Lake, New Zealand, between 2016 and 2022. Time-frequency
-    analysis of the heat input rate shows dominant periods of heating cycles
-    ranged between 100 - 250 days. Some years are dominated by shorter cycles
-    and others by longer cycles which potentially points to changes in the
-    magmatic heat supply and the hydrothermal system.
+    Mt. Ruapehu's Crater Lake, New Zealand, between 2016 and 2022.
+    Time-frequency analysis of the heat input rate shows dominant periods of
+    heating cycles ranged between 100 - 250 days. The period between 2017 and
+    2020 was dominated by shorter cycles and greater-than-average heat input
+    rate which points to changes in the magmatic heat supply and the
+    hydrothermal system during this time.
 \end{abstract}
@@
 emission rate together with a high carbon-sulfur ratio of the emitted gas may
-point to the formation of a hydrothermal seal.
+point to the formation of a hydrothermal seal
+\cite{christensonCyclicProcessesFactors2010}.
 
@@
 
-Ruapehu Crater Lake (Te Wai A-Moe in Te Reo Maori and referred to as RCL from
+Ruapehu Crater Lake (Te Wai \={a}-moe in Te Reo Maori and referred to as RCL from
 hereon) on Mt. Ruapehu, an active andesitic stratovolcano in the center of New
@@
 Ruapehu erupted at least 602 times since 1830 with many phreatic and
-phreato-magmatic eruptions and two major magmatic episodes \citep{Scott2013}.
-Although the latest magmatic event between June 1995 to November 1997 and a
-subsequent dam collapse changed the shape and volume of the lake, it has been
-largely unchanged since then and the lake's bathymetry is well known from times
-when it was empty [Tony Hurst, personal communication].
+phreato-magmatic eruptions and two major magmatic episodes
+\citep{Scott2013,HistoricEruptiveActivity2022}. Although the latest magmatic
+event between June 1995 to November 1997 and a subsequent dam collapse changed
+the shape and volume of the lake, it has been largely unchanged since then and
+the lake's bathymetry is well known from times when it was empty.
 
@@
 monitored continuously and water samples for ion concentration measurements have
-been taken roughly every two to three months. This enables us to continuously
+been taken on an approximately monthly basis. This enables us to continuously
 estimate heat input into the lake and, in principle, automatically detect
@@
 
-Inference of the heat input is typically done through a mass and energy balance
-calculation: the sum of mass and energy flow into and out of the lake has to
-account for the observed changes in lake temperature, water level and ion
-dilution \cite{Hurst1981, Hurst1991, Stevenson1992, Fournier2009, Scott1994}. At
-RCL, and probably at most other volcanic lakes, this is an underdetermined
-problem, that is the number of unknowns exceeds the number of constraints.
-Hence, it is necessary to constrain some of the parameters in the mass and
-energy balance model using prior knowledge. Further, some observations and input
-parameters come with large uncertainties and are sampled irregularly.
+We typically infer the heat input from a mass and energy balance calculation:
+the total mass and energy flow into and out of the lake has to account for the
+observed changes in lake temperature, water level and ion dilution
+\cite{Hurst1981, Hurst1991, Stevenson1992, Fournier2009, Scott1994}. At RCL, and
+probably at most other volcanic lakes, this is an underdetermined problem, that
+is the number of unknowns exceeds the number of constraints. Hence, it is
+necessary to constrain some of the parameters in the mass and energy balance
+model using prior knowledge. Further, some observations and input parameters
+come with large uncertainties and are sampled irregularly.
 
@@
 
-\begin{figure}
-\includegraphics[width=12 cm]{figures/overview.png}  
-\caption{Image of Mt. Ruapehu's crater lake from 2008 (\textcopyright Karen
-	Britten, GNS Science). The red O marks the location of the lake outlet. The
-	location of Mt. Ruapehu is shown as a red triangle on the insert map.}
-  \label{overview}
-\end{figure}
-
 \section{Probabilistic inference method}\label{Pim}
@@
 
-The mass and energy balance model can be written in terms of three coupled ordinary
-differential equations \cite{Hurst1991, Stevenson1992}. The rate of lake water
-mass ($M$) change is described as:
+The mass and energy balance model can be written in terms of three coupled
+ordinary differential equations \cite{Hurst1991, Stevenson1992}. The rate of
+lake water mass change ($\dot{M}$) is the difference between the rate of steam
+input through the vent system beneath the lake ($\dot{M_i}$), the water inflow
+rate through melt and precipitation ($\dot{M_s}$), the outflow rate through
+seepage and overflow ($\dot{M_o}$), and the rate of mass loss through
+evaporation ($\dot{M_e}$): 
 \begin{equation}\label{ode_M}
-	\frac{dM}{dt} = \dot{M_s} - \dot{M_o} - \dot{M_e} + \dot{M_i}
+	\frac{dM}{dt} = \dot{M_i} + \dot{M_s} - \dot{M_o} - \dot{M_e} 
 \end{equation}
 
-Where $\dot{M_s}$ is the rate of water inflow through melt and precipitation,
-$\dot{M_o}$
-is the outflow rate through seepage and overflow, $\dot{M_e}$ is the rate of mass
-loss through evaporation and $\dot{M_i}$ is the rate of steam input through the vent
-system beneath the lake.
-
-The energy content of the lake can be described with the notion of
-enthalpy, $H$. At constant pressure, enthalpy is defined as:
+The energy content of the lake can be described with the notion of enthalpy,
+$H$, which is defined relative to a reference enthalpy. In the following we take
+the enthalpy of the lake at temperature $T=0~^{\circ}C=T_0$ as reference
+$H_0=H(T_0)$. At constant pressure, the enthalpy of the lake can then be written
+as a function of $T$, $T_0$, $M$ and the specific heat of water, $c_w$:
 \begin{equation}
-	H(T) = H(T_0) + \int_{T_0}^{T}c_pM\tau d\tau
+	H(T) = H_0 + \int_{T_0}^{T}c_wM\tau d\tau
 \end{equation}
 
-where $T$ is temperature and $c_p$ is the specific heat. For all following
-equations we take $H(T_0=0~^{\circ}C)$ as our reference enthalpy and
-$c_p=c_w=constant$, with $c_w$ the specific heat of water, such that the
-enthalpy of lake water at temperature $T$ can be written as $H(T) = c_wMT$.
+For all following equations we take $c_w=constant$, such that the enthalpy of
+lake water at temperature $T$ can be written as $H(T) = H_0 + c_wMT$.
+
+The rate of enthalpy change of the lake water can then be defined as a function
+of the rate of energy input from volcanic sources ($\dot{Q_i}$), the rate of
+energy gain due to solar irradiation from short-wavelength radiation
+($\dot{Q_r}$), the rate of energy loss through outflow/seepage
+($c_wT\dot{M_o}$), the rate of surface energy losses due to evaporation (forced
+and free convection), sensible heat, and long-wavelength radiation
+($\dot{Q_e}$), and the rate of energy loss from heating incoming surface water
+($c_wT\dot{M_s}$):
+
+\begin{equation}\label{ode_E}
+    \frac{dH}{dt}=\frac{d(c_wTM)}{dt}=c_wT\dot{M} + c_wM\dot{T} =
+	\dot{Q_i} + \dot{Q_r} - c_wT\dot{M_o} - \dot{Q_e} - c_wT\dot{M_s} 
+\end{equation}
+
 Because RCL is at an altitude of $\sim$~2500~m and big parts of its catchment
 are covered by snow and ice all year round we assume the temperature of incoming
-water to be at 0 $^{\circ}C$.
+surface water to be at 0 $^{\circ}C$. Several empirical and theoretical equations exist
+to describe the surface losses and we refer the reader to previous publications
+for more detailed discussions \cite{Stevenson1992, hurstCraterLakeEnergy2015}.
+There is no consensus on the best quantitative model of evaporation from warm
+lakes and, in fact, the most suitable equation may depend on the location of and
+available observations from the lake. Appendix \ref{A} describes the equations
+used in this study in detail which include the recent development on forced
+(i.e. wind-induced) convection \cite{sartoriCriticalReviewEquations2000}. It is
+important to note here that $Q_e$ is a function of $T$ and the windspeed above
+the lake.
 
-The rate of enthalpy change of the lake water can thus be written as:
-\begin{equation}\label{ode_E}
-    \frac{dH}{dt}=\frac{d(c_wTM)}{dt}=c_wT\dot{M} + c_wM\dot{T} =
-	-\dot{Q_e} - c_wT\dot{M_o} + \dot{Q_{eff}} + \dot{Q_r}
-\end{equation}
-
-where $T$ is the lake water temperature, $\dot{Q_r}$ is the rate of energy gain
-due to solar irradiation (short-wavelength radiation), and $c_wT\dot{M_o}$ is
-the rate of energy loss through outflow/seepage. $\dot{Q_e}$ is the rate of
-surface energy losses due to evaporation (forced and free convection), sensible
-heat, and long-wavelength radiation (net loss). Several empirical and
-theoretical equations exist to describe the surface losses and we refer the
-reader to previous publications for more detailed discussions
-\cite{Stevenson1992, hurstCraterLakeEnergy2015}. There is no consensus on the
-best quantitative model of evaporation from warm lakes and, in fact, the most
-suited equation may depend on the location of and available observations from
-the lake. Appendix \ref{A} describes the equations used in this study in detail
-which include the recent development on forced (i.e. wind-induced) convection
-\cite{sartoriCriticalReviewEquations2000}. It is important to note here that
-$Q_e$ is a function of $T$ and the windspeed above the lake.
-
-$\dot{Q_{eff}}$ is the effective rate of energy gain from volcanic sources after
-heating incoming water:
-\begin{equation}
-	\dot{Q_{eff}}=\dot{Q_i} - c_wT\dot{M_s}
-\end{equation}
-with $\dot{Q_i}$ the rate of energy input from volcanic sources.
-
-Following from Equation \ref{ode_E} the temperature change $\frac{dT}{dt}$ can then be written as:
+Following from Equation \ref{ode_E} the temperature change $\frac{dT}{dt}$ can
+then be written as:
 \begin{equation}\label{ode_T}
@@
 
-Finally, the change in ion concentration of, for example, $Mg^{2+}$ can be expressed as:
+Finally, the change in ion concentration expressed in total ion amount ($X$) of,
+for example, $Mg^{2+}$ can be defined as:
 \begin{equation}\label{ode_X}
@@
 
-Where $X$ is the total ion amount in the lake. This equation is only true under
-the assumption that $X$ only changes due to dilution and there is no re-supply
-between eruptive episodes. Equation \ref{ode_X} gives an estimate of the average
-outflow over periods on the order of months but does not capture shorter term
-variations in the outflow.
+This equation is only true under the assumption that $X$ only changes due to
+dilution and there is no re-supply between eruptive episodes. Equation
+\ref{ode_X} gives an estimate of the average outflow over periods on the order
+of months but does not capture shorter term variations in the outflow.
 
@@
 as shown in Figure \ref{sens}.
-
-\begin{figure}
-    \includegraphics[width=12 cm]{figures/sensitivity_analysis.png}  
-    \caption{Surface energy loss ($Q_e$) with respect to lake water temperature
-             ($T$). The model parameters that have been held fixed are shown in  
-             the title.}
-      \label{sens}
-\end{figure}
 
@@
 
-where $\bm{y}_{1:t}$ is the time series of $m$ observations and $x_k$ comprises
+where $\bm{y}_{1:t}$ is the time series of $m$ observations and $\bm{x}_k$ comprises
 the $n$ model parameters described in Section \ref{phm} at time step $k$. Note
-that $x_k$ also includes $T$, $M$, and $X$. We distinguish here between the
+that $\bm{x}_k$ also includes $T$, $M$, and $X$. We distinguish here between the
 state of the lake and observations of this state ($T$, $M$, $X$) which are to
@@
 exponentially with the length of the timeseries and the number of model
-parameters. An alternative are Bayesian Filtering and Smoothing algorithms of
+parameters. Alternatives are Bayesian Filtering and Smoothing algorithms of
 which the Kalman Filter and Smoother are the closed form solution for linear
@@
 Equations \ref{bsm1} to \ref{zetk} describe a recursive way to compute Equation
-\ref{bsm} in which we only need to know $\bm{x}_1$,
-$p(\bm{x}_k|\bm{x}_{k-1})$, and $p(\bm{y}_k|\bm{x}_k)$. Equation \ref{bsm1} and
-\ref{bfilt} are also known as the Bayesian Smoothing and Bayesian Filtering
-equation, respectively. If $p(\bm{x}_k|\bm{x}_{k-1})$ and
-$p(\bm{y}_k|\bm{x}_k)$ are linear transformations with normally distributed
-errors, the \textit{Kalman filter} \cite{Kalman1960} and Kalman smoother
-\cite{Rauch1965} represent closed form solutions. In our case, these
+\ref{bsm} in which we only need to know $\bm{x}_1$, $p(\bm{x}_k|\bm{x}_{k-1})$,
+and $p(\bm{y}_k|\bm{x}_k)$. Equation \ref{bsm1} and \ref{bfilt} are also known
+as the Bayesian Smoothing and Bayesian Filtering equation, respectively. If
+$p(\bm{x}_k|\bm{x}_{k-1})$ and $p(\bm{y}_k|\bm{x}_k)$ are linear transformations
+with normally distributed errors, the Kalman filter \cite{Kalman1960} and Kalman
+smoother \cite{Rauch1965} represent closed form solutions. In our case, these
 transformations are:
@@
 ($\bm{q}_{k} \sim N(\bm{0}, \bm{Q}_{k})$, $\bm{r}_k \sim N(\bm{0}, \bm{R}_k)$),
-we can compute an approximate solution to \ref{bsm1} using the \textit{Unscented
-Kalman Smoother} \citep{Merwe2004, sarkkaBayesianFilteringSmoothing2013}, a
-non-linear Kalman Smoother.
+we can compute an approximate solution to Equation \ref{bsm1} using the
+Unscented Kalman Smoother \citep{Merwe2004,
+sarkkaBayesianFilteringSmoothing2013}, a non-linear Kalman Smoother.
 
@@
 not know $\bm{x}_1$ and $\bm{Q}_{k}$. Setting $\bm{x}_1$ to an arbitrary value
-within physically reasonable bounds but large uncertainty works well in
-practice. While $\bm{Q}_{k}$ can be formally optimised we treat it here as a
-design parameter that we modify such that the observations are matched within
-their error bounds and the inferred heat input rate varies smoothly. The latter
-is motivated by the fact that the physical model described in \ref{phm} is not
-able to model short-term transient changes in lake volume and temperature. These
-can be caused by rain or increased meltwater influx and the underlying
-assumption of a perfectly mixed lake typically does not apply in these
-situations.
+within physically reasonable bounds but large uncertainty works well in practice
+and has very little influence on the results. While $\bm{Q}_{k}$ can be formally
+optimised we treat it here as a design parameter that we modify such that the
+observations are matched within their error bounds and the inferred heat input
+rate varies smoothly. The latter is motivated by the fact that the physical
+model described in \ref{phm} is not able to model short-term transient changes
+in lake volume, temperature, and ion concentration. These can be caused by rain
+or increased meltwater influx, or the influx of ion-enriched fluids. The
+underlying assumption of a perfectly mixed lake typically does not apply in
+these situations.
+
+The ability to account for these epistemic uncertainties explicitly is a
+particular strength of the Bayesian Filtering and Smoothing equations. It allows
+us to gain valuable insights even from very simple models.
 
@@
 
-Figure \ref{syn_example} shows the result of the probabilistic inference of
-heat input compared to the true input of the synthetic model. The observations
-are well fit within their error bounds but we see some descrepancies between
-the original and the recovered heat input rates, especially for very steep
-changes in the input rate. This is the result of preferring a smooth solution
-as already mentioned in Section \ref{inf}.
+Figure \ref{syn_example} shows the result of the probabilistic inference of heat
+input compared to the true input of the synthetic model. The observations fit
+well within their error bounds but we see some descrepancies between the
+original and the recovered heat input rates, especially for very steep changes
+in the input rate. This is the result of preferring a smooth solution as already
+mentioned in Section \ref{inf}.
 
 It is noteworthy that the mass outflux rates are recovered quite accurately
-despite providing very little synthetic observations.
-
-\begin{figure}
-	\includegraphics[width=12 cm]{figures/synthetic_inversion_uks.png}  
-    \caption{Inversion result for the synthetic test described in Section 
-        \ref{syn_test}. Turquoise lines and markers represent synthetic 
-        observations; red dashed lines are the inversion results and black
-        dashed lines are the true input to the synthetic test. Error bars 
-        represent $\pm$ 3 standard deviations.}
-  \label{syn_example}
-\end{figure}
+despite providing very few synthetic observations.
 
 \subsection{Real data}\label{realdata}
+
 Temperature at RCL is measured by an integrated-circuit temperature device
-(Texas Instruments LM35) located roughly 1.9 m below the lake's current
-overflow level. The water level sensor is a hydrostatic pressure sensor (First
-Sensor KTE/KTU/KTW6000\ldots CS Series) and it is co-located with the
-temperature sensor. Both sensors have a sampling interval of 15 minutes and
-transmit data several times a day via satellite to GNS Science, which monitors
-New Zealand's geohazards through its GeoNet project \cite{GeoNetHome}.
-Water samples are taken 8-10 times a year from which subsequently, among many
-other analytes, concentrations of $Mg^{++}$ and $Cl^-$ are determined at GNS
-Science's Geothermal Analytical Laboratory. We will focus here on $Mg^{++}$
-concentration as it shows less deviation from the assumption of constant
-dilution than $Cl^-$.
+(Texas Instruments LM35) located roughly 1.9 m below the lake's current overflow
+level. The water level sensor is a hydrostatic pressure sensor (First Sensor
+KTE/KTU/KTW6000\ldots CS Series) and it is co-located with the temperature
+sensor. Both sensors have a sampling interval of 15 minutes and transmit data
+several times a day via satellite to GNS Science, which monitors New Zealand's
+geohazards through its GeoNet project \cite{GeoNetHome}. Due to the type of
+satellite connection only about 30\% of the temperature and level measurements
+arrive at GeoNet's datacenter; the rest of the data is lost. Water samples are
+taken manually 8-12 times a year from which subsequently, among many other
+analyses, concentrations of $Mg^{++}$ and $Cl^-$ are determined at GNS Science's
+Geothermal Analytical Laboratory. We will focus here on $Mg^{++}$ concentration
+as it shows less deviation from the assumption of constant dilution than $Cl^-$.
+Because of the sparse sampling, new information on $Mg^{++}$ concentration is
+available approximately once per month.
 
@@
 Otherwise, their uncertainty is taken to be the median of directly estimated
-uncertainties.
+uncertainties from when multiple samples were taken.
 
-$\dot{M_o}$ is measured once or twice per year and we further know from field
-observations that the lake is not overflowing beneath a certain water level. We
-fit a sigmoid function to these observations assuming a maximum outflow rate of
-250 l/s. We assume the uncertainty in $\dot{M_o}$ as the 95 percent confidence
-interval of the curve-fitting.
+Lake outflow ($\dot{M_o}$) is measured once or twice per year and we further
+know from field observations that the lake is not overflowing beneath a certain
+water level. We fit a sigmoid function to these observations assuming a maximum
+outflow rate of 250 l/s. We assume the uncertainty in $\dot{M_o}$ as the 95
+percent confidence interval of the curve-fitting.
 
@@
 wavelet spectrum can be interpreted similarly to the Power Spectral Density.
-Figure \ref{inference_result} shows that periodicity is dominated by periods
+Figure \ref{inference_result}c shows that periodicity is dominated by periods
 between 100 and 250 days. Which period is dominant at any one time changes
@@
 
-\begin{figure}
- 	\includegraphics[width=12 cm]{figures/time_frequency_analysis.png}  
-    \caption{ a) Inference results for the heat input rate $\dot{Q_i}$ at RCL
-             between 4 March 2016 and 1 February 2022. The solid line shows the
-             marginal probability and the uncertainty is displayed as shaded
-             region; b) Continuous wavelet transform of the time series shown in
-             Panel a; c) Power spectral density of the heat input rate shown as a
-             solid line with uncertainty displayed as shaded region.}
-\label{inference_result}
-\end{figure}
+\section{Discussion and conclusion}
+Inferring the heat input rate ($\dot{Q_i}$) into a crater lake is an important
+part of estimating the total energy budget of a volcanic system
+\cite{Brown1989a}. Similar to a volcano's gas budget, changes in the heat input
+rate can either indicate changes in the magmatic heat source or in the
+hydrothermal system.
 
-\section{Discussion and conclusion}
-Inferring the heat input rate into a crater lake is an important part in
-estimating the total energy budget of a volcanic system \cite{Brown1989a}.
-Similar to a volcano's gas budget, changes in the heat input rate can either
-indicate changes in the magmatic heat source or in the hydrothermal system
-above.
-
-We would expect a significant change in the heat source to cause a change in
-slope of the cumulative heat input rate (Figure \ref{cumulative}). The fact that
-we do not observe such changes may indicate a more or less constant supply of
-heat during the studied period. Experiments on analog models of hydrothermal
-systems suggest that changes in the heat source and the hydrothermal system
-should also change the length of heating cycles
-\cite{vandemeulebrouckAnalogueModelingInstabilities2005,fitzgerald1997TwentySecondWorkshop}.
+Experiments on analog models of hydrothermal systems suggest that
+changes in the heat source and the hydrothermal system should also change the
+length of heating cycles
+\cite{vandemeulebrouckAnalogueModelingInstabilities2005,fitzgeraldStabilityTwoPhaseGeothermal1997}.
 Figure \ref{inference_result}b shows the heating cycles before mid-2017 and
@@
 
-\begin{figure}
- 	\includegraphics[width=12 cm]{figures/cumulative_heat.png}  
-    \caption{Cumulative heat input into RCL (solid line) and its uncertainty 
-             (shaded region) calculated for the time-series in
-             Figure \ref{inference_result} a.} 
-\label{cumulative}
-\end{figure}
+This is further supported by the cumulative difference to the mean for the heat
+input rate as shown in Figure \ref{cumulative}. This type of graph helps
+highlighting periods when the heat input rate was above or below the long-term
+mean. Before 2017 and after 2019 the difference appears to fluctuate more or
+less around the mean value. After a dip beginning in 2017 a period of higher
+than mean values starts towards the end of 2017 and ends early 2020.
 
 To further explore hypotheses like this, the next logical step is to include
-numerical models of magma degassing and hydrothermal heat and fluid flow. While
-beyond the scope of this study, our inversion method is well suited for such
-more complex models as its runtime scales linearly with the number of
-parameters. Several studies have demonstrated the value of non-linear Kalman
-Filters for inverse problems with non-linear, high-dimensional numerical models
+numerical models of magmatic gas and heat release as well as hydrothermal heat
+and fluid flow. Our inversion method is well suited for such more complex models
+as its runtime scales linearly with the number of parameters. Several studies
+have demonstrated the value of non-linear Kalman Filters for inverse problems
+with non-linear, high-dimensional numerical models
 \cite[e.g.][]{White2018a,huangIteratedKalmanMethodology2022}.
@@
 boundary conditions such as negative values for the inflow rate of meteoric
-water ($\dot{M_s}$) \cite{hurstCraterLakeEnergy2015, Stevenson1992}. Their
+water ($\dot{M_s}$) \cite{Hurst1991, hurstCraterLakeEnergy2015}. Their
 explanation was that as rising hot fluids enter the lake, relatively cool lake
 water flows back into the vent. This would lead to an overall lower addition of
-mass to the lake and resulting in an enthalpy value that is double than expected
+mass to the lake and resulting in an enthalpy value that is double the expected
 for saturated steam at hydrostatic pressures likely to be found at the bottom of
@@
 In conclusion, the combination of the mass and energy model, developed in
-previous studies \cite{hurstCraterLakeEnergy2015, Stevenson1992}, with the
+previous studies \cite{Hurst1991,hurstCraterLakeEnergy2015}, with the
 probabilistic inference method we developed here provides a powerful method to
 continuously infer the heat input rate into RCL. It allows us to combine
-disparate data streams, quantify our beliefs in the results and thereby augment
-our monitoring capabilities by keeping a finger on Mt. Ruapehu's energy budget.
+disparate data streams, invert them for probabilistic estimates of the heat
+input rate and thereby keep a finger on Mt. Ruapehu's energy budget.
 
@@
 
+\section*{Acknowledgements}
+The authors would like to thank Nico Fournier for providing feedback and ideas
+throughout the project and Sophie Pearson-Grant for her constructive and
+insightful feedback on this manuscript. GNS Science's volcano monitoring group
+has been an invaluable team to test and improve this work. Graphs have been
+produced using the open-source graphing libraries Plotly
+(https://plotly.com/python/), matplotlib (https://matplotlib.org/) and cartopy
+(https://scitools.org.uk/cartopy/docs/latest/).
+
 \section*{Competing interests}
@@
 \section*{Author's contributions}
-    Text for this section \ldots
+  Y.B. wrote the main manuscript text, produced graphs, and implemented the
+  Kalman Smoothing. S.S. worked on data collection and interpretation. T.H.
+  developed and implemented the mass and energy balance model. All
+  authors reviewed the manuscript.
 
-\section*{Acknowledgements}
-  Text for this section \ldots
+\section*{Availability of data and materials}%% if any
+The source code for this work can be accessed at
+https://github.com/yannikbehr/pumahu. All data used in this study is provided
+free of charge by the GeoNet program (https://www.geonet.org.nz)
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %% Do not use \listoffigures as most will included as separate files
+%\newpage
+\section*{Figures}
 
-%% \section*{Figures}
+\begin{figure}[h!]
+    %\includegraphics[width=12 cm]{figures/overview.png}  
+    \caption{Image of Mt. Ruapehu's crater lake from 2008 (\textcopyright Visual
+        Media Library, GNS Science, ID 6277, photographer Karen Britten). The red
+        circle marks the location of the lake outlet. The location of Mt. Ruapehu is
+        shown as a red triangle on the inset map.}
+      \label{overview}
+\end{figure}
+
+\begin{figure}[h!]
+   %\includegraphics[width=12 cm]{figures/sensitivity_analysis.png}  
+    \caption{Surface energy loss ($Q_e$) with respect to lake water temperature
+             ($T$). The model parameters that have been held fixed are shown in  
+             the title.}
+      \label{sens}
+\end{figure}
+
+\begin{figure}[h!]
+	%\includegraphics[width=12 cm]{figures/synthetic_inversion_uks.png}  
+    \caption{Inversion result for the synthetic test described in Section
+        \ref{syn_test}. Turquoise lines and markers represent synthetic
+        observations; red dashed lines are the inversion results and black
+        dashed lines are the true input to the synthetic test. Error bars and
+        shaded areas represent $\pm$ 3 standard deviations.}
+  \label{syn_example}
+\end{figure}
+
+\begin{figure}[h!]
+ 	%\includegraphics[width=12 cm]{figures/time_frequency_analysis.png}  
+    \caption{ a) Inference results for the heat input rate $\dot{Q_i}$ at RCL
+             between 4 March 2016 and 1 February 2022. The solid line shows the
+             marginal probability and the uncertainty is displayed as shaded
+             region; b) Continuous wavelet transform of the time series shown in
+             a; c) Power spectral density of the heat input rate shown as a
+             solid line with uncertainty displayed as shaded region.}
+\label{inference_result}
+\end{figure}
+
+\begin{figure}[h!]
+ 	%\includegraphics[width=12 cm]{figures/cumulative_heat.png}  
+    \caption{Cumulative difference to the mean for the heat input rate into RCL
+             (solid line) and its uncertainty (shaded region) calculated from
+             the time-series in Figure \ref{inference_result}a.} 
+\label{cumulative}
+\end{figure}
+
 %%   \begin{figure}[h!]

### 2022-04-19 src/pumahu/forward_model.py
Correct forward model

@@ -317,8 +317,8 @@ class Forwardmodel:
         self.evap = (qe, me)
         # energy loss due to outflow
         qo = state[5]*state[0]*cw
-        g0 = 1./(cw*state[1])*(-qe + solar + qi - qo)
         g1 = state[4] + steam - me - state[5]
+        g0 = 1./(cw*state[1])*(-qe + solar + qi - qo) - state[0]/state[1]*g1
         # dX/dt = -M_out*(X_t/m_t)
         # X_t is the total amount of a chemical
         # species at time t

### 2022-04-19 tests/test_fd.py
Add test for historic data

@@ -2,8 +2,10 @@ import unittest
 
 import numpy as np
 
+from pumahu import get_data
 from pumahu.fd import fd 
 from pumahu.syn_model import SynModel, setup_test
+from pumahu.data import LakeData
 
 
 class FDTestCase(unittest.TestCase):
@@ -18,6 +20,17 @@ class FDTestCase(unittest.TestCase):
         q_diff = (q_test[:-1] - q_exp[:-1])
         rmse = np.sqrt(((q_diff) ** 2).mean())
         self.assertAlmostEqual(rmse, 15.64, 2)
+
+    def test_historic_data(self):
+        """
+        Test with dataset used by Hurst et al. [2015]. 
+        """
+        ld = LakeData(csvfile=get_data('data/data.csv'))
+        xdf = ld.get_data('2000-1-1', '2021-1-1',
+                          smoothing={'Mg': 2.6, 'T': 0.4, 'z': 0.01})
+        xdf = xdf.dropna('dates', how='all')
+        rs = fd(xdf, results_file=None, new=False, use_drmg=False)
+        print(rs)
 
 
 if __name__ == '__main__':

### 2022-04-19 tests/test_uks.py
Add more tests for historic data

@@ -90,7 +90,7 @@ class UKSTestCase(unittest.TestCase):
         logl_mean = np.nansum(rs.p_samples.values)/rs.dates.shape[0]
         self.assertAlmostEqual(logl_mean, -6.40177, 4)
         
-    def test_historic_data(self):
+    def test_Hurstetal_data(self):
         """
         Test with dataset used by Hurst et al. [2015]. 
         """
@@ -99,6 +99,29 @@ class UKSTestCase(unittest.TestCase):
         xdf = ld.get_data('2003-1-16', '2010-1-29',
                           smoothing={'Mg': 2.6, 'T': 0.4, 'z': 0.5})
         xdf = xdf.interpolate_na(dim='dates')
+        P0 = OrderedDict(T=1e0, M=1e0, X=1e0, q_in=1e1,
+                 m_in=1e1, m_out=1e1, h=1e-1, W=1e-1,
+                 dqi=1e-2, dMi=1e-3, dMo=1e-3, dH=1e-3, 
+                 dW=1e-3)
+        P0 = np.eye(len(P0))*list(P0.values())
+        Q = OrderedDict(T=1e0, M=1e2, X=1e-1, q_in=3e1,
+                        m_in=1e0, m_out=1e0, h=1e-3, W=1e-3,
+                        dqi=1e-3, dMi=0, dMo=0, dH=0, dW=0)
+        Q = np.eye(len(Q))*list(Q.values())
+        uks = UnscentedKalmanSmoother(data=xdf, Q=Q, P0=P0,
+                                      initvals={'qi': 0., 'm_in': 0., 'm_out': 20., 'X': 5.})
+        xds_uks = uks() 
+
+    def test_pre1995_data(self):
+        """
+        Test with dataset from before 1995-96 eruptions. 
+        """
+        startdate2 = '1993-4-20'
+        enddate2 = '1995-7-4'
+        ld = LakeData(enthalpy=3.0, windspeed=3.5, m_out=0, m_out_err=1.)
+        xdf = ld.get_data(startdate2, enddate2, smoothing={'Mg': 2.6, 'T': 0.4, 'z': 0.5}, ignore_cache=True)
+        xdf = xdf.interpolate_na(dim='dates')
+        
         P0 = OrderedDict(T=1e0, M=1e0, X=1e0, q_in=1e1,
                  m_in=1e1, m_out=1e1, h=1e-1, W=1e-1,
                  dqi=1e-2, dMi=1e-3, dMo=1e-3, dH=1e-3, 

### 2022-04-19 tests/test_forward_model.py
still work in progress

@@ -4,7 +4,10 @@ import unittest
 import numpy as np
 
 from pumahu import Forwardmodel, SynModel
-from pumahu.syn_model import setup_test
+from pumahu.syn_model import (setup_test,
+                              setup_realistic,
+                              resample,
+                              make_sparse)
 
 
 class ForwardModelTestCase(unittest.TestCase):
@@ -249,6 +252,12 @@ class ForwardModelTestCase(unittest.TestCase):
         self.assertEqual(xds.shape, (4,4))
         self.assertEqual(list(xds.params.values),
                          ['q_in', 'm_in', 'h', 'W'])
+        
+
+    def test_resample(self):
+        xds = SynModel().run(setup_realistic(sinterval=120), addnoise=True, ignore_cache=True)        
+        na = resample(xds)
+        na = make_sparse(na, ['m_out', 'm_in', 'X'])
 
 
 if __name__ == '__main__':

### 2022-04-19 job_scheduler.py
initial commit

@@ -1,0 +1,15 @@
+import schedule
+import subprocess
+import time
+
+def job():
+    subprocess.run(['heat_mcmc', '--rdir', '/opt/data', '-f', '-p', '-d'])
+    subprocess.run(['heat_uks', '--rdir', '/opt/data',
+                    '-f', '-p', '-d',
+                    '--benchmark', '/opt/data/mcmc_sampling.nc'])
+
+schedule.every().day.at("11:00").do(job)
+
+while True:
+    schedule.run_pending()
+    time.sleep(10)

### 2022-04-19 setup.py

@@ -1,0 +1,25 @@
+from ez_setup import use_setuptools
+use_setuptools()
+from setuptools import setup
+
+####################################################################
+#                    CONFIGURATION
+####################################################################
+
+# do the build/install
+setup(
+    name="pumahu",
+    version="0.1",
+    description="Python package to compute crater lake energy and mass balance.",
+    long_description="Python package to compute crater lake energy and mass balance.",
+    author="Yannik Behr",
+    author_email="yannikbehr@yanmail.de",
+    url="",
+    license="GPL v3",
+    package_dir={'': 'src'},
+    packages=['pumahu'],
+    package_data={'pumahu': ['data/*.dat', 'data/*.csv', 'data/*.npz', 'notebook/*.ipynb']},
+    entry_points={'console_scripts':
+                  ['heat_mcmc=pumahu.mcmc:main',
+                   'heat_uks=pumahu.uks:main']}
+)

### 2022-04-19 src/pumahu/__init__.py

@@ -1,0 +1,11 @@
+import pkg_resources
+
+
+def get_data(filename):
+    return pkg_resources.resource_filename(__name__, filename)
+
+from .forward_model import Forwardmodel
+from .mcmc import ns_sampling 
+from .syn_model import SynModel
+from .uks import UnscentedKalmanSmoother
+from .data import LakeData

### 2022-04-19 src/pumahu/data.py

@@ -1,0 +1,720 @@
+from collections import defaultdict, abc
+from datetime import date, datetime, timezone, timedelta
+from functools import lru_cache
+import inspect
+import os
+
+from cachier import cachier
+import numpy as np
+import pandas as pd
+from scipy.interpolate import interp1d
+from filterpy.kalman import UnscentedKalmanFilter as UKF
+from filterpy.kalman import KalmanFilter as KF
+from filterpy.kalman import MerweScaledSigmaPoints
+from filterpy.common import Q_continuous_white_noise
+import xarray as xr
+
+from . import Forwardmodel, get_data
+
+def fits_hash(args, kwargs):
+    key = []
+    for _a in args[1:]:
+        if isinstance(_a, abc.Mapping):
+            key.append(sorted(_a.items()))
+        else:
+            key.append(_a)
+    key = tuple(key)    
+    for k,v in sorted(kwargs.items()):
+        if isinstance(v, abc.Mapping):
+            key += tuple((k, tuple(sorted(v.items()))))
+        else:
+            key += tuple((k,v))
+    hashValue = hash(key)
+    return hashValue
+
+class LakeData:
+    """
+    Load the lake measurements from the FITS database.
+    """
+
+    def __init__(self, url="https://fits.geonet.org.nz/observation",
+                 csvfile=None, windspeed=5., m_in=np.nan, m_in_err=np.nan,
+                 q_in=np.nan, q_in_err=np.nan, m_out=10., m_out_err=.25,
+                 enthalpy=3.):
+        """
+        Parameters:
+        -----------
+        url : str
+              FITS url to retrieve data from
+        csvfile : str
+                  If not None, read data from a csv file instead of
+                  requesting it from FITS
+        windspeed : float
+                  Default windspeed [m/s].
+        enthalpy : float
+                   Default enthalpy [MJ/kg/day]
+        m_out : Default outflow rate [kt/day]
+        """
+        self.base_url = url
+        self.xdf = None
+        self.ws = windspeed
+        self.ws_err = 0.5
+        self.m_in = m_in
+        self.m_in_err = m_in_err 
+        self.m_out = m_out
+        self.m_out_err = m_out_err 
+        self.q_in = q_in
+        self.q_in_err = q_in_err 
+        self.h = enthalpy
+        self.h_err = 0.01
+        self.prms = ['T', 'z', 'Mg', 'X', 'V',
[…]

### 2022-04-19 src/pumahu/fd.py

@@ -1,0 +1,131 @@
+"""
+Compute energy and mass balance for crater lakes from measurements of water
+temperature, wind speed and chemical dilution.
+"""
+
+import os
+
+import numpy as np
+from numpy.ma import masked_invalid, masked_less
+import pandas as pd
+from scipy.interpolate import interp1d
+from scipy.stats import gaussian_kde
+from scipy.integrate import trapz, cumtrapz
+import xarray as xr
+
+from nsampling import (NestedSampling, Uniform,
+                       Normal, InvCDF, Constant)
+
+from filterpy.kalman import UnscentedKalmanFilter as UKF
+from filterpy.kalman import MerweScaledSigmaPoints
+from filterpy.common import Q_continuous_white_noise
+
+from pumahu.forward_model import Forwardmodel
+
+
+def el(t1, t2, vol, cw):
+    """
+    Change in Energy stored in the lake [TJ]
+    """
+    return (t2 - t1) * vol * cw
+
+
+def fd(data, results_file=None, new=False, use_drmg=False,
+       level2volume=None):
+    """
+    Compute the amount of steam and energy that has to be put into a crater
+    lake to cause an observed temperature change. This computation runs
+    the model backward using a finite difference approach.
+    """
+    if results_file is not None:
+        if not new and os.path.isfile(results_file):
+            res = xr.open_dataset(results_file)
+            res.close()
+            return res
+    parameters = ['q_in', 'm_in', 'm_out',
+                  'h', 'T', 'M', 'X', 'dq_in']
+    nparams = len(parameters)
+    dates = pd.to_datetime(data['dates'].values)
+    ndates = dates.size
+    exp = np.zeros((ndates, nparams, 2))*np.nan
+    
+    data_z = data.loc[:, 'z', 'val'].values
+    data_T = data.loc[:, 'T', 'val'].values
+    data_mg = data.loc[:, 'Mg', 'val'].values
+    data_W = data.loc[:, 'W', 'val'].values
+    data_h = data.loc[:, 'h', 'val'].values
+    fm = Forwardmodel(level2volume=level2volume)
+    cw=0.0042
+
+    # es returns nan entries if wind is less than 0.0
+    nd = (dates[1:] - dates[:-1])/pd.Timedelta('1D')
+    # time interval in Megaseconds
[…]

### 2022-04-19 src/pumahu/forward_model.py

@@ -1,0 +1,334 @@
+from datetime import datetime, timedelta
+
+import numpy as np
+from scipy.optimize import brentq
+
+
+class Forwardmodel:
+
+    def __init__(self, method='euler', level2volume=None, mass2area=None):
+        """
+        Physical model of the crater lake energy and mass balance.
+
+        Besides the physical model this class also provides different methods
+        to integrate the ODEs using either Eulers method, 2nd order Runge-Kutta
+        or 4th order Runge-Kutta.
+
+        Parameters
+        ----------
+        method : string
+                 The integration method. Can be either 'euler'
+                 (Euler's method), 'rk2' (2nd order Runge-Kutta),
+                 or 'rk4' (4th order Runge-Kutta)
+
+        level2volume : function
+                       A function to compute lake volume and surface area from
+                       the lake level. Takes the lake level as input and return
+                       lake surface area and volume (in that order).
+
+        mass2area : function
+                    A function to compute lake surface area from the lake mass
+                    and temperature. Takes the mass and temperature as input
+                    and return surface area and volume (in that order).
+        """
+
+        self.steam = None
+        self.evap = None
+        if method == 'euler':
+            self.int_method = self.euler
+        elif method == 'rk2':
+            self.int_method = self.rk2
+        elif method == 'rk4':
+            self.int_method = self.rk4
+        else:
+            msg = 'Please choose one of the following'
+            msg += 'integration methods: euler, rk2, rk4.'
+            raise ValueError(msg)
+
+        if level2volume is None:
+            self.level2volume = self.fullness
+        else:
+            self.level2volume = level2volume
+
+        if mass2area is None:
+            self.mass2area = self.inverse_fullness
+        else:
+            self.mass2area = mass2area
+
+    def get_evap(self):
+        return self.evap
+
+    def get_steam(self):
+        return self.steam
+
+    def surface_loss(self, T, w, a, T_air=0.9, l=500., vp_air=6.5,
+                     vd_air=0.0022, pa=800., ca=1005., da=0.948,
+                     lh=2400000.):
+        """
+        Energy and mass loss rate from lake surface.
+
[…]

### 2022-04-19 src/pumahu/mcmc.py

@@ -1,0 +1,331 @@
+"""
+Compute energy and mass balance for crater lakes from measurements of water
+temperature, wind speed and chemical dilution.
+"""
+from argparse import ArgumentParser
+from collections import OrderedDict
+from datetime import datetime, timedelta
+import logging
+import os
+import time
+
+import numpy as np
+import pandas as pd
+import schedule
+from scipy.interpolate import interp1d
+from tqdm import tqdm
+import xarray as xr
+
+from nsampling import (NestedSampling, Uniform,
+                       Normal, InvCDF, Constant)
+
+from . import get_data 
+from .forward_model import Forwardmodel
+from .data import LakeData
+from .visualise import (trellis_plot,
+                        mcmc_heat_input,
+                        heat_vs_rsam)
+
+
+logging.basicConfig(level=logging.DEBUG, format='%(asctime)s %(message)s')
+
+
+class LikeliHood:
+
+    def __init__(self, data, date, dt, ws, cov, intmethod='rk4',
+                 mass2area=None):
+        self.data = data
+        self.date = date
+        self.dt = dt
+        self.ws = ws
+        # the precision matrix is the inverse of the
+        # covariance matrix
+        self.prec = np.linalg.inv(cov)
+        self.factor = -np.log(np.power(2.*np.pi, cov.shape[0])
+                              + np.sqrt(np.linalg.det(cov)))
+        self.samples = []
+        self.fm = Forwardmodel(method=intmethod,
+                               mass2area=mass2area)
+
+    def get_samples(self):
+        return np.array(self.samples)
+
+    def run_lh(self, vals, sid):
+        """
+        Callback function to compute the log-likelihood for nested sampling.
+        """
+        try:
+            qi = vals[0]*0.0864
+            Mi = vals[1]
+            Mo = vals[2]
+            H = vals[3]
+            T = vals[4]
+            M = vals[5]
+            X = vals[6]
+            dq = vals[7]
+            dp = np.zeros(5)
+            dp[0] = dq*0.0864
+            y0 = np.array([T, M, X, qi, Mi, Mo, H, self.ws])
+            y_new = self.fm.integrate(y0, self.date, self.dt, dp)
+            lh = self.factor - 0.5*np.dot(y_new[0:3]-self.data,
+                                          np.dot(self.prec,
+                                                 y_new[0:3]-self.data))
[…]

### 2022-04-19 src/pumahu/syn_model.py

@@ -1,0 +1,464 @@
+import hashlib
+from datetime import date, timedelta
+
+from cachier import cachier
+import numpy as np
+import pandas as pd
+from scipy.signal import tukey
+from scipy.stats import gamma
+import xarray as xr
+from tqdm import tqdm
+
+from .forward_model import Forwardmodel
+
+
+def setup_test(q_in=1000.):
+    """
+    A synthetic scenario for unit testing.
+
+    Parameters:
+    -----------
+    q_in: float 
+          Maximum heat input rate [MW]
+
+    Returns:
+    --------
+    :class:`xarray.DataArray`
+            An xarray DataFrame containing the input
+            data for the synthetic model
+    """
+    dates = pd.date_range(start='2017-01-01', end='2017-01-04',
+                          freq='D')
+    data = np.zeros((4,4))
+    params = ['q_in', 'm_in', 'h', 'W']
+    # qi
+    data[0, :] = np.array([0., .2, .6, 1.])*q_in
+    # Mi
+    data[1, :] = np.ones(4)*10.
+    # H
+    data[2, :] = np.ones(4)*6
+    # W
+    data[3, :] = np.ones(4)*4.5
+    return xr.DataArray(data, coords=[('params', params),
+                                      ('times', dates)])
+
+
+def setup_realistic(q_in=1000., tmax=90., sinterval=10):
+    """
+    Synthetic scenario where the heat influx is a
+    combination of 3rd order polynomial and sinusoid.
+
+    :param q_in: Maximum heat input rate [MW]
+    :type q_in: float
+    :param tmax: Length of the time-series [days]
+    :type tmax: float
+    :param sinterval: Sampling interval [mins]
+    :type sinterval: int
+    :returns: An xarray DataFrame containing the input
+              data for the synthetic model
+    :rtype: :class:`xarray.DataArray`
+    """
+    dt = (sinterval * 60)/86400.
+    t = np.arange(0,tmax+dt, dt)
+    dates = pd.date_range(start='2020-03-03', freq='%dmin'%sinterval,
+                          periods=t.size)
+    a = 1.
+    b = -20
+    c = b**2/4+3
+    y = 1./(a*t**2 + b*t + c)
+    y = q_in*y/max(y)
+
+    npts = y.size
+    y_last = y[npts//2]
+    f = 1./20.
+    t1 = t[npts//2::]
[…]

### 2022-04-19 src/pumahu/uks.py

@@ -1,0 +1,403 @@
+"""
+Solve the mass and energy balance model using
+the Unscented Kalman Smoother.
+"""
+from collections import OrderedDict
+from datetime import datetime, timedelta
+from functools import partial
+import os
+
+from filterpy.kalman import (UnscentedKalmanFilter,
+                             unscented_transform)
+
+import numpy as np
+import pandas as pd
+from tqdm import tqdm
+import xarray as xr
+
+from . import get_data 
+from .syn_model import SynModel
+from .forward_model import Forwardmodel
+from .sigma_points import MerweScaledSigmaPoints
+from .data import LakeData, get_outflow
+from .visualise import trellis_plot, plot_qin_uks
+ 
+import ipdb
+
+
+class Fx:
+
+    def __init__(self, nparams, ngrads, test=False):
+        """
+        Run the forward model.
+
+        Parameters:
+        -----------
+        nparams : int
+                  Number of free parameters
+        ngrads : int
+                 Number of free gradients
+        test: bool
+              If 'True' use a mass2area function
+              corresponding to the synthetic test setup.
+        """
+        self.nparams = nparams
+        self.ngrads = ngrads
+        if test:
+            s = SynModel()
+            self.fm = Forwardmodel(method='rk4', mass2area=s.mass2area)
+        else:
+            self.fm = Forwardmodel(method='rk4')
+
+    def run(self, x, dt, date):
+        """
+        Forward model lake temperature
+        """
+        y = np.zeros(self.nparams)
+        y[:] = x[0:self.nparams]
+        dp = [0.] * self.ngrads 
+        dp[:] = x[self.nparams:]
+        y_next = self.fm.integrate(y, date, dt, dp)
+        return np.r_[y_next[0:self.nparams], x[self.nparams:]]
+
+
+def h_x(x):
+    """
+    Measurement function
+    """
+    return [x[0], x[1], x[2], x[4], x[5], x[6], x[7]]
+
+
+class UnscentedKalmanSmoother:
+
+    def __init__(self, data, X0=None, P0=None, Q=None):
+        self.dates = pd.to_datetime(data['dates'].values)
+        self.ndates = len(self.dates)
+        self.dt = (self.dates[1] - self.dates[0])/pd.Timedelta('1D')
[…]

### 2022-04-19 src/pumahu/visualise.py

@@ -1,0 +1,397 @@
+"""
+Code to visualize results from the crater lake energy and mass balance
+computations.
+"""
+from collections import defaultdict
+import os
+import urllib
+import ssl
+
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+import matplotlib.dates as mdates
+import matplotlib.lines as mlines
+import numpy as np
+from obspy import read, UTCDateTime, Stream
+import pandas as pd
+import plotly.graph_objects as go
+from plotly.subplots import make_subplots
+import plotly.io as pio
+from sklearn.neighbors import KernelDensity
+
+from pumahu import get_data
+
+
+class TrellisPlot:
+    def __init__(self):
+        self.line_colours = ['rgb(26,166,183)',
+                             'rgb(255,65,77)',
+                             'rgb(0,45,64)']
+        self.line_dash = ['solid', 'dash', 'dash'] 
+        self.error_colours = ['rgba(26,166,183,.3)',
+                              'rgba(255,65,77, .3)',
+                              'rgba(0,45,64,.3)']
+        self.legend_traces = []
+        self._ntraces = defaultdict(lambda: -1)
+
+    def get_traces(self, data, key, err_min=1e-3,
+                   nanthresh=0.4, dropzeros=True):
+        """
+        Extract traces
+        
+        Parameters:
+        -----------
+        data : :class:`xarray.DataArray`
+               Data structure that contains the values
+               and error estimates.
+        key : str
+              Dimension name of the data to extract.
+        err_min : float
+                  Minimum error to plot in case max
+                  and min traces are too close.
+        nanthresh : float
+                    Interpolate if the number of NaNs is
+                    less than nanthresh.
+        """
+        if 'val_std' in data.dims:
+            # ignore warnings due to NaNs
+            ymean = data.loc[:, key, 'val'].values
+            if dropzeros:
+                ymean = np.where(ymean < 0., 0., ymean)
+            yerr = data.loc[:, key, 'std'].values
+            # Interpolate over negative values
+            idx = np.where(yerr<0)[0]
+            yerr[idx] = np.nan
+            # calculate the percentage of NaN values
+            nanperc = float(np.sum(np.isnan(ymean))/data.shape[0])
+            if nanperc < nanthresh:
+                yerr = pd.Series(yerr).interpolate().values
+                ymean = pd.Series(ymean).interpolate().values
+            ymin = ymean - 3*yerr
+            if dropzeros:
+                ymin = np.where(ymin < 0., 0., ymin)
[…]

### 2022-04-19 tests/test_data.py

@@ -1,0 +1,246 @@
+from collections import defaultdict
+from datetime import datetime
+import inspect
+import os
+import unittest
+import warnings
+
+import numpy as np
+import pandas as pd
+import xarray as xr
+
+from pumahu import get_data
+from pumahu.data import LakeData, WindData
+
+
+class DataTestCase(unittest.TestCase):
+
+    def load_test_results(self):
+        with open(os.path.join(self.data_dir, 'LAKEOUTm.DAT')) as f:
+            # skip header lines
+            for i in range(4):
+                f.readline()
+            # now read until the next empty line
+            data = defaultdict(list)
+            dates = []
+            while True:
+                l = f.readline()
+                if l == '\n':
+                    break
+                yr, mon, day, temp, stfl, pwr, evfl, fmelt, inf, drm, mmp,\
+                    fmg, fcl, o18, o18m, o18f, h2, h22, m = l.split()
+                dates.append(
+                    np.datetime64('{}-{:02d}-{:02d}'.format(
+                        yr, int(mon), int(day))))
+                data['t'].append(float(temp))
+                data['steam'].append(float(stfl))
+                data['pwr'].append(float(pwr))
+                data['evfl'].append(float(evfl))
+                data['fmelt'].append(float(fmelt))
+                data['inf'].append(float(inf))
+                data['fmg'].append(float(fmg))
+                data['fcl'].append(float(fcl))
+                data['mass'].append(float(m))
+
+        return pd.DataFrame(data, index=dates)
+
+    def load_input(self):
+        with open(os.path.join(self.data_dir, 'input.dat')) as f:
+            # skip header line
+            f.readline()
+
+            data = defaultdict(list)
+            while True:
+                l = f.readline()
+                if l == '\n':
+                    break
+                if not l:
+                    break
+                try:
+                    yr, mo, dy, t, h, fl, m, c, dv, nd, w, o18, h2 = l.split()
+                except:
+                    print(l)
+                    raise
+                data['date'].append(np.datetime64('{}-{:02d}-{:02d}'.format(
+                    yr, int(mo), int(dy))))
+                data['temp'].append(float(t))
+                data['hgt'].append(float(h))
+                data['mg'].append(float(m))
+                data['cl'].append(float(c))
+                data['wind'].append(float(w))
+                data['o18'].append(float(o18))
[…]

### 2022-04-19 tests/test_fd.py

@@ -1,0 +1,37 @@
+import unittest
+
+import numpy as np
+
+from pumahu import get_data
+from pumahu.fd import fd 
+from pumahu.syn_model import SynModel, setup_test
+from pumahu.data import LakeData
+
+
+class FDTestCase(unittest.TestCase):
+
+    def test_synthetic_euler(self):
+        sm = SynModel(integration_method='euler')
+        data = sm.run(setup_test(), ignore_cache=True)
+        rs = fd(data.exp, results_file=None, new=False, use_drmg=False,
+                level2volume=sm.synth_fullness)
+        q_test = rs.input.loc[:, 'q_in', 'val'].values
+        q_exp = rs.exp.loc[:, 'q_in', 'val'].values
+        q_diff = (q_test[:-1] - q_exp[:-1])
+        rmse = np.sqrt(((q_diff) ** 2).mean())
+        self.assertAlmostEqual(rmse, 15.64, 2)
+
+    def test_historic_data(self):
+        """
+        Test with dataset used by Hurst et al. [2015]. 
+        """
+        ld = LakeData(csvfile=get_data('data/data.csv'))
+        xdf = ld.get_data('2000-1-1', '2021-1-1',
+                          smoothing={'Mg': 2.6, 'T': 0.4, 'z': 0.01})
+        xdf = xdf.dropna('dates', how='all')
+        rs = fd(xdf, results_file=None, new=False, use_drmg=False)
+        print(rs)
+
+
+if __name__ == '__main__':
+    unittest.main()

### 2022-04-19 tests/test_forward_model.py

@@ -1,0 +1,264 @@
+from datetime import datetime
+import unittest
+
+import numpy as np
+
+from pumahu import Forwardmodel, SynModel
+from pumahu.syn_model import (setup_test,
+                              setup_realistic,
+                              resample,
+                              make_sparse)
+
+
+class ForwardModelTestCase(unittest.TestCase):
+
+    def test_esol(self):
+        fm = Forwardmodel()
+        e = fm.esol(2e5, datetime(2003,1,19,0,0,0))
+        self.assertEqual(e, 4.5)
+
+    def test_surface_loss(self):
+        fm = Forwardmodel()
+        loss, ev = fm.surface_loss(35.0, 5.0, 200000)
+        self.assertAlmostEqual(loss, 20.61027, 5)
+        self.assertAlmostEqual(ev, 4.87546, 5)
+    
+    def test_fullness(self):
+        fm = Forwardmodel()
+        a, vol = fm.fullness(np.ones(10) * 2529.4)
+        fvol = np.ones(10) * 8880.29883
+        diffvol = abs(vol - fvol) / fvol * 100.
+        fa = np.ones(10) * 196370.188
+        diffa = abs(a - fa) / fa * 100.
+        # Probably due to different precisions the numbers between the
+        # original Fortran code and the Python code differ slightly
+        self.assertTrue(np.all(diffvol < 0.0318))
+        self.assertTrue(np.all(diffa < 0.000722))
+
+    def test_inverse_fullness(self):
+        fm = Forwardmodel()
+        a, vol, z = fm.inverse_fullness(np.array([8880]), 20.)
+        self.assertAlmostEqual(z, 2529.548, 3)
+
+    def test_without_q_gradient(self):
+        """
+        Test forward model assuming no gradient
+        of the heat input rate between time steps.
+        """
+        # First test with Euler
+        T0 = 30.
+        dt = 1.
+        Qi0 = 400. * 0.0864
+        dQi = 0.
+        Mi = 10
+        dMi = 0.
+        Mo = 5.
+        dMo = 0.
+        H = 2.8
+        dH = 0.
+        ws = 4.5
+        dws = 0.
+        density = 1.003 - 0.00033 * T0
+        M0 = 8875.*density
+        X0 = 2.
+        dtime = datetime(2019, 1, 1, 0, 0, 0)
[…]

### 2022-04-19 tests/test_mcmc.py

@@ -1,0 +1,150 @@
+from collections import defaultdict
+import inspect
+import os
+import tempfile
+import unittest
+import warnings
+
+import numpy as np
+import pandas as pd
+import pytest
+
+from pumahu.mcmc import ns_sampling, main
+from pumahu.syn_model import SynModel, setup_test
+
+
+class MCMCTestCase(unittest.TestCase):
+
+    @pytest.mark.slow
+    def test_synthetic_euler(self):
+        data = SynModel(integration_method='euler').run(setup_test(),
+                                                        ignore_cache=True)
+        rs = ns_sampling(data.exp, nsamples=2000, nresample=-1,
+                         q_in_lim=(0., 1500.), m_in_lim=(0., 40.),
+                         m_out_lim=(0., 40.), new=True, tolZ=1e-3, 
+                         tolH=3e30, H=6., ws=4.5, seed=42, intmethod='euler',
+                         gradient=False)
+        q_in_val = rs['exp'].loc[:, 'q_in', 'val'].data
+        q_in_var = rs['exp'].loc[:, 'q_in', 'std'].data
+        z_val = rs.p_samples.loc[:, :, 'Z'].max(axis=1).data 
+        z_var = rs.p_samples.loc[:, :, 'Z_var'].max(axis=1).data 
+        np.testing.assert_array_almost_equal(q_in_val,
+                                             np.array([150.20788797,
+                                                       308.42346672,
+                                                       626.16570884,
+                                                       np.nan]),
+                                             decimal=6)
+        np.testing.assert_array_almost_equal(q_in_var,
+                                             np.array([10452.13754637,
+                                                       39601.25850057,
+                                                       54910.75889361,
+                                                       np.nan]),
+                                             decimal=6)
+        np.testing.assert_array_almost_equal(z_val,
+                                             np.array([-9.54064536,
+                                                       -9.92588325,
+                                                       -9.33493868,
+                                                       np.nan]),
+                                             decimal=6)
+        np.testing.assert_array_almost_equal(z_var,
+                                             np.array([0.17426234,
+                                                       0.17471695,
+                                                       0.1573276,
+                                                       np.nan]),
+                                             decimal=6)
+
+    @pytest.mark.slow
+    def test_synthetic_rk2(self):
+        """
+        Test the inversion of a synthetic model using fourth-order
+        Runge-Kutta integration.
+        """
+        data = SynModel(integration_method='rk2').run(setup_test(),
+                                                      ignore_cache=True)
[…]

### 2022-04-19 tests/test_notebooks.py

@@ -1,0 +1,40 @@
+import inspect
+import os
+import unittest
+import nbformat
+from nbconvert.preprocessors import ExecutePreprocessor
+
+
+class NBTestCase(unittest.TestCase):
+
+    def setUp(self):
+        self.cwd = os.path.dirname(os.path.abspath(
+            inspect.getfile(inspect.currentframe())))
+        self.nbdir = os.path.join(self.cwd, '..', 'src', 'pumahu', 'notebook')
+
+    def test_estimation(self):
+        nbfn = os.path.join(self.nbdir, 'estimation.ipynb')
+        with open(nbfn) as f:
+            nb = nbformat.read(f, as_version=4)
+        ep = ExecutePreprocessor(timeout=1200, kernel_name='python3')
+        ep.preprocess(nb, {'metadata': {'path': self.nbdir}})
+
+    def test_synthetics(self):
+        nbfn = os.path.join(self.nbdir, 'forward_model.ipynb')
+        with open(nbfn) as f:
+            nb = nbformat.read(f, as_version=4)
+        ep = ExecutePreprocessor(timeout=1200, kernel_name='python3')
+        ep.preprocess(nb, {'metadata': {'path': self.nbdir}})
+
+    def test_outflow(self):
+        nbfn = os.path.join(self.nbdir, 'Outflow_estimate.ipynb')
+        with open(nbfn) as f:
+            nb = nbformat.read(f, as_version=4)
+        ep = ExecutePreprocessor(timeout=1200, kernel_name='python3')
+        ep.preprocess(nb, {'metadata': {'path': self.nbdir}})
+
+
+
+if __name__ == '__main__':
+    unittest.main()
+

### 2022-04-19 tests/test_uks.py

@@ -1,0 +1,214 @@
+from collections import OrderedDict
+import numpy as np
+import unittest
+import pandas as pd
+import xarray as xr
+
+from filterpy.kalman import unscented_transform
+
+from pumahu import get_data
+from pumahu.uks import UnscentedKalmanSmoother
+from pumahu.syn_model import (SynModel,
+                              setup_test,
+                              setup_realistic)
+from pumahu.data import LakeData
+from pumahu.sigma_points import MerweScaledSigmaPoints
+
+
+class UKSTestCase(unittest.TestCase):
+
+    def setUp(self):
+        self.xds = SynModel(integration_method='rk4').run(setup_test(),
+                                                          gradient=True)
+        self.Q = OrderedDict(T=1e-3, M=1e-3, X=1e-3, q_in=1e-10,
+                             m_in=1e1, m_out=1e1, h=1e-3, W=1e1,
+                             dqi=1e3, dMi=1e4, dMo=1e4, dH=1e-3, dW=1e1)
+        self.Q = np.eye(len(self.Q))*list(self.Q.values())
+
+        self.P0 = OrderedDict(T=1e3, M=1e3, X=1e3, q_in=1e3,
+                              m_in=1e3, m_out=1e3, h=1e-1, W=1e-1,
+                              dqi=1e3, dMi=1e3, dMo=1e3, dH=1e-1, dW=1e-1)
+        self.P0 = np.eye(len(self.P0))*list(self.P0.values())
+
+        self.xds_nan = self.xds.copy(deep=True)
+        self.xds_nan.exp[2, 0, 0] = np.nan
+        self.xds_nan.exp[2, 0, 1] = np.nan
+
+    def test_init(self):
+        """
+        Test that the init function works correctly
+        """
+        uks = UnscentedKalmanSmoother(data=self.xds.exp)
+        self.assertEqual(len(uks.X0), uks.nvar)
+        self.assertTrue(isinstance(uks.X0, list))
+
+    def test_synthetic(self):
+        uks = UnscentedKalmanSmoother(data=self.xds.exp)
+        res = uks(test=True)
+        log_lh = np.nansum(res['p_samples'].loc[dict(p_parameters='lh')].values)
+        self.assertAlmostEqual(log_lh, -29.2659392, 4) 
+        
+    def test_synthetic_extensive(self):
+        """
[…]

### 2022-04-19 tests/test_visualise.py

@@ -1,0 +1,24 @@
+import unittest
+
+import xarray as xr
+
+from pumahu.uks import UnscentedKalmanSmoother
+from pumahu.syn_model import (SynModel,
+                              setup_test,
+                              setup_realistic,
+                              resample,
+                              make_sparse)
+from pumahu.visualise import trellis_plot
+
+
+class VisTestCase(unittest.TestCase):
+
+    def test_trellis_plot(self):
+        xds4 = SynModel().run(setup_realistic(sinterval=120), addnoise=True)
+        na = resample(xds4)
+        na = make_sparse(na, ['m_out', 'X'])
+        fig = trellis_plot(xds4, xr.Dataset({'exp': na}))
+        fig.write_image('./trellis_plot.png')
+
+if __name__ == '__main__':
+    unittest.main()

### 2022-04-19 buildnrun.sh
Fix deploy and update build

@@ -8,7 +8,6 @@
 IMAGE=pumahu
 TAG=0.1
 BUILD=false
-PUSH=false
 INTERACTIVE=false
 
 function usage(){
@@ -22,8 +21,6 @@ Optional Arguments:
     -i, --interactive       Start the container with a bash prompt.
     --image                 Provide alternative image name.
     --tag                   Provide alternative tag
-    --push                  Push to registry. Note: the registry
-                            has to be part of the image name
 EOF
 }
 
@@ -35,7 +32,6 @@ do
         -i | --interactive) INTERACTIVE=true;;
         --image) IMAGE="$2";shift;;
         --tag) TAG="$2";shift;;
-        --push) PUSH=true;;
         -h) usage; exit 0;;
         -*) usage; exit 1;;
 esac
@@ -48,11 +44,6 @@ if [ "${BUILD}" == "true" ]; then
     docker build --build-arg NB_USER=$(whoami) \
         --build-arg NB_UID=$(id -u) \
         -t "${IMAGE}:${TAG}" .
-fi
-
-if [ "${PUSH}" != "false" ]; then
-    docker tag ${IMAGE}:${TAG} huta17-d.gns.cri.nz:5000/yannik/pumahu:${TAG}
-    docker push huta17-d.gns.cri.nz:5000/yannik/pumahu:${TAG}
 fi
 
 if [ "${INTERACTIVE}" == "true" ]; then

### 2022-04-19 docker_build.sh

@@ -1,0 +1,51 @@
+#!/bin/bash
+
+set -e
+
+###### Standard section to container images ####
+PARAMS=""
+
+while (( "$#" )); do
+  case "$1" in
+    -i|--image)
+      if [ -n "$2" ] && [ "${2:0:1}" != "-" ]; then
+        IMAGE=$2
+        shift 2
+      else
+        echo "Error: Argument for $1 is missing" >&2
+        exit 1
+      fi
+      ;;
+    -h|--help)
+      echo "Some text explaining the options"
+      exit 0
+      ;;
+    -*) # unsupported flags
+      echo "Error: Unsupported flag $1" >&2
+      exit 1
+      ;;
+    *) # preserve positional arguments
+      PARAMS="$PARAMS $1"
+      shift
+      ;;
+  esac
+done
+
+# set positional arguments in their proper place
+eval set -- "$PARAMS"
+
+## Find target server
+if [ -z "$IMAGE" ]
+then
+  echo "Image name is required"
+  exit 1
+fi
+
+echo "Building $IMAGE"
+
+DOCKER_BUILDKIT=1 docker build . -t "$IMAGE"
+
+echo "Pushing $IMAGE"
+docker image push "$IMAGE"
+
+exit 0

### 2022-04-19 docker_deploy.sh

@@ -1,11 +1,5 @@
 #!/bin/bash
 
-######## Set up default project variables #######
-IMAGE='huta17-d.gns.cri.nz:5000/yannik/pumahu:0.1'
-APP_NAME=Pumahu
-APP_SERVER="Vulkan"
-TEAM="Volcano"
-#################################################
 ####### Docker config ###########################
 DOCKER_CONFIG=(HostConfig:='{"RestartPolicy": {"Name":"always" } , "Binds": [ "pumahu_data:/opt/data" ] }' \
 Cmd:='["python", "./pumahu/job_scheduler.py"]')
@@ -95,23 +89,16 @@ then
 fi
 
 # Get the API auth token
-if [[ -z "$portainer_user" || -z "$portainer_auth" ]]
+if [[ -z "$portainer_auth" ]]
 then
-  echo "Environment variables undefined"
+  echo "Environment auth token undefined"
   exit 2
 else
-  echo "Building as $portainer_user and deploying to $APP_SERVER"
+  echo "Deploying to $APP_SERVER"
 fi
 
-TOKEN=$(https POST "$PORTAINER_HOST"/api/auth Username="$portainer_user" Password="$portainer_auth" $STD_HTTPS_OPTS | jq .jwt -r)
+AUTH="X-API-key:$portainer_auth"
 
-if [[ -z "$TOKEN" ]]
-then
-  echo "unauthorised access"
-  exit 3
-fi
-
-AUTH="Authorization: Bearer $TOKEN"
 ## Find the target server id
 APP_SERVER_ID=$( https GET "$PORTAINER_HOST"/api/endpoints "$AUTH" $STD_HTTPS_OPTS -b | jq --arg APP_SERVER "$APP_SERVER" '.[] | select(.Name == $APP_SERVER) | .Id')
 

### 2022-04-19 manuscript/figures/psd_analysis.py
update Makefiles to build manuscript from scratch; remove obsolete scripts

@@ -1,42 +1,0 @@
-import plotly.graph_objects as go
-import numpy as np
-from mtspec import mtspec
-from scipy import signal
-import xarray as xr
-
-fn = 'data/uks.nc'
-xdf = xr.open_dataset(fn)
-data = xdf.exp.loc[:, 'q_in', 'val'].values
-data -= data.mean()
-spec, freq, jackknife, _, _ = mtspec(data=data, delta=1.0, time_bandwidth=3.5,
-    number_of_tapers=10, nfft=2048, statistics=True)
-idx = freq<0.2
-freq = np.r_[1./freq[1:], 0]
-
-fig = go.Figure()
-fig.add_trace(go.Scatter(x=freq, y=spec, name='Multi taper', mode='lines',
-                         showlegend=False))
-if False:
-    freqs, psd = signal.welch(data, nfft=2048, nperseg=256)
-    freqs = np.r_[1./freqs[1:], 0]
-    fig.add_trace(go.Scatter(x=freqs, y=psd, mode='lines', line_dash='dash',
-                             name='Welch'))
-fig.add_trace(go.Scatter(name='Upper Bound',
-                         x=freq,
-                         y=jackknife[:, 1],
-                         mode='lines',
-                         marker=dict(color="#444"),
-                         line=dict(width=0),
-                         showlegend=False))
-fig.add_trace(go.Scatter(name='Lower Bound',
-                         x=freq,
-                         y=jackknife[:, 0],
-                         mode='lines',
-                         marker=dict(color="#444"),
-                         line=dict(width=0),
-                         fillcolor='rgba(68, 68, 68, 0.3)',
-                         fill='tonexty',
-                         showlegend=False))
-fig.update_xaxes(type="log", title='Period [days]')
-fig.update_layout(font_size=22)
-fig.write_image('psd_analysis.png', width=1500, scale=2)

### 2022-04-19 manuscript/figures/uks_real_data.py

@@ -1,15 +1,0 @@
-import os
-import xarray as xr
-from pumahu.visualise import plot_qin_uks
-from pumahu.uks import main as main_uks
-
-fn = 'data/uks.nc'
-if not os.path.isfile(fn):
-    main_uks(['--rdir', './data', '-s', '2016-03-04',
-              '-e', '2021-07-07', '-f'])
-
-xdf = xr.open_dataset('data/uks.nc')
-fig = plot_qin_uks(xdf)
-fig.update_layout(showlegend=False,
-                  font_size=22)
-fig.write_image('uks_real_data.png', width=1500)

### 2022-09-30 src/pumahu/__init__.py
remove obsolete imports

@@ -8,4 +8,4 @@ from .forward_model import Forwardmodel
 from .mcmc import ns_sampling 
 from .syn_model import SynModel
 from .uks import UnscentedKalmanSmoother
-from .data import LakeData
+from .data import LakeData, get_MetService_wind, get_outflow

### 2022-09-30 manuscript/manuscript.tex
adding extra sections at the end

@@
 %\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
-\documentclass[doublespacing, linenumbers]{bmcart}
+%\documentclass[doublespacing, linenumbers]{bmcart}
+\documentclass[doublespacing]{bmcart}
 
@@
 %\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-%\usepackage{graphicx}
+\usepackage{graphicx}
 
@@
 
-\def\includegraphic{}
-\def\includegraphics{}
+%\def\includegraphic{}
+%\def\includegraphics{}
 
@@
 
-\section*{Acknowledgements}
+\section*{Declarations}
+\subsection*{Acknowledgements}
 The authors would like to thank Nico Fournier for providing feedback and ideas
@@
 
-\section*{Competing interests}
+\subsection*{Funding}%% if any
+This project was funded in parts by the Strategic Science Investment Funding to
+GNS Science within the Datascience programme and by the GeoNet programme.
+
+\subsection*{Abbreviations}%% if any
+  \begin{tabular}{ll}
+    RCL  & Ruapehu Crater Lake \\ 
+    $\dot{M}$ & Lake water mass change \\
+    $\dot{M_i}$ & Steam input \\ 
+    $\dot{M_s}$ & Melt water and precipitation inflow \\
+    $\dot{M_o}$ & Seepage and overflow \\ 
+    $H$ & Enthalpy \\
+    $T$ & Lake temperature \\
+    $c_w$ & Specific heat of water \\
+    $\dot{Q_i}$ & Energy input from volcanic sources \\
+    $\dot{Q_r}$ & Solar irradiation \\
+    $\dot{Q_e}$ & Surface energy loss \\
+    $X$ & Total ion amount in the lake\\
+  \end{tabular}
+
+\subsection*{Availability of data and materials}%% if any
+The source code for this work can be accessed at
+https://github.com/yannikbehr/pumahu. All data used in this study is provided
+free of charge by the GeoNet program (https://www.geonet.org.nz).
+
+\subsection*{Ethics approval and consent to participate}%% if any
+Not applicable
+
+\subsection*{Competing interests}
 The authors declare that they have no competing interests.
 
-\section*{Author's contributions}
+\subsection*{Consent for publication}%% if any
+Not applicable
+
+\subsection*{Author's contributions}
   Y.B. wrote the main manuscript text, produced graphs, and implemented the
@@
 
-\section*{Availability of data and materials}%% if any
-The source code for this work can be accessed at
-https://github.com/yannikbehr/pumahu. All data used in this study is provided
-free of charge by the GeoNet program (https://www.geonet.org.nz)
+\subsection*{Authors' information}%% if any
+All authors are affiliated with GNS Science. TH is Emeritus Volcano
+Geophysicist, SS is Senior Volcano Geophysicist and YB is Volcano Science
+Operations Specialist.
 
@@
 \begin{figure}[h!]
-    %\includegraphics[width=12 cm]{figures/overview.png}  
+    \includegraphics[width=12 cm]{figures/overview.png}  
     \caption{Image of Mt. Ruapehu's crater lake from 2008 (\textcopyright Visual
@@
 \begin{figure}[h!]
-   %\includegraphics[width=12 cm]{figures/sensitivity_analysis.png}  
+   \includegraphics[width=12 cm]{figures/sensitivity_analysis.png}  
     \caption{Surface energy loss ($Q_e$) with respect to lake water temperature
@@
 \begin{figure}[h!]
-	%\includegraphics[width=12 cm]{figures/synthetic_inversion_uks.png}  
+	\includegraphics[width=12 cm]{figures/synthetic_inversion_uks.png}  
     \caption{Inversion result for the synthetic test described in Section
@@
 \begin{figure}[h!]
- 	%\includegraphics[width=12 cm]{figures/time_frequency_analysis.png}  
+ 	\includegraphics[width=12 cm]{figures/time_frequency_analysis.png}  
     \caption{ a) Inference results for the heat input rate $\dot{Q_i}$ at RCL
@@
 \begin{figure}[h!]
- 	%\includegraphics[width=12 cm]{figures/cumulative_heat.png}  
+ 	\includegraphics[width=12 cm]{figures/cumulative_heat.png}  
     \caption{Cumulative difference to the mean for the heat input rate into RCL

### 2022-09-30 tests/test_data.py
trying to fix tests

@@ -9,7 +9,7 @@ import numpy as np
 import pandas as pd
 import xarray as xr
 
-from pumahu import get_data
+from pumahu import get_data, get_MetService_wind, get_outflow
 from pumahu.data import LakeData, WindData
 
 
@@ -115,7 +115,7 @@ class DataTestCase(unittest.TestCase):
 
     def test_get_Mg(self):
         """
-        Test different ways of receiving and smoothing temperature readings.
+        Test different ways of receiving and smoothing Mg readings.
         """
         ld = LakeData()
         df = ld.get_Mg(tend='2019-07-31', smoothing='dv')
@@ -136,6 +136,9 @@ class DataTestCase(unittest.TestCase):
         self.assertAlmostEqual(dkf1.iloc[0]['Mg'], 383.5)
 
     def test_get_ll(self):
+        """
+        Testing reception and smoothing of lakelevel data.
+        """
         index = pd.date_range('2019-01-01', '2019-01-03')
         kf_test_frame = pd.DataFrame({'z': np.array([2529.32191667,
                                                      2529.33634745,
@@ -161,6 +164,9 @@ class DataTestCase(unittest.TestCase):
         pd.testing.assert_frame_equal(dv_test_frame, ddv)
 
     def test_lake_data_fits(self):
+        """
+        Test downloading lake data.
+        """
         warnings.filterwarnings("ignore", message="numpy.dtype size changed")
         warnings.filterwarnings("ignore", message="numpy.ufunc size changed")
         warnings.filterwarnings("ignore", message="can't resolve package from")
@@ -185,13 +191,17 @@ class DataTestCase(unittest.TestCase):
         with open(get_data('data/wind.dat')) as wb:
             dl = WindData(wb, default=0.0)
             df = dl.get_data(start='2003-01-16', end='2010-01-29')
-            ws = [w for d, w in df.iteritems()]
-            np.testing.assert_array_almost_equal(ws, ti['wind'], 1)
+            ws = np.array([w for d, w in df.iteritems()])
+            err = ws - ti['wind'].values
+            rms = np.sqrt(np.sum(err*err)/err.size)/(ws.max() - ws.min())
+            self.assertAlmostEqual(rms, .005, 3)
 
         dl1 = WindData(get_data('data/wind.dat'), default=0.0)
         df1 = dl1.get_data(start='2003-01-16', end='2010-01-29')
-        ws = [w for d, w in df1.iteritems()]
[…]

### 2022-09-30 tests/test_fd.py

@@ -19,7 +19,7 @@ class FDTestCase(unittest.TestCase):
         q_exp = rs.exp.loc[:, 'q_in', 'val'].values
         q_diff = (q_test[:-1] - q_exp[:-1])
         rmse = np.sqrt(((q_diff) ** 2).mean())
-        self.assertAlmostEqual(rmse, 15.64, 2)
+        self.assertAlmostEqual(rmse, 17.37, 2)
 
     def test_historic_data(self):
         """
@@ -30,7 +30,8 @@ class FDTestCase(unittest.TestCase):
                           smoothing={'Mg': 2.6, 'T': 0.4, 'z': 0.01})
         xdf = xdf.dropna('dates', how='all')
         rs = fd(xdf, results_file=None, new=False, use_drmg=False)
-        print(rs)
+        rmse = np.sqrt(np.nanmean(rs.exp.loc[:, 'q_in', 'val'].values ** 2))
+        self.assertAlmostEqual(rmse, 149.6, 2)
 
 
 if __name__ == '__main__':

### 2022-09-30 tests/test_forward_model.py

@@ -69,7 +69,7 @@ class ForwardModelTestCase(unittest.TestCase):
         fw = Forwardmodel()
         y_new = fw.integrate(y0, dtime, dt, dp)
         y_new = np.array(tuple(y_new), dtype=npdp)
-        self.assertAlmostEqual(y_new['T'], 30.61, places=2)
+        self.assertAlmostEqual(y_new['T'], 30.56, places=2)
         self.assertAlmostEqual(y_new['M'], 8827.93, places=2)
         self.assertAlmostEqual(y_new['X'], 1.999, places=3)
         self.assertAlmostEqual(fw.get_steam(), 12.34, places=2)
@@ -80,23 +80,23 @@ class ForwardModelTestCase(unittest.TestCase):
         fw1 = Forwardmodel(method='rk2')
         y_new = fw1.integrate(y0, dtime, dt, dp)
         y_new = np.array(tuple(y_new), dtype=npdp)
-        self.assertAlmostEqual(y_new['T'], 30.6, places=2)
-        self.assertAlmostEqual(y_new['M'], 8827.85, places=2)
+        self.assertAlmostEqual(y_new['T'], 30.55, places=2)
+        self.assertAlmostEqual(y_new['M'], 8827.86, places=2)
         self.assertAlmostEqual(y_new['X'], 1.999, places=3)
         self.assertAlmostEqual(fw1.get_steam(), 12.34, places=2)
-        self.assertAlmostEqual(fw1.get_evap()[0], 14.9, places=2)
+        self.assertAlmostEqual(fw1.get_evap()[0], 14.88, places=2)
         self.assertAlmostEqual(fw1.get_evap()[1], 3.25, places=2)
 
         # Finally test with fourth-order Runge-Kutta
         fw2 = Forwardmodel(method='rk4')
         y_new = fw2.integrate(y0, dtime, dt, dp)
         y_new = np.array(tuple(y_new), dtype=npdp)
-        self.assertAlmostEqual(y_new['T'], 30.60, places=2)
-        self.assertAlmostEqual(y_new['M'], 8827.85, places=2)
+        self.assertAlmostEqual(y_new['T'], 30.55, places=2)
+        self.assertAlmostEqual(y_new['M'], 8827.86, places=2)
         self.assertAlmostEqual(y_new['X'], 1.999, places=3)
         self.assertAlmostEqual(fw2.get_steam(), 12.34, places=2)
[…]

### 2022-09-30 tests/test_uks.py

@@ -6,7 +6,7 @@ import xarray as xr
 
 from filterpy.kalman import unscented_transform
 
-from pumahu import get_data
+from pumahu import get_data, get_outflow
 from pumahu.uks import UnscentedKalmanSmoother
 from pumahu.syn_model import (SynModel,
                               setup_test,
@@ -46,7 +46,7 @@ class UKSTestCase(unittest.TestCase):
         uks = UnscentedKalmanSmoother(data=self.xds.exp)
         res = uks(test=True)
         log_lh = np.nansum(res['p_samples'].loc[dict(p_parameters='lh')].values)
-        self.assertAlmostEqual(log_lh, -29.2659392, 4) 
+        self.assertAlmostEqual(log_lh, -37.8449, 4) 
         
     def test_synthetic_extensive(self):
         """
@@ -66,7 +66,7 @@ class UKSTestCase(unittest.TestCase):
         uks = UnscentedKalmanSmoother(data=xds3.exp, P0=P0, Q=Q)
         rs = uks(test=True, smooth=True)
         logl_mean = np.nansum(rs.p_samples.values)/rs.dates.shape[0]
-        self.assertAlmostEqual(logl_mean, -12.56475, 4)
+        self.assertAlmostEqual(logl_mean, -15.7801, 4)
 
     def test_nans(self):
         """
@@ -77,18 +77,18 @@ class UKSTestCase(unittest.TestCase):
         logl = rs.p_samples[:, 0]
         np.testing.assert_array_almost_equal(logl,
                                              np.array([np.nan,
-                                                       -12.0177,
-                                                       -8.8329,
-                                                       -7.8945]), 4)
+                                                       -15.9388,
+                                                       -11.4676,
+                                                       -10.001]), 4)
 
     def test_real_data(self):
         ld = LakeData()
         data = ld.get_data('2019-01-01', '2019-02-01', smoothing='dv')
-        ld.get_outflow()
+        data = get_outflow(data)
         uks = UnscentedKalmanSmoother(data=data)
         rs = uks(test=True)
[…]
